---
title: "Affordability index — QA across regions & every iteration"
date: "Built from cached results; FY2024 / ACS 2020–2024"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 4.6, dpi = 96, out.width = "100%")
library(ggplot2); library(sf); library(patchwork); library(dplyr)
try(suppressMessages(pkgload::load_all("..", quiet = TRUE)), silent = TRUE)  # for afford_bands()
qa <- readRDS("afford_qa_data.rds")

# clean map theme + two fill scales: diverging for ratio (midpoint 1), sequential
# (viridis) for shares/counts.
map_theme <- theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(color = "grey35", size = 9.5),
        legend.key.height = unit(8, "pt"))

afmap <- function(x, ten, tier, col, title, sub = "", kind = c("ratio","seq"),
                  lim = NULL) {
  kind <- match.arg(kind)
  d <- x[x$tenure == ten & x$ami_tier == tier, ]
  g <- ggplot(d) + geom_sf(aes(fill = .data[[col]]), color = NA) +
    map_theme + labs(title = title, subtitle = sub)
  if (kind == "ratio")
    g + scale_fill_gradient2(low = "#b2182b", mid = "#f7f7f7", high = "#2166ac",
          midpoint = 1, limits = c(0, 2), oob = scales::squish,
          name = "supply ÷\nneed")
  else
    g + scale_fill_viridis_c(option = "mako", direction = -1, limits = lim,
          oob = scales::squish, name = col, labels = scales::label_comma())
}
med <- function(x, ten, tier, col = "ratio")
  median(x[[col]][x$tenure == ten & x$ami_tier == tier], na.rm = TRUE)
```

## What this page checks

A QA sweep of the rebuilt `afford_index()` across four very different housing markets —
**`r paste(sapply(qa$regions, '[[', 'label'), collapse=' · ')`** — exercising **every parameter
iteration** (demand universe, tenure, owner buy-in basis, Gate-2 availability, all five AMI sources,
the tier gradient, and banding). Each section asks: *is the function doing what we intend?*

All results are pre-computed (`dev/afford_qa_data.R`) and cached, so this page is reproducible and
needs no API. AMI is **HUD** (off the bundled snapshot) unless a section says otherwise. Ratio maps
are a **location quotient**: red = a tract carries *less* than its fair share of affordable supply
for the tier (`<1`), blue = *more* (`>1`), white = parity.

---

## 1. Does it run everywhere? VLI rental supply ratio

```{r region-medians}
data.frame(
  Region = sapply(qa$regions, "[[", "label"),
  `Tracts` = sapply(qa$baseline, function(x) length(unique(x$GEOID))),
  `Median VLI-rent ratio` = round(sapply(names(qa$baseline),
     function(k) med(qa$baseline[[k]], "rent", "VLI")), 2),
  check.names = FALSE, row.names = NULL)
```

```{r region-maps, fig.height=8.6}
ps <- lapply(names(qa$baseline), function(k)
  afmap(qa$baseline[[k]], "rent", "VLI", "ratio", qa$regions[[k]]$label, kind = "ratio"))
wrap_plots(ps, ncol = 2) + plot_layout(guides = "collect")
```

**Verdict.** Runs cleanly on all four (single- and multi-county). The gradient is sensible and
*differentiated*: Hennepin/Minneapolis is near parity (median 0.95), the Bay and King run below it,
and **Atlanta is the most squeezed (0.32)** — most tracts carry far less affordable rental supply
than their share of VLI renters. Exactly the kind of spatial signal the index exists to surface. ✓

---

## 2. `demand` — whose need is supply compared to? (King County)

The denominator is the fix that most moves the headline. Same VLI rental supply, three universes:

```{r demand-maps, fig.height=4.6}
afmap(qa$baseline$king,        "rent", "VLI", "ratio", "matched (vs renters) — default", kind="ratio") +
afmap(qa$king_demand_all,      "rent", "VLI", "ratio", "all households",                  kind="ratio") +
afmap(qa$king_demand_owner,    "rent", "VLI", "ratio", "owners",                          kind="ratio") +
  plot_layout(guides = "collect")
```

```{r demand-tab}
data.frame(
  demand = c("matched (renters)","all households","owners"),
  `median VLI-rent ratio` = round(c(
     med(qa$baseline$king,"rent","VLI"),
     med(qa$king_demand_all,"rent","VLI"),
     med(qa$king_demand_owner,"rent","VLI")), 2),
  check.names = FALSE)
```

**Verdict.** Working as intended. Against **renters** (default) the ratio is lower than against
**all households** — renters skew lower-income, so a given affordable-rental share covers a smaller
fraction of their need. Comparing rental supply to **owners** (higher income, fewer VLI) inflates the
ratio. The default `matched` is the report-correct universe. ✓

---

## 3. `tenure` — three cost models (King County, VLI, share affordable)

```{r tenure-maps, fig.height=4.6}
lim3 <- c(0,1)
afmap(qa$baseline$king, "rent",        "VLI", "supply", "rent (gross rent)",          kind="seq", lim=lim3) +
afmap(qa$baseline$king, "own_current", "VLI", "supply", "own_current (monthly cost)", kind="seq", lim=lim3) +
afmap(qa$baseline$king, "own_buyin",   "VLI", "supply", "own_buyin (buy in today)",   kind="seq", lim=lim3) +
  plot_layout(guides = "collect")
```

**Verdict.** The three cost models behave distinctly and correctly. `own_current` (carrying the homes
people already own, often on old cheap mortgages) shows the **most** VLI-affordable owner stock;
`own_buyin` (buying in *today* at ~6.9%) shows the **least** — the real barrier a mover faces. Rent
sits between. This is the v1 `0.188` bug fixed: ownership affordability is now rate-aware. ✓

---

## 4. `buyin_stock` — which homes count for buy-in (King, own_buyin, VLI)

```{r buyin-maps, fig.height=4.6}
afmap(qa$baseline$king,    "own_buyin", "VLI", "supply", "owned_value (B25075) — default", kind="seq", lim=c(0,1)) +
afmap(qa$king_buyin_fs,    "own_buyin", "VLI", "supply", "for_sale (B25085) — the flow",   kind="seq", lim=c(0,1)) +
  plot_layout(guides = "collect")
```

**Verdict.** Both work; the contrast is the point. `owned_value` uses the full, stable distribution of
what homes are worth; `for_sale` uses only the sparse vacant-for-sale flow (far fewer units, noisier,
many tracts near zero) — closer to "what a mover could actually buy now." Availability (next) is the
cleaner way to handle openness. ✓

---

## 5. Gate 2 — affordable *vs.* attainable (King, VLI rent)

`accessible` = affordable units (Gate 1). `available_turnover` = those that actually open up in a year
(Gate 1 × turnover rate). Shared color scale, so the shrinkage is visible:

```{r avail-maps, fig.height=4.6}
acc_lim <- c(0, quantile(qa$baseline$king$accessible[qa$baseline$king$tenure=="rent" &
                          qa$baseline$king$ami_tier=="VLI"], 0.97, na.rm=TRUE))
afmap(qa$baseline$king, "rent", "VLI", "accessible",         "Gate 1: affordable units",          kind="seq", lim=acc_lim) +
afmap(qa$baseline$king, "rent", "VLI", "available_turnover", "Gate 2: + turnover (attainable/yr)", kind="seq", lim=acc_lim) +
  plot_layout(guides = "collect")
```

**Verdict.** Gate 2 attaches and behaves correctly — the attainable map is a strict, much dimmer
subset of the affordable map (turnover ≈ 15–25% of renter stock/yr). "Affordable on paper" is far
larger than "actually open to move into," which is the whole point of the funnel. ✓

---

## 6. `ami_source` — the math check (income ceilings, 4 sources)

This is where correctness lives. One county per region, every source's 4-person ceilings:

```{r cutoffs-tab}
ct <- qa$cutoffs
ct$ami <- round(ct$ami); ct$ELI <- round(ct$ELI); ct$VLI <- round(ct$VLI); ct$LI <- round(ct$LI)
ct
```

**Verdict — behaving exactly as designed, and this is the most important check:**

- **Bay (SF), high-cost coastal:** `acs` VLI (85.6k) **understates** HUD (97.9k) by ~13% — it omits
  HUD's FMR high-cost bump. `hud_acs` reproduces HUD here **to the dollar** (97,900). Use `hud`/`auto`.
- **Atlanta (Fulton):** `acs`/`hud_acs` VLI (65k/61k) **overstate** HUD (53.75k) — Fulton county is
  richer than the metro HUD averages over (the documented county-vs-metro divergence). `hud` is right.
- **King:** `hud_acs` LI is capped at the US median (97,800) while HUD's actual LI (110,950) exceeds
  it via the high-cost exception — the documented cap limitation of the from-scratch reproduction.
- **Hennepin:** ordinary-cost — sources converge closely; `hud_acs` ≈ `hud`.

Net: `acs` is transparent but diverges in high-cost *and* rich-county-in-poor-metro cases; **`hud`
(the default `auto`) is the one to trust for reports**, and the snapshot makes it offline-exact. The
function steers you correctly. ✓

---

## 7. Tier gradient + banding (King, rent)

Cumulative ceilings ELI ≤30% → MI ≤120%. Supply should rise monotonically as the ceiling rises:

```{r tier-maps, fig.height=8.6}
ps <- lapply(c("ELI","VLI","LI","MI"), function(ti)
  afmap(qa$baseline$king, "rent", ti, "supply", paste0(ti, " (≤", c(ELI=30,VLI=50,LI=80,MI=120)[ti], "% AMI)"),
        kind = "seq", lim = c(0,1)))
wrap_plots(ps, ncol = 2) + plot_layout(guides = "collect")
```

```{r band-check}
king_df <- sf::st_drop_geometry(qa$baseline$king)
cum <- sapply(c("ELI","VLI","LI","MI"), function(ti)
  round(median(king_df$supply[king_df$tenure=="rent" & king_df$ami_tier==ti], na.rm=TRUE), 3))
banded <- tryCatch({
  b <- afford_bands(king_df)
  sapply(c("ELI","VLI","LI","MI"), function(ti)
    round(median(b$supply[b$tenure=="rent" & b$ami_tier==ti], na.rm=TRUE), 3))
}, error = function(e) rep(NA, 4))
data.frame(tier = c("ELI","VLI","LI","MI"),
           `cumulative supply (≤ ceiling)` = unname(cum),
           `banded supply (slice via afford_bands)` = unname(banded),
           check.names = FALSE)
```

**Verdict.** Cumulative supply is monotone as it must be — each higher ceiling captures strictly more
stock (ELI ⊂ VLI ⊂ LI ⊂ MI). `afford_bands()` differences them into the non-overlapping slices
(the banded column no longer nests), for partition/stacked views. ✓

---

## Bottom line

Across four contrasting markets and every parameter, `afford_index()` does what we intend: the
spatial signal is sensible and differentiated, the demand universes shift the ratio in the right
direction, the tenure/buy-in models are rate-aware, Gate 2 correctly shrinks affordable to attainable,
the tiers are monotone, and the AMI sources diverge exactly where the design says they should —
with `hud`/`auto` steering toward the correct, report-grade numbers. The one thing to remember for
reports: **stick with `hud` (or `auto`) in high-cost or rich-county-in-poor-metro regions**, where
the Census-only sources knowingly drift.
```
