---
title: "Stable destinations (P1): joining affordability with HPRM stability"
subtitle: "A walkthrough of the first cut — King County, WA (Seattle) and the 9-county Bay Area, VLI renters"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 11, dpi = 96, out.width = "100%")
library(ggplot2); library(sf); library(patchwork); library(dplyr)
suppressMessages(pkgload::load_all("..", quiet = TRUE))   # nt_map() + afford_*/inflow_index()
d <- readRDS("p1_eval_data.rds"); s <- d$summ; sb <- d$summ_bay
ck <- d$cap_king[d$cap_king$ami_tier == "VLI", ]   # free-will capacity, VLI renters
cb <- d$cap_bay[d$cap_bay$ami_tier == "VLI", ]
king <- d$king; vli <- king[king$ami_tier == "VLI", ]
bay  <- d$bay;  bay_vli <- bay[bay$ami_tier == "VLI", ]
dvlim <- range(c(king$dis_value, bay$dis_value), na.rm = TRUE)   # shared inflow colour scale
mt <- theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(color = "grey35", size = 9),
        legend.key.height = unit(7, "pt"))
map_aff <- function(v) ggplot(v) + geom_sf(aes(fill = supply), color = NA) + mt +
  scale_fill_gradient(low = "#f7fbff", high = "#08519c", limits = c(0,1),
                      labels = scales::label_percent(), name = "% of rentals\naffordable to VLI") +
  labs(title = "Where VLI can afford", subtitle = "darker = a larger share of rentals within reach")
map_stab <- function(v) ggplot(v) + geom_sf(aes(fill = stability), color = NA) + mt +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850", midpoint = 0.5,
                       limits = c(0,1), breaks = c(0,.5,1), labels = c("0\nprecarious","0.5","1\nstable"),
                       name = "stability") +
  labs(title = "Where it's stable", subtitle = "red = precarious (high HPRM), green = stable")
bivar_cat <- function(v) factor(dplyr::case_when(
    v$supply <  median(v$supply, na.rm = TRUE) ~ "Less affordable",
    v$stability >= 0.8                          ~ "Affordable & stable (>= 0.8)",
    v$stability >= 0.5                          ~ "Affordable, elevated risk (0.5-0.8)",
    TRUE                                        ~ "Affordable, high precarity (< 0.5)"),
  levels = c("Affordable & stable (>= 0.8)","Affordable, elevated risk (0.5-0.8)",
             "Affordable, high precarity (< 0.5)","Less affordable"))

# --- interactive MapLibre choropleths (the ggplot map_*() below are now unused) ---
vli$cat <- bivar_cat(vli); bay_vli$cat <- bivar_cat(bay_vli)
prep <- function(v) {   # trim to display columns + lightly simplify for a responsive web map
  v <- v[, c("GEOID", "county", "supply", "stability", "dis_value", "cat",
             "verdict", "stable_cat")]
  v <- sf::st_transform(v, 3857)
  v <- sf::st_simplify(v, dTolerance = 60, preserveTopology = TRUE)
  sf::st_transform(v, 4326)
}
vli_m <- prep(vli); bay_m <- prep(bay_vli)
dvmax <- max(c(king$dis_value, bay$dis_value), na.rm = TRUE)   # shared inflow ceiling
# headline maps render VERDICTS, not gradients (the across-the-board rule):
# a continuous scale that shows a place as "mid" reads as "no problem here".
ml_verdict <- function(v) { v <- v[!is.na(v$verdict), ]
  nt_map(v, "verdict", type = "categorical",
    colors = c("#1a9850", "#fdae61", "#d73027"),
    legend_title = "verdict for a VLI renter", before_id = "water") }
ml_stabcat <- function(v) { v <- v[!is.na(v$stable_cat), ]
  nt_map(v, "stable_cat", type = "categorical",
    colors = c("#1a9850", "#fdae61", "#d73027"),
    legend_title = "stability verdict", before_id = "water") }
ml_dv    <- function(v) nt_map(v, "dis_value", type = "continuous",
    colors = c("#d73027", "#fdae61", "#1a9850"), breaks = c(50, dvmax),
    legend_title = "net low-income migration (pivot +50)", before_id = "water")
ml_bivar <- function(v) nt_map(v, "cat", type = "categorical",
    colors = c("#1a9850", "#fdae61", "#d73027", "#d9d9d9"),
    legend_title = "afford x stability", before_id = "water")
map_bivar <- function(v) { v$cat <- bivar_cat(v)
  ggplot(v) + geom_sf(aes(fill = cat), color = "white", linewidth = 0.04) + mt +
  scale_fill_manual(values = c("Affordable & stable (>= 0.8)"="#1a9850",
                               "Affordable, elevated risk (0.5-0.8)"="#fdae61",
                               "Affordable, high precarity (< 0.5)"="#d73027",
                               "Less affordable"="#d9d9d9"), name = NULL, drop = FALSE) +
  labs(title = "Where can a VLI renter afford — and stay?",
       subtitle = "affordable tracts graded by stability: green >= 0.8, orange 0.5-0.8, red < 0.5") +
  theme(legend.position = "bottom", legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(nrow = 2)) }
map_dv <- function(v) ggplot(v) + geom_sf(aes(fill = dis_value), color = "white", linewidth = 0.04) + mt +
  scale_fill_gradient2(low = "#d73027", mid = "#fdae61", high = "#1a9850", midpoint = 50,
    limits = dvlim, oob = scales::squish, name = "net low-income\nmigration\n(dis_value)") +
  labs(title = "Where are low-income households moving to?",
       subtitle = "EDR net migration, pivot +50: red = outflow (displacement), orange ≈ break-even, green = inflow (growth)")
```

## What this evaluates

This is the first cut of the **"good place" / stable-destination layer** — the step that turns the
affordability index into a *destination* measure. The logic, per the design (`dev/good-place-design.md`)
and the theory (`hprm/manuscripts/a-good-place.md`):

> **A good place is not a tract's score — it's a household's *access*: where it can afford, reach, and
> *stay*.** Affordability tells us where a VLI renter *could* live; it says nothing about whether
> they'd be pushed back out. So we intersect the afford-and-reach set with **housing stability**, which
> we read directly from ERN's own HPRM (displacement + eviction risk) rather than from any "opportunity"
> score.

Here we run it for **King County, WA** (Seattle), VLI renters (≤50% AMI), renting — the cleanest test
region (single county; both HPRM signals present and non-CA, so no out-of-sample EER caveat) — then
extend to the **9-county Bay Area**, the high-displacement contrast, below.

## How "stability" is measured

Stability is built from two HPRM fields, each mapped to a 0–1 score using **HPRM's own published risk
anchors**, then **equal-weighted** (your call):

- **Displacement** — `dis_value` (the EDR composite). It is deliberately *not* a raw sum of the AMI-tier
  net-migration predictions: it sums only the tiers at or below the **+50 BART-uncertainty buffer**, so
  a confidently in-migrating tier can't mask a displacing one. Anchors: `+50 → 1.0` (stable), `−300`
  (Extreme) `→ 0`.
- **Eviction** — `ev_value` (the EER filing-rate ratio; 1.0 = state average). Anchors: `0.8 → 1.0`,
  `2.0` (Extreme) `→ 0`.

```{r curves, fig.height=2.8, fig.width=10}
g1 <- ggplot(d$fn_edr, aes(dis_value, s_edr)) +
  annotate("rect", xmin=-Inf, xmax=-300, ymin=-Inf, ymax=Inf, fill="#b2182b", alpha=.08) +
  geom_line(linewidth=1, color="#1f6f78") +
  geom_vline(xintercept=c(50,-100,-300), linetype=3, color="grey55") +
  annotate("text", x=c(60,-95,-295), y=0.05, label=c("At Risk","Elevated","Extreme"),
           hjust=0, size=2.7, color="grey40") +
  labs(title="Displacement → stability", x="dis_value (EDR)", y="s_edr") + theme_minimal(base_size=10)
g2 <- ggplot(d$fn_eer, aes(ev_value, s_eer)) +
  annotate("rect", xmin=2, xmax=Inf, ymin=-Inf, ymax=Inf, fill="#b2182b", alpha=.08) +
  geom_line(linewidth=1, color="#1f6f78") +
  geom_vline(xintercept=c(0.8,1.5,2.0), linetype=3, color="grey55") +
  annotate("text", x=c(0.82,1.52,2.02), y=0.95, label=c("At Risk","High","Extreme"),
           hjust=0, size=2.7, color="grey40") +
  labs(title="Eviction → stability", x="ev_value (EER)", y="s_eer") + theme_minimal(base_size=10)
g1 + g2
```
*Figure 1. Each HPRM risk field mapped to a 0-1 stability score on its published anchors.*

`stability = 0.5 · s_edr + 0.5 · s_eer`, on every tract — so the 0–1 scale reads in plain terms:

- **1.0 = fully stable** — no measurable displacement (`dis_value ≥ +50`) *and* below-average eviction
  (`ev_value ≤ 0.8`): low-income renters are holding on here.
- **0.0 = maximally precarious** — Extreme on *both* (`dis_value ≤ −300` *and* `ev_value ≥ 2.0`).
- **0.5 = the midpoint** — e.g., "High" risk on both dimensions, or Extreme on one and stable on the other.

**EDR vs EER — and a switch.** These aren't the same harm. **EDR** (`dis_value`) is *net out-migration* —
low-income households actually **leaving**. **EER** (`ev_value`) is the *added precarity of the
formal-eviction channel*. They can diverge — HPRM's "policy substitution," where tenant protections
suppress eviction filings while market displacement intensifies. So stability should take a
**`stability_source`** option: **`both`** (default), **`edr`** (displacement only — the purest "are
low-income people being pushed out?"), or **`eer`** (eviction only). For California, where EER is
out-of-sample, `edr` is the robust choice. *(`stability_source` is now a coded parameter of
`afford_stability()`; this walkthrough uses `both`.)*

**Honesty note:** the HPRM signal is *built* from individual data (the Data Axle panel + court records —
Carlson's individual approach, not the population/compositional method that invites the ecological
fallacy), but its *output* is a tract rate. So we treat `stability` as a **place-level risk descriptor**,
not a guarantee for any one household — the individual-level check is the Gate-4 observed-move residual,
later.

## The data join

HPRM is FY2022 (2020-census tracts); the afford index is 2024 (also 2020-census tracts), so they join
cleanly on `GEOID`. Coverage in King County: **`r s$join_matched` of `r s$join_total` tracts** (the one
miss is a no-renter edge tract).

## The free-will arithmetic

Before any map: if every VLI renter household could re-sort freely, paying at most 30% of income
in rent, what could each region absorb? The `afford_capacity()` readout (VLI renters, rent):

```{r capacity}
cap <- rbind(cbind(region = "King County", as.data.frame(ck)),
             cbind(region = "Bay Area (9-county)", as.data.frame(cb)))
knitr::kable(dplyr::transmute(cap, region,
  `VLI renter households`      = scales::comma(tier_households, accuracy = 100),
  `affordable units (at 30%)`  = scales::comma(affordable_units, accuracy = 100),
  `per 100 households`         = round(per100_affordable, 1),
  shortfall                    = scales::comma(shortfall, accuracy = 100),
  `per 100 at 50% (stretch)`   = round(per100_stretch, 1),
  `open per year, per 100`     = round(per100_open_year, 1),
  `open now, per 100`          = round(per100_open_now, 1),
  `tracts < 1 opening/yr`      = tracts_lt1_open))
```

King County's rental stock holds **`r round(ck$per100_affordable)` affordable units per 100 VLI
renter households** — a structural deficit of **`r scales::comma(round(ck$shortfall))` units even
under perfect free sorting** — and only **`r round(ck$per100_open_year)` per 100 open up in a
year** (**`r round(ck$per100_open_now, 1)` per 100** are vacant right now). The Bay holds
**`r round(cb$per100_affordable)` per 100** (shortfall **`r scales::comma(round(cb$shortfall))`**),
with **`r round(cb$per100_open_year)` per 100** opening yearly. At the 50%-of-income *stretch*
line the reachable stock roughly doubles (King **`r round(ck$per100_stretch)`**, Bay
**`r round(cb$per100_stretch)`** per 100) — the shortage lives at the 30% standard, not at 50%.
*These are price-only ceilings: an affordable unit occupied by a higher-income household still
counts. The NLIHC-style "affordable and available" competition adjustment is the planned
refinement (see `dev/affordability-deep-dive.md`).*

## The maps

The two ingredients first, each as a **verdict** — per the across-the-board presentation rule:
policy audiences read maps by color, and a gradient that renders a tract "mid" is read as "no
problem here," so headline maps classify. For affordability the two burden lines give the classes
plain meaning at the tract-majority (= median-unit) rule: **affordable** — the typical rental
costs a VLI household ≤ 30% of income; **roughly affordable** — 30–50% (reachable only by
stretching); **not affordable** — more than half their income, even stretching.

**All maps below are interactive** - pan, scroll to zoom, hover a tract for its class or value,
click for its GEOID, on a light Positron basemap.

```{r map_aff}
ml_verdict(vli_m)
```
*Figure 2. The verdict — where King County is affordable to a VLI renter: green = affordable (typical rental ≤ 30% of income), orange = roughly affordable (30–50%), red = not affordable (> 50% even stretching).*

King splits **`r s$v_aff` affordable / `r s$v_rough` roughly affordable / `r s$v_not` not
affordable**. The majority of the county is *orange* — reachable only by stretching past the 30%
standard — which a continuous blue gradient rendered as unremarkable mid-tones.

```{r map_stab}
ml_stabcat(vli_m)
```
*Figure 3. The stability verdict — green = stable (≥ 0.8), orange = elevated (0.5–0.8), red = precarious (< 0.5), on HPRM's published anchors (from `afford_stability()`'s `stable_cat`).*

Stability gets the same verdict grammar. Precarity concentrates in the dense urban core; the rural
east is broadly stable.

Now grade the *affordable* tracts by stability — with a **traffic-light** cut, because (as you flagged)
an **Elevated**-risk tract isn't a settled destination: **green = stability ≥ 0.8** (genuinely stable,
≈ HPRM "At Risk or better"); **orange = 0.5–0.8** (Elevated risk — reachable but not settled);
**red = < 0.5** (High/Extreme precarity).

```{r map_bivar}
ml_bivar(vli_m)
```
*Figure 4. Where a VLI renter can afford - and stay. Affordable tracts graded by stability: green >= 0.8, orange 0.5-0.8, red < 0.5; grey = less affordable.*

**Green** = affordable and genuinely stable (≥ 0.8). **Orange** = affordable but at **Elevated** risk
(0.5–0.8) — reachable, not a settled destination; *this is the band you flagged, now visible instead of
buried inside "green."* **Red** = affordable but High/Extreme precarity (< 0.5). **Grey** = less
affordable. In King's more-affordable VLI half: **`r s$nc_green` green, `r s$nc_orange` orange, `r s$nc_red` red**.
Raising the bar to 0.8 surfaces the cautionary orange ring the 0.5 cut was hiding — and, as the Bay
section below shows, the *same* split recurs almost unchanged in a very different, higher-displacement market.

**Switching to `edr`-only** (displacement only, dropping the eviction channel) keeps the trade-off: the
motivating cases from earlier this session — **San Diego ≈ −0.48, Salt Lake ≈ −0.43** (both-signal) — are
EDR-driven, so this is a *people-leaving* phenomenon first; eviction adds to it but doesn't create it.
(`stability_source = "edr"` recomputes any of these directly.)

## The three combination modes

The design specifies three, all available; the **continuous weight (the map above) is the default**.

**(a) Report-alongside** — the afford metrics with the stability fields attached, so nothing is hidden.
The most-affordable VLI tracts, and how precarious each is:

```{r tab}
vli |> sf::st_drop_geometry() |>
  arrange(desc(supply)) |>
  transmute(GEOID, `afford supply` = round(supply,2), dis_value = round(dis_value),
            ev_value = round(ev_value,2), hprm_score, stability = round(stability,2),
            `stable?` = ifelse(stable, "yes","no")) |>
  head(8)
```

**(b) Hard "stable destination" filter** — affordable + reachable + `stability ≥ 0.8` (the **green**
tracts in the map above). In King's affordable half, **`r s$nc_green`** clear the 0.8 bar, **`r s$nc_orange`** sit
in the orange Elevated band, and **`r s$nc_red`** are red — and the Bay (below) reproduces almost the same
split (green ≈ `r round(100*sb$nc_green/(sb$nc_green+sb$nc_orange+sb$nc_red))`% of its affordable half, vs `r round(100*s$nc_green/(s$nc_green+s$nc_orange+s$nc_red))`% here).

**(c) Continuous stability weight** — the **default**: each tract's `available_turnover × stability`
(**availability now folded in** — the annual openings a group can afford, smoothly down-weighted by
stability rather than hard-cut), so a moderately-precarious tract still counts, partially. This is
the `stable_dest` column `afford_stability()` returns.

## The finding

The point of the layer, in one number: across King County tracts, **affordability and stability
correlate `r round(s$cor_aff_stab, 2)`** for VLI renters. The places a VLI household can most afford are
systematically the *least* stable — affordable tracts run median stability **`r round(s$med_stab_aff,2)`**
vs **`r round(s$med_stab_less,2)`** for less-affordable ones. This is the affordability-precarity
trade-off the opportunity paradigm can't see and the stable-destination layer is built to surface: a
VLI renter's reachable, *holdable* options are fewer, and they skew to the edge.

**Caveats (carry these):**

- The maps are **visually area-weighted** — the large rural-east polygons dominate the eye. The
  "periphery" read needs **unit/population weighting** before it's a finding, not just a picture.
- HPRM is **2022**, afford is **2024** (a 2-year vintage gap; both 2020-census tracts).
- King has both HPRM signals in-sample. For **California**, EER (`ev_value`) is an *out-of-sample*
  prediction (CA excluded from EER training) — usable, but flag it.
- The stability function is the agreed default (equal weights; published anchors) — configurable.

## Aside: where are low-income households actually *moving to*?

You flagged something the stability layer deliberately sets aside. `dis_value` is **signed net
migration of low-income households** (codebook, `nmr_pred_*`: *"Negative = net outflow"*). The
stability score treats everything at or above the +50 buffer as equally "stable" — but the
*positive* tail is not merely the absence of displacement, it's **low-income neighborhood
growth**: tracts where ELI/VLI/LI households are, on net, *moving in*. The codebook is explicit —
when all three tiers clear +50, `dis_value` is *"backfilled with the raw sum (positive =
low-income growth)."*

So here is the raw `dis_value` range, pivoted at **+50** (the same buffer the HPRM risk bands use):
red = net outflow (displacement), orange ≈ break-even, green = net inflow (low-income growth).

```{r map_dv}
ml_dv(vli_m)
```
*Figure 5. Where low-income households are moving to (King) - EDR net migration, pivot +50: red = outflow (displacement), green = inflow (growth).*

In King County, **`r s$n_growth`** tracts are confident low-income **growth** (`dis_value > +50`,
green), **`r s$n_disp`** are net **outflow** (`dis_value < 0`, red), and **`r s$n_even`** sit in the
0-to-50 break-even zone the model can't distinguish from stable (`dis_value` spans
`r round(s$dv_min)` to `r round(s$dv_max)`).

**A caution you raised — this is a *prediction*, not a headcount.** `dis_value`/`nmr_pred_*` are BART
predictions of net migration (Data Axle panel, INLA-smoothed), so a "growth" tract is where the model
*expects* low-income in-migration to concentrate. The model's validated target is *displacement* (net
outflow), so the positive/growth tail is comparatively **less validated** and is shrunk toward regional
means. Treat it as a modeled signal to **triangulate** against observed flows (ACS
geographic-mobility-by-income; IRS SOI county migration by AGI), not as ground truth — a *more recent*
Data Axle vintage would refresh recency but not close the prediction-vs-observation gap or the panel's
coverage bias (it underrepresents the poorest and undocumented). `inflow_index()` is source-agnostic
here: pass observed net migration in the same shape and it works unchanged.

> **Face-validity failure - do not read the positive/green tail as "destinations."** Mapped for the
> Bay (Figure 8), positive `dis_value` reads *exclusionary and receiving suburbs* as major low-income
> growth - **Marin median +338**, Sonoma +448, Contra Costa +400 - while the expensive cores go
> negative. Marin as a low-income destination is not credible. EDR is a *displacement* model; its
> positive tail is unvalidated extrapolation dominated by spatial smoothing where the low-income base
> is tiny. **Decision (July 2026): center destinations on affordability + stability, not this.**
> Positive `dis_value`'s one good use is *absence of displacement* feeding stability (which clamps the
> magnitude away, so stability is robust to this). A purpose-built growth model is specified in
> `hprm/HPRM_3.0_growth_model.md`. The two inflow maps below are kept as the *evidence* for this.

**This is a different question from stability, and it deserves its own measure.** Stability asks
*"can a household stay?"* (low displacement risk); this map asks *"where are low-income households
actually going?"* — a **revealed-destination / demand** signal. The two are related but not the
same: a growth tract is stable *for the incoming group*, but the flow itself may be **channeled**
(the "highways of migration" — place stratification; Krysan & Crowder's cycles of segregation), not
free. The right home for this is a **separate function** (a revealed-in-migration layer) built on
the **raw `nmr_pred_el/vl/l`** fields — tier-specific (*"where are VLI households moving?"*), not the
displacement-preserving composite. The payoff is the comparison the whole project is after: **where
a group *can afford* to go** (`afford_index`) versus **where it *actually* goes** (revealed
in-migration). That gap — potential minus revealed destinations — is the economic-segregation
signal itself.

## The 9-county Bay Area

Does the affordability–precarity trade-off hold in a high-displacement metro, or is it a King artifact? Same recipe (VLI renters,
`afford_stability()` + `inflow_index()`) across the **9-county Bay Area** (Alameda, Contra Costa, Marin,
Napa, San Francisco, San Mateo, Santa Clara, Solano, Sonoma): **`r sb$join_matched` of `r sb$join_total`**
tracts joined.

*Two method notes.* (1) Three Bay counties — Napa, Solano, Sonoma — aren't in the offline HUD snapshot, so
the Bay uses the **`hud_acs`** source (the validated ACS-based HUD approximation, MFI within ~1% of HUD)
for all nine; King used the exact `hud` snapshot. (2) The Bay is **California, so EER (`ev_value`) is
out-of-sample** — stability below is `both`, but `edr`-only is the robust read here.

First the verdict (same classes as Figure 2): the Bay splits **`r sb$v_aff` affordable /
`r sb$v_rough` roughly affordable / `r sb$v_not` not affordable** for VLI renters.

```{r bay_verdict_map}
ml_verdict(bay_m)
```
*Figure 6. Bay Area - the affordability verdict for a VLI renter (same classes as Figure 2).*

```{r bay_bivar_map}
ml_bivar(bay_m)
```
*Figure 7. Bay Area - where a VLI renter can afford and stay (same traffic-light scale as Figure 4).*

The answer is **yes, at almost exactly the same strength**: the trade-off is **`r round(sb$cor_aff_stab, 2)`**
for Bay VLI renters — identical to King's **`r round(s$cor_aff_stab, 2)`**. Affordable tracts run median
stability **`r round(sb$med_stab_aff, 2)`** against **`r round(sb$med_stab_less, 2)`** for less-affordable
ones (King: `r round(s$med_stab_aff, 2)` vs `r round(s$med_stab_less, 2)` — a hair wider gap in the Bay).
Of the affordable VLI half, **`r sb$nc_green` green / `r sb$nc_orange` orange / `r sb$nc_red` red** — green
is **`r round(100*sb$nc_green/(sb$nc_green+sb$nc_orange+sb$nc_red))`%** of the affordable half, essentially
King's **`r round(100*s$nc_green/(s$nc_green+s$nc_orange+s$nc_red))`%**. The finding here is *not* that the
Bay is worse — it's that the trade-off is **structural and metro-invariant**: two very different housing
markets, the same −0.49. That is a stronger claim for the manuscript than a Bay-specific effect would be.

And where are low-income households actually moving (shared +50-pivot scale)?

```{r bay_dv_map}
ml_dv(bay_m)
```
*Figure 8. Bay Area - where low-income households are moving to (same +50-pivot scale as Figure 5).*

**`r sb$n_growth`** Bay tracts are confident low-income **growth** (`> +50`), **`r sb$n_disp`** net
**outflow** (`< 0`), **`r sb$n_even`** break-even (`dis_value` `r round(sb$dv_min)` to `r round(sb$dv_max)`).
*Here* is where the Bay actually diverges from King: **`r round(100*sb$n_disp/sb$join_total)`%** of Bay
tracts are net outflow, against **`r round(100*s$n_disp/s$join_total)`%** in King — the higher displacement
pressure that the equal trade-off *strength* conceals shows up in the raw migration, not in the bivariate.
Read the two Bay maps together: where VLI can most afford (upper map) and where low-income households
actually land (green, lower map) only partly coincide — and that gap between *afford* and *inflow* is the
economic-segregation signal the project is built to measure.

## Next steps

1. ✅ **Wrapped into a package function** — `afford_stability()` now folds HPRM stability onto an
   `afford_index()` result, with the `stability_source` switch (`both`/`edr`/`eer`), the CA-EER
   out-of-sample flag, and **availability folded into the chain** (`stable_dest = available_turnover
   × stability`, not `supply × stability`). This walkthrough is now generated *through* it.
2. ✅ **Revealed-in-migration function** — `inflow_index()` (`dis_value` overall; raw `nmr_pred_*` per
   tier), joinable onto an `afford_index()` result for the potential-vs-revealed comparison. **Now flagged
   experimental** - its positive `dis_value` tail fails face validity (Marin reads as low-income
   growth; see the caveat above Figures 5/8), so use it for the *displacement* read only; the
   purpose-built growth model is specified in `hprm/HPRM_3.0_growth_model.md`.
3. ✅ **Verdict + capacity (this build)** — `afford_index()` now carries both burden lines
   (`burden` 30% / `stretch` 50%), `afford_verdict()` renders the three-class verdict the headline
   maps above use (per the across-the-board rule: classify, don't shade), and `afford_capacity()`
   produces the free-will table (units per 100, shortfall, openings). See
   `dev/affordability-deep-dive.md` for the v1 autopsy behind this.
4. **Competition adjustment** — NLIHC-style "affordable *and available*" (a cheap unit occupied by
   a higher-income household isn't available): HUD CHAS at tract level. Also: gate the bivariate
   (Figs 4/7) on the absolute verdict instead of the median split, and add MOE/reliability flags.
5. **Extend** to AMI × race (Bay Area started above) and to Atlanta / Minneapolis — does the
   affordability-precarity trade-off hold, and is it worse for Black and Latino renters? And
   **triangulate** `inflow_index()` against observed ACS/IRS migration (it's a prediction).
6. **P2 — the headline**: the Constraint Gap by AMI × race.
```
