---
title: "Report-style comparison: the published v1 affordability layer vs v2"
subtitle: "San Diego County & Salt Lake County — 'Limited Renter Affordability' with the EDR displacement hatch. Interactive: pan or zoom any map and its row follows."
output:
  html_document:
    toc: false
    self_contained: true
---

<style>
.main-container { max-width: 1500px !important; }
.hatch-note  { color: #7a1f1a; font-size: 0.8rem; margin: 6px 0 0 0; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(sf); library(dplyr); library(htmltools)
suppressMessages(pkgload::load_all("..", quiet = TRUE))
options(tigris_use_cache = TRUE, tigris_progress = FALSE)

# HPRM v5 2022: the displacement hatch = any income group at displacement risk
# (dis_group non-NA) -- the v5 analog of the reports' "Overall Displacement"
# overlay ("the number of income groups experiencing any displacement risk").
h <- arrow::read_parquet(
  "/Users/buffalo/data/evictionresearch/hprm/hprm_v5_full_2022.parquet") |>
  transmute(GEOID = geoid, dis_value, dis_group)

# report palette (screenshot hues); all three class schemes share it so the
# eye compares like with like
lev1 <- c("Not affordable", "Less affordable", "More affordable")
lev2 <- c("Not affordable (>50%)", "Roughly affordable (30-50%)",
          "Affordable (<=30%)")
pal <- c("#f4a582", "#fbf2b4", "#74b56e")
rate_cls <- function(rate) factor(case_when(   # the published Not/Less/More
  rate <= 100 ~ lev1[1],                       # cutpoints: accessible rentals
  rate <= 200 ~ lev1[2],                       # per 100k tier households
  TRUE        ~ lev1[3]), levels = lev1)

# 45-degree hatch lines clipped to the displacement tracts
hatch_lines <- function(g, spacing = 1200) {
  g <- sf::st_union(sf::st_transform(g, 3857))
  bb <- sf::st_bbox(g)
  hgt <- unname(bb["ymax"] - bb["ymin"])
  x0 <- seq(unname(bb["xmin"]) - hgt, unname(bb["xmax"]), by = spacing)
  ln <- sf::st_sfc(lapply(x0, function(x)
    sf::st_linestring(rbind(c(x, unname(bb["ymin"])),
                            c(x + hgt, unname(bb["ymax"]))))), crs = 3857)
  ln <- suppressWarnings(sf::st_intersection(ln, g))
  sf::st_as_sf(data.frame(hatch = "displacement"),
               geometry = sf::st_transform(ln, 4326))
}
simp <- function(g) sf::st_transform(
  sf::st_simplify(sf::st_transform(g, 3857), dTolerance = 60,
                  preserveTopology = TRUE), 4326)

build <- function(state, co) {
  v1 <- suppressMessages(afford(state, co, 0.8, 2024))
  v1$cls <- rate_cls(v1$tr_rent_rate)
  v2 <- suppressMessages(afford_index(state, co, 2024, tenure = "rent",
                                      ami_source = "hud")) |> afford_verdict()
  li  <- v2 |> filter(ami_tier == "LI")     # the layer's group: 0-80% AMI
  li$cls <- rate_cls(li$rate)               # published classes, corrected engine
  vli <- v2 |> filter(ami_tier == "VLI")    # pilot group: 0-50% AMI
  vli$cls <- factor(case_when(
    vli$verdict == "not affordable"     ~ lev2[1],
    vli$verdict == "roughly affordable" ~ lev2[2],
    vli$verdict == "affordable"         ~ lev2[3]), levels = lev2)
  geo <- simp(sf::st_transform(
    tigris::tracts(state = state, county = co, cb = TRUE, year = 2024), 4326)[, "GEOID"])
  hp <- h[startsWith(h$GEOID, paste0(state, co)), ]
  gj <- function(d) left_join(geo, as.data.frame(d)[, c("GEOID", "cls")],
                              by = "GEOID") |> left_join(hp, by = "GEOID")
  list(g1 = gj(v1), g2 = gj(li), g3 = gj(vli),
       li_verdict = table(li$verdict, useNA = "ifany"),
       hatch = hatch_lines(geo[geo$GEOID %in% hp$GEOID[!is.na(hp$dis_group)], ]))
}

# ---- synced MapLibre panels: nt_map_sync(), the package function this
# document's inline JS was promoted into --------------------------------------
mk_map <- function(g, legend_title, hatch) {
  g <- g[!is.na(g$cls), c("GEOID", "cls")]
  nt_map(g, "cls", type = "categorical", colors = pal,
         legend_title = legend_title, before_id = "water") |>
    mapgl::add_line_layer(id = "disp_hatch", source = hatch,
                          line_color = "#c1272d", line_width = 0.9,
                          line_opacity = 0.85) |>
    nt_declutter_basemap()   # drop park/landcover tints that wash the fills
}
map_row <- function(x, region, group) tagList(
  nt_map_sync(
    mk_map(x$g1, "v1 classes (per-100k rate)", x$hatch),
    mk_map(x$g2, "same classes, corrected engine", x$hatch),
    mk_map(x$g3, "VLI verdict (30%/50% burden)", x$hatch),
    group = group, height = 480,
    titles = c(paste0("v1 — as published (", region, ")"),
               "v2 — same classes, corrected engine",
               "v2 — the verdict, VLI (0–50% AMI)"),
    subtitles = c(
      "Limited Renter Affordability, 0–80% AMI: legacy afford() rate classes (accessible rentals per 100k low-income households: ≤100 / 100–200 / >200)",
      "identical cutpoints on the rebuilt rate: HUD income limits (v1's were 15–26% low) + renter-matched need",
      "the pilot group's headline map: what the typical unit costs a VLI household (majority rule)")),
  div(class = "hatch-note",
      "Red hatching (all panels) = displacement: any income group at risk (HPRM v5 2022 dis_group). Pan or zoom any map — the row moves together; hover a tract for its class."))

xtab <- function(g) {
  sf::st_drop_geometry(g) |> filter(!is.na(cls)) |>
    group_by(class = cls) |>
    summarize(tracts = dplyr::n(),
              hatched = sum(!is.na(dis_group)),
              `share hatched` = scales::percent(mean(!is.na(dis_group)), 1),
              .groups = "drop")
}

sd_  <- build("06", "073")
slc_ <- build("49", "035")
```

**What this shows.** Each region three times, in the visual grammar of the published EDR reports
([SANDAG](https://urban-displacement.github.io/edr-sandag/),
[Salt Lake City](https://urban-displacement.github.io/edr-ut/slc_edr_report/)): the *"Limited
Renter Affordability"* fill with the red **displacement hatch** on top. Left = **v1 exactly as
published** (0–80% AMI, the rate cutpoints its code drew the report maps with). Middle = **the
same map with only the engine corrected** — identical class cutpoints, but HUD's actual income
limits (v1's were 15–26% too low) and renter-household need. Right = **the v2 verdict for the
pilot VLI group** — the new presentation (burden classes), same hatch. **The three maps in each
row are synchronized** (the package's `nt_map_sync()`) — pan or zoom any one and the others
follow, so you can inspect an area of interest across all three at once.

*Method notes: the published reports ran on report-era EDR (SLC: 2015–2019 data); the hatch here
is HPRM v5 2022 — compare the pattern, not exact tract sets. Salt Lake is Salt Lake County only
(the published Wasatch region adds Weber/Davis/Utah counties; the 2026-07-02 HUD API outage
limits v2 to snapshot-exact counties). See `dev/affordability-deep-dive.md` §2–2.3 and
`dev/slc_sd_replication.R`.*

## San Diego County

```{r sd}
map_row(sd_, "San Diego", "sd")
```

```{r sd_tab}
knitr::kable(lapply(sd_[c("g1","g2","g3")], xtab),
  caption = "San Diego — v1 as published / v2 same classes / v2 VLI verdict, against the displacement hatch")
```

## Salt Lake County

```{r slc}
map_row(slc_, "Salt Lake", "slc")
```

```{r slc_tab}
knitr::kable(lapply(slc_[c("g1","g2","g3")], xtab),
  caption = "Salt Lake — v1 as published / v2 same classes / v2 VLI verdict, against the displacement hatch")
```

## How to read the difference

- **The hatch is the same in every panel** — displacement doesn't move; the fill under it does.
- **Left (v1): the published finding is visible and extreme.** In San Diego, v1 put just
  **16 tracts** in "More Affordable" — and **16 of 16 sit under the displacement hatch** (Salt
  Lake: 115 green, 61% hatched, against 3–7% in the other classes). That is the reports'
  *"no 'more affordable' neighborhoods outside displacement-risk areas."*
- **Middle: correct only the engine and the sliver becomes a real class.** Same cutpoints, HUD
  cutoffs, renter-matched need: San Diego's green class grows **16 → 207**. The hatch still
  concentrates on the green/yellow (reachable) side — the *gradient* survives the correction
  (share hatched climbs 20% → 60% → 90% across the classes); the *"every affordable tract is
  displacing"* absolutism does not. The published green-sliver effect was substantially the
  artifact of income cutoffs 15–26% below HUD's.
- **Right: the verdict for the group decisions are made about (VLI).** Plain-language classes
  (affordable / roughly / not at 30%/50% of income), and displacement again lands almost
  exclusively in the reachable classes.
- **Why not the verdict at 0–80% AMI?** It saturates by construction: the 0–80% band's *ceiling*
  income affords the typical unit in most tracts (San Diego: `r sd_$li_verdict[["affordable"]]`
  of ~730 tracts read "affordable"), because a burden verdict answers *"can the band's top
  afford here?"* while the reports' rate classes answer *"is there enough stock for the size of
  the need?"* Different constructs: v2 carries both (`verdict` and `ratio`/`rate`), and for broad
  bands the scarcity-vs-need rate (middle panel) is the right report-style layer, while the
  verdict is sharpest for the tiers decisions target (VLI/ELI).

---

# Map 3 alone — the VLI verdict across four more markets

```{r regions_data, include=FALSE}
pe <- readRDS("p1_eval_data.rds")   # Bay + King: verdicts already computed
                                    # (hud / hud_acs with the FMR-era cutoffs pulled 2026-07-01)
vcls <- function(verdict) factor(case_when(
  verdict == "not affordable"     ~ lev2[1],
  verdict == "roughly affordable" ~ lev2[2],
  verdict == "affordable"         ~ lev2[3]), levels = lev2)
prep_vli <- function(sfdat) {
  v <- sfdat[sfdat$ami_tier == "VLI" & !is.na(sfdat$verdict), ]
  v$cls <- vcls(v$verdict)
  simp(v[, c("GEOID", "cls")])
}
verdict_map <- function(v, height = 560) {
  m <- nt_map(v, "cls", type = "categorical", colors = pal,
              legend_title = "verdict for a VLI renter", before_id = "water") |>
    mapgl::add_line_layer(id = "disp_hatch",
      source = hatch_lines(v[v$GEOID %in% h$GEOID[!is.na(h$dis_group)], ]),
      line_color = "#c1272d", line_width = 0.9, line_opacity = 0.85) |>
    nt_declutter_basemap()   # drop park/landcover tints that wash the fills
  m$height <- height
  m
}
bay_v  <- prep_vli(pe$bay)
king_v <- prep_vli(pe$king)
tc_raw <- suppressMessages(afford_index("27",
  c("Hennepin", "Ramsey", "Anoka", "Dakota", "Washington", "Scott", "Carver"),
  2024, tenure = "rent", ami_source = "hud_acs", geometry = TRUE)) |> afford_verdict()
tc_v   <- prep_vli(tc_raw)
atl_raw <- suppressMessages(afford_index("13", c("121", "089", "067", "135"),
  2024, tenure = "rent", ami_source = "hud", geometry = TRUE)) |> afford_verdict()
atl_v  <- prep_vli(atl_raw)
vc <- function(v) table(v$cls)   # counts in lev2 order: not / roughly / affordable
```

**The reading.** *This is how much a VLI household will pay if they can move in there*:
**green** — the typical unit takes **no more than 30%** of their income; **orange** — **30–50%**
(attainable only by stretching); **red** — **more than half their income**. For the red class,
the descriptor:

> **Even at the rents current tenants pay, a VLI family would give up more than half their
> income in these neighborhoods — and entry prices run higher still.**

**The "what ifs" this map assumes — read them with every figure:**

- **"If they can move in."** The map is price-only. It says nothing about whether a unit is
  *open* (availability — the next pane) or whether the applicant is *accepted* (records, credit,
  vouchers — screening). A green tract with no vacancies admits no one.
- **At incumbent prices.** Classes are priced at what *current tenants* pay (ACS gross rent,
  utilities included). Asking rents on the units that actually open run higher, so the map
  **understates** the entry barrier — the constraint is at least as bad as shown.
- **At the top of the VLI range.** The share is computed at the VLI income ceiling
  (household-size-adjusted HUD limit). Households deeper in the band give up a larger share of
  income for the same unit — green is "manageable" for the best-off VLI household.

Red hatching on every map = displacement (any income group at risk, HPRM v5 2022), as above.

### 9-county San Francisco Bay Area

**`r vc(bay_v)[[3]]` affordable / `r vc(bay_v)[[2]]` roughly affordable / `r vc(bay_v)[[1]]` not
affordable.** *(Cutoffs: the validated ACS reproduction of the HUD limits, including the
high-cost VLI bump, pulled 2026-07-01.)*

```{r bay_verdict}
verdict_map(bay_v)
```

### Twin Cities, MN — 7-county metro (Hennepin, Ramsey, Anoka, Dakota, Washington, Scott, Carver)

**`r vc(tc_v)[[3]]` affordable / `r vc(tc_v)[[2]]` roughly affordable / `r vc(tc_v)[[1]]` not
affordable.** *(Cutoffs: `hud_acs` — the HUD-outage fallback. In this ordinary-cost market the
approximation converges with HUD (QA §6, Hennepin), and the FMR bump the outage disables does
not bind here; suburban county-vs-metro divergence is possible — re-run on exact `hud` when the
API returns.)*

```{r tc_verdict}
verdict_map(tc_v)
```

### King County, WA

**`r vc(king_v)[[3]]` affordable / `r vc(king_v)[[2]]` roughly affordable / `r vc(king_v)[[1]]`
not affordable.** *(Cutoffs: exact HUD snapshot.)*

```{r king_verdict}
verdict_map(king_v)
```

### Atlanta metro, GA — Fulton, DeKalb, Cobb, Gwinnett

**`r vc(atl_v)[[3]]` affordable / `r vc(atl_v)[[2]]` roughly affordable / `r vc(atl_v)[[1]]` not
affordable.** *(Cutoffs: exact HUD snapshot.)*

```{r atl_verdict}
verdict_map(atl_v)
```

---

# The three panes: affordability → availability → stability

```{r pane2_data, include=FALSE}
options(timeout = 1200)                      # first-run block downloads are large
zori   <- afford_zori(quiet = TRUE)          # Zillow ZIP asking rents (cached)
slim <- function(v) {                        # simplify geometry, keep all columns
  g <- sf::st_transform(sf::st_geometry(v), 3857)
  sf::st_geometry(v) <- sf::st_transform(
    sf::st_simplify(g, dTolerance = 60, preserveTopology = TRUE), 4326)
  v
}
w_king <- nt_zcta_weights("53", "033")       # block-exact ZCTA x tract weights
# pane 2 (entry) then pane 3 (stability: HPRM displacement + eviction anchors)
h_stab <- arrow::read_parquet(
  "/Users/buffalo/data/evictionresearch/hprm/hprm_v5_full_2022.parquet",
  col_select = c("geoid", "dis_value", "ev_value"))
king3  <- afford_stability(slim(afford_entry(pe$king, zori, weights = w_king)), h_stab)
# the other three markets: afford_entry() derives counties from the data and
# builds its own weights (block pulls are cached by tigris)
bay3   <- afford_stability(slim(afford_entry(pe$bay, zori)), h_stab)
tc3    <- afford_stability(slim(afford_entry(tc_raw, zori)), h_stab)
atl3   <- afford_stability(slim(afford_entry(atl_raw, zori)), h_stab)
k_vli  <- king3[king3$ami_tier == "VLI", ]
kv1 <- table(k_vli$verdict); kv2 <- table(k_vli$entry_verdict)
kv3 <- table(k_vli$stable_cat)
prem_med <- median(k_vli$entry_premium, na.rm = TRUE)
prem_p90 <- quantile(k_vli$entry_premium, 0.9, na.rm = TRUE)
n_fallback <- sum(k_vli$premium_fallback, na.rm = TRUE)

# one synced three-pane row (affordability / entry / stability) per region;
# the hatch stays on the price panes only -- pane 3's fill IS the risk read
pane3_row <- function(x3, region, group) {
  v <- x3[x3$ami_tier == "VLI", ]
  k1 <- table(v$verdict); k2 <- table(v$entry_verdict); k3 <- table(v$stable_cat)
  nt_map_sync(
    afford_map(x3, "VLI", displacement = h,
               legend_title = "verdict at incumbent prices"),
    afford_map(x3, "VLI", verdict_col = "entry_verdict", displacement = h,
               legend_title = "entry verdict (asking prices)"),
    afford_map(x3, "VLI", verdict_col = "stable_cat"),
    group = group, height = 500,
    titles = c(sprintf("%s — 1 · affordability", region),
               "2 · availability at entry",
               "3 · stability"),
    subtitles = c(
      sprintf("what sitting tenants pay — a / r / n: %d / %d / %d",
              k1[[1]], k1[[2]], k1[[3]]),
      sprintf("what the door costs today — a / r / n: %d / %d / %d (median premium %.2fx)",
              k2[[1]], k2[[2]], k2[[3]], median(v$entry_premium, na.rm = TRUE)),
      sprintf("can they stay? (HPRM; no hatch — the fill is the risk) — s / e / p: %d / %d / %d",
              k3[[1]], k3[[2]], k3[[3]])))
}
pane2_cap <- function(x3) {
  cp <- afford_capacity(x3)
  knitr::kable(dplyr::transmute(cp[cp$ami_tier == "VLI", ],
    `affordable per 100 (standing)`  = round(per100_affordable, 1),
    `open per year, per 100`         = round(per100_open_year, 1),
    `affordable per 100 (at entry)`  = round(per100_entry, 1),
    `open AND entry-affordable, per 100/yr` = round(per100_open_entry_year, 1),
    `...AND stability-weighted, per 100/yr` = round(per100_open_entry_stable, 1)))
}
p2stats <- function(x3) {
  v <- x3[x3$ami_tier == "VLI", ]
  cp <- afford_capacity(x3)
  list(prem = median(v$entry_premium, na.rm = TRUE),
       fb = sum(v$premium_fallback, na.rm = TRUE), n = nrow(v),
       v1 = table(v$verdict), v2 = table(v$entry_verdict),
       s = table(v$stable_cat),
       cap = as.data.frame(cp[cp$ami_tier == "VLI", ]))
}
k_cap <- p2stats(king3)$cap
bs <- p2stats(bay3); ts <- p2stats(tc3); as_ <- p2stats(atl3)
saveRDS(list(king = p2stats(king3), bay = bs, tc = ts, atl = as_),
  "/private/tmp/claude-501/-Users-buffalo-git-evictionresearch-neighborhood/fece173f-44e2-4643-a35a-5414bacdc86b/scratchpad/pane2_regions.rds")
```

Pane 1 prices each tract at what **sitting tenants** pay. Pane 2 re-prices it at **entry**: every
tract gets an **entry premium** — Zillow's ZIP-level asking-rent index (`afford_zori()`, month
`r unique(zori$month)`) over the same ZIP's standing rents, brought onto tracts with block-exact
housing-unit weights (`nt_zcta_weights()`) — and the tract's rent distribution is shifted by it
before the same 30%/50% verdict is applied (`afford_entry()`). King County's median entry premium
is **`r sprintf("%.2f", prem_med)`x** (90th percentile `r sprintf("%.2f", prem_p90)`x;
`r n_fallback` of `r nrow(k_vli)` tracts fell back to the area-median premium for thin listing
coverage). The third map is **stability** (`afford_stability()`: HPRM displacement + eviction
risk on the published anchors, equal-weighted): whoever gets in — can they *stay*? Each row of
three is synced — pan to any neighborhood and read all three at once.

```{r pane2_maps}
pane3_row(king3, "King County", "king2")
```

```{r pane2_caption}
afford_caption("VLI", pane = "availability")
```

```{r pane3_caption}
afford_caption("VLI", pane = "stability")
```

**The capacity funnel (VLI renters, King County):**

```{r pane2_cap}
pane2_cap(king3)
```

Read the row left to right as the funnel: *of every 100 VLI renter households in King County,
the standing stock could house `r round(k_cap$per100_affordable)` at ≤ 30% of income; only
`r round(k_cap$per100_open_year)` of those units open in a year;
`r round(k_cap$per100_open_entry_year)` open **and** rent at ≤ 30% at today's asking prices; and
weighting by stability leaves about `r round(k_cap$per100_open_entry_stable)` openings a year in
neighborhoods a household can also hold onto* — before competition (a cheap unit occupied by a
higher-income household still counts; the CHAS layer is next) and before screening. One honest
note: the earlier ZIP-level teaser (6 of 69 ZIPs asking-affordable) compared listing *levels* to
the cost line; this model shifts each tract's own rent *distribution* by its premium, which is
the defensible estimate — the wedge is real
(green −`r round(100 * (1 - kv2[[1]] / kv1[[1]]))`%, red +`r round(100 * (kv2[[3]] / kv1[[3]] - 1))`%)
but smaller than the teaser implied. Method, validation plan, and caveats:
`dev/availability-design.md`.

## The same three panes, three more markets

Same construction everywhere (each region's own ZIP premiums, block-exact weights, the same
verdict and stability grammar; each row is synced independently).

### 9-county Bay Area

*(Median entry premium `r sprintf("%.2f", bs$prem)`x; `r bs$fb` of `r bs$n` tracts on the
fallback premium. Cutoffs: the FMR-era ACS-HUD reproduction, pulled 2026-07-01. Stability note:
California's eviction signal (EER) is out-of-sample — the displacement side carries the read.)*

```{r pane2_bay}
pane3_row(bay3, "Bay Area", "bay2")
```

```{r pane2_bay_cap}
pane2_cap(bay3)
```

### Twin Cities, MN — 7-county metro

*(Median entry premium `r sprintf("%.2f", ts$prem)`x; `r ts$fb` of `r ts$n` tracts on the
fallback premium. Cutoffs: `hud_acs`, the HUD-outage fallback — converges with HUD in this
ordinary-cost market.)*

```{r pane2_tc}
pane3_row(tc3, "Twin Cities", "tc2")
```

```{r pane2_tc_cap}
pane2_cap(tc3)
```

### Atlanta metro, GA — Fulton, DeKalb, Cobb, Gwinnett

*(Median entry premium `r sprintf("%.2f", as_$prem)`x; `r as_$fb` of `r as_$n` tracts on the
fallback premium. Cutoffs: exact HUD snapshot.)*

```{r pane2_atl}
pane3_row(atl3, "Atlanta", "atl2")
```

```{r pane2_atl_cap}
pane2_cap(atl3)
```
