---
title: "The Affordability Index: methodology and interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Affordability Index: methodology and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What the index is for

The affordability index answers one question at the census-tract scale:

> **For households at a given income level, how much of a tract's housing is within reach,
> and how does that compare to the region's share of households at that income level?**

It is the housing-supply counterpart to the Eviction Research Network's displacement and
eviction-risk models. Conceptually it supports the *highways of migration* idea: people priced
out of one place can only move to the few places that can actually house them. The index maps
where that affordable housing is — and, by comparing supply to the regional count of
same-income households, where it is scarce relative to need.

> **Note — two implementations.** `afford_index()` is the rebuilt v2 engine and the recommended
> entry point: it fixes the AMI basis (county median *family* income), interpolates within income
> brackets, supports standardized ELI/VLI/LI/MI tiers and household sizes, and offers both owner
> cost bases (current cost vs. cost to buy in). `afford()` is the original, kept for backward
> compatibility. This vignette documents the legacy `afford()`; see `?afford_index` and the design
> notes in `dev/affordability-index-design.md` for v2.

`afford()` is the original implementation. **Read the "Assumptions and known limitations"
section below before using its output in a report** — the ownership numbers in particular carry a
hard-coded interest-rate assumption, and a fuller methodological review with a remediation plan
lives in [`dev/affordability-index-review.md`](https://github.com/evictionresearch/neighborhood).

## Quick start

```{r}
library(neighborhood)

# 5-county Bay Area, households at 50% of Area Median Income, 2024 ACS
bay <- afford(
  state    = "06",
  counties = c("001", "013", "041", "075", "081"),
  ami_limit = 0.5,
  year     = 2024
)

# With tract geometry for mapping (returns an sf object)
wasatch <- afford(
  state    = "49",
  counties = c("011", "035", "049", "057"),
  ami_limit = 0.5,
  year     = 2024,
  geometry = TRUE
)
```

`ami_limit` is the income level of interest as a fraction of Area Median Income: `0.3` ≈
extremely low income, `0.5` ≈ very low income, `0.8` ≈ low income. **Use FIPS codes** for
`state` and `counties` (e.g. `"06"`, `"041"`), not names — see limitations.

## How it is calculated

For a chosen income level `ami_limit`, the index is built in four steps.

### 1. Set the income threshold

```
AMI        = median of tract median household incomes in the region   (B19013_001)
threshold  = ami_limit * AMI                                          (e.g. 0.5 * AMI)
```

All affordability is judged against this single regional `threshold`.

### 2. Translate housing costs into the income needed to afford them

Each housing-cost bracket from the ACS is converted into the household income required to
afford it without spending more than **30%** of income on housing:

| Tenure | Source | Cost → required income |
|---|---|---|
| **Rental** | gross rent, `B25063` | `income = (monthly_rent / 0.30) * 12` |
| **Ownership** | value `B25075` + price asked `B25085` | `income = price * 0.188` |

The rental rule is the standard 30%-of-gross-rent affordability test. The ownership factor
`0.188` is a *fixed* approximation of the income needed to carry a home of a given price (see
limitations — it embeds an interest-rate assumption).

### 3. Count what each tract offers the income group

Within each tract, a unit is **accessible** to the group if the income it requires is at or
below the `threshold`. The index then forms, per tract and per tenure:

```
supply = accessible_units / total_units_in_tract
       = "what share of this tract's stock is affordable to the group"
```

### 4. Compare supply to regional need

Let `class_prop` be the region's share of households at or below the threshold. Then:

```
ratio = supply / class_prop
      = location quotient: >1 means the tract carries MORE than its
        "fair share" of affordable housing for this income group; <1 means less

rate  = (accessible_units / regional_group_households) * 100000
      = accessible units per 100,000 of the region's same-income households
```

`ratio` is the analytical heart of the index — it is the same idea the EDR reports lead with
("fewer than one affordable rental unit per three extremely-low-income households").

## Reading the output

`afford()` returns one row per tract. The most useful columns:

| Column | Meaning |
|---|---|
| `tr_rent_total`, `tr_own_total` | rental / owned units in the tract |
| `tr_rent_accessible`, `tr_own_accessible` | units affordable to the income group |
| `tr_rent_supply`, `tr_own_supply` | accessible ÷ total (share affordable) |
| `tr_rent_ratio`, `tr_own_ratio` | supply ÷ regional group share (**location quotient**) |
| `tr_rent_rate`, `tr_own_rate` | accessible per 100k regional group-households |
| `reg_total_pop`, `reg_class_pop` | region households (all / in the income group) |
| `ami_limit` | the income level used |
| `*_cat` | 3-class map categories (fixed cut points) |
| `popup` | preformatted HTML for `leaflet` maps |

A simple read: tracts with **`ratio` well above 1** are the region's affordable-housing
reservoirs for that income group; tracts with **`ratio` near 0** cannot house the group at all.
Overlaying low-`ratio` tracts with displacement-risk tracts reveals the migration squeeze the
theory predicts — *nowhere affordable to move that isn't already under pressure.*

## Assumptions and known limitations

These are summarized here so results are not over-read; the full treatment and fixes are in
`dev/affordability-index-review.md`.

1. **Ownership affordability embeds a fixed interest rate.** The `0.188` factor corresponds to a
   ~3.8%, 30-year, 0%-down mortgage with no taxes/insurance. At 2024 rates (~7%) it **overstates**
   ownership affordability by roughly 30–40%. Treat `tr_own_*` columns as provisional.
2. **"AMI" is the median of tract medians**, not HUD AMI and not a population-weighted regional
   median. It also collapses multi-county regions into a single cutoff.
3. **Occupied stock, not availability.** The index scores the affordability of *existing* units,
   not whether units are *vacant/available to move into*. A tract can be "affordable" yet have
   nowhere to actually go.
4. **Price/value mixes two census universes** (owner-occupied value + vacant-for-sale price asked).
5. **No non-price barriers.** Discrimination, voucher rejection, and eviction-record/credit
   screening ("the scarlet E") are not represented — though they are central to who can *actually*
   move where.
6. **Margins of error are dropped.** Small-denominator tracts (e.g. 1 of 2 units affordable) are
   presented with the same confidence as large ones.
7. **Inputs must be FIPS codes.** Passing `state = "CA"` / `county = "Marin"` (names) returns an
   empty internal join and silently distorts the regional totals.

## Data sources

All inputs are American Community Survey 5-year estimates via `tidycensus::get_acs()`; a
`CENSUS_API_KEY` must be set. Tables used: `B19013` (median income), `B19001` (income
distribution), `B25075`/`B25085` (owner value / price), `B25063` (gross rent). Geometry, when
requested, comes from `tigris::tracts()`.

## See also

- `dev/affordability-index-review.md` — full methodological review and remediation plan.
- `evictionresearch/library/code/d_calc_ami_incomplete.R` — the organization's HUD-style AMI tier
  engine (county AMI + within-bracket interpolation) that future versions should build on.
- `ntdf()` — neighborhood racial typologies, joinable to the index on `GEOID` for equity analysis.
