---
title: "The Affordability Index: methodology and interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Affordability Index: methodology and interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)

# fig() embeds a pre-rendered PNG from vignettes/figures/ if it exists, and
# silently renders nothing if it is missing -- so a stale figure never breaks
# the package build.
fig <- function(f) {
  p <- file.path("figures", f)
  if (file.exists(p)) knitr::include_graphics(p)
}
```

## What the index is for

The affordability index answers one question at the census-tract scale:

> **For households at a given income level, how much of a tract's housing is within reach,
> and how does that compare to the region's share of households at that income level?**

It is the housing-supply counterpart to the Eviction Research Network's displacement and
eviction-risk models. Conceptually it supports the *highways of migration* idea: people priced
out of one place can only move to the few places that can actually house them. The index maps
where that affordable housing is — and, by comparing supply to the regional count of
same-income households, where it is scarce relative to need.

In the larger design (`dev/affordability-index-design.md`) affordability is **Gate 1** of a
three-gate "choice" funnel — *Afford → Available → Allowed*. This vignette documents Gate 1, the
supply engine, which is the part that ships today.

> **Note — two implementations, lead with the new one.** `afford_index()` is the rebuilt **v2
> engine** and the recommended entry point. `afford()` is the original v1 implementation, kept for
> backward compatibility and documented in an appendix at the end. New work should use
> `afford_index()`; the v1→v2 crosswalk below explains exactly what it fixes.

---

# `afford_index()` — the recommended engine

`afford_index()` rebuilds the index on a defensible footing: AMI from county median **family**
income (HUD's anchor), income brackets parsed from ACS variable *labels* with within-bracket
interpolation, standardized ELI/VLI/LI/MI tiers, HUD household-size adjustment, and explicit cost
models for both renting and owning. Its output is **tidy and long**: one row per
tract × tenure × tier.

All inputs are American Community Survey 5-year estimates via `tidycensus::get_acs()`, so a
`CENSUS_API_KEY` must be set. The `"hud"` AMI source additionally needs the `hudr` package and a
free `HUD_API_KEY`. Because every example below touches the Census (and sometimes the HUD) API, the
code chunks are shown but **not executed** at build time.

## Quick start

The very simplest call needs only `state` and `counties` (FIPS codes or names both work). Everything
else has a default: `year = 2024`, all three tenures, all four tiers, the pure-ACS AMI source.

```{r}
library(neighborhood)

# San Francisco County, defaults for everything else
sf_idx <- afford_index("06", "075")

# Look at one slice: very-low-income (50% AMI) rental supply, per tract
sf_idx[sf_idx$tenure == "rent" & sf_idx$ami_tier == "VLI",
       c("GEOID", "ami", "income_cutoff", "accessible", "total", "supply", "ratio")]
```

`state` and `counties` accept either FIPS codes (`"06"`, `"075"`) **or** names (`"CA"`,
`"San Francisco"`) — unlike legacy `afford()`, which only takes FIPS. A multi-county region is just
a vector of counties:

```{r}
# 5-county Bay Area core, 2023 ACS
bay <- afford_index(
  state    = "CA",
  counties = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo"),
  year     = 2023
)
```

## The full argument reference

Every argument is a real formal of `afford_index()` (defaults shown). The mortgage knobs in the
lower block only affect the `"own_buyin"` tenure.

| Argument | Default | What it does |
|---|---|---|
| `state` | — (required) | State FIPS (`"06"`) or name/abbreviation (`"CA"`). |
| `counties` | — (required) | One or more 3-digit county FIPS or names. |
| `year` | `2024` | ACS 5-year endpoint year. |
| `tenure` | `c("rent","own_current","own_buyin")` | Which tenures to compute (any subset; see below). |
| `ami_source` | `"acs"` | `"acs"`, `"hud"`, or `"hud_acs"` (the last not yet implemented). |
| `ami_tiers` | `c(ELI=.30, VLI=.50, LI=.80, MI=1.20)` | Named tier fractions of AMI. |
| `hud_hh_size` | `4` | Household size 1–8 for the family-size adjustment; `NULL` = unadjusted. |
| `interest_rate` | `NULL` | Annual mortgage rate for `own_buyin`; `NULL` = built-in by-year value. |
| `term_years` | `30` | `own_buyin` mortgage term. |
| `down_pct` | `0.10` | `own_buyin` down payment fraction. |
| `tax_ins_rate` | `0.0125` | `own_buyin` annual taxes + insurance as a fraction of value. |
| `geometry` | `FALSE` | If `TRUE`, attach tract polygons (an `sf` tibble). |

### `tenure` — three housing-cost models

`tenure` is the most important conceptual choice. Renting has one model; owning has **two**, because
"can you afford the homes here?" has two honest meanings.

```{r}
# Just renters
afford_index("06", "075", tenure = "rent")

# Renters plus the "could a mover buy in today?" owner basis
afford_index("06", "075", tenure = c("rent", "own_buyin"))
```

| `tenure` value | ACS table | Cost model | Question it answers |
|---|---|---|---|
| `"rent"` | `B25063` (gross rent) | affordable if `rent <= income_cutoff * 0.30 / 12` | What share of *rentals* is affordable to the tier? |
| `"own_current"` (basis A) | `B25094` (selected monthly owner cost) | same 30%-of-income rule as rent | Affordability of the *existing* owned stock as currently financed (most owners locked in old, low payments). |
| `"own_buyin"` (basis B) | `B25075` (home value) | value → required income via an explicit mortgage model | Could a *mover* afford to **buy in here today** at this year's rate? |

The distinction matters: `own_current` describes who could carry the homes people already own
(often very affordable thanks to old mortgages), while `own_buyin` describes the true barrier facing
someone trying to move in now. Reports about *displacement* and *migration* almost always want
`own_buyin`.

### `ami_source` — where the income limits come from

```{r}
# Pure ACS (default): AMI = county median FAMILY income (B19113)
afford_index("06", "075", ami_source = "acs")

# Official HUD published limits (needs hudr + HUD_API_KEY)
afford_index("06", "075", ami_source = "hud")
```

- **`"acs"`** (default): AMI is the county median **family** income (ACS `B19113`), and each tier is
  a simple fraction of it. Fully reproducible from Census alone.
- **`"hud"`**: ELI/VLI/LI come straight from HUD's *published* per-household-size limits (so HUD's
  caps, floors, and high-cost adjustments are baked in); any tier without a HUD limit (e.g.
  `MI = 120%`) is *derived* as fraction × area median × size factor. Requires the `hudr` package and
  a `HUD_API_KEY` (register free at huduser.gov; e.g. in `~/.Renviron`), unless a HUD snapshot is
  bundled with the package.
- **`"hud_acs"`**: the full HUD-cascade reproduction; **not yet implemented** — calling it errors
  with a pointer to `dev/hud-income-limits-architecture.md`.

### `ami_tiers` — which income tiers to compute

A named numeric vector of AMI fractions. The defaults are the standard HUD-style cumulative ceilings;
a VLI household can use anything affordable at or below 50% AMI.

```{r}
# Default tiers, named explicitly
afford_index("06", "075",
             ami_tiers = c(ELI = 0.30, VLI = 0.50, LI = 0.80, MI = 1.20))

# A custom, coarser set
afford_index("06", "075", ami_tiers = c(VeryLow = 0.50, Moderate = 1.00))
```

The names become the `ami_tier` factor levels in the output (in the order you supply), and each
tier appears as its own block of rows per tract.

### `hud_hh_size` — family-size adjustment

HUD scales income limits by household size (a 4-person household is the unadjusted basis = 1.00).
`hud_hh_size` applies that factor to the tier cutoffs; pass `NULL` to leave AMI on the 4-person
basis.

```{r}
# A single-person household (limits scaled to 0.70 of the 4-person basis)
afford_index("06", "075", hud_hh_size = 1)

# No size adjustment at all
afford_index("06", "075", hud_hh_size = NULL)
```

The factors are HUD's: `1 = 0.70, 2 = 0.80, 3 = 0.90, 4 = 1.00, 5 = 1.08, 6 = 1.16, 7 = 1.24,
8 = 1.32`. Values outside 1–8 error.

### The `own_buyin` mortgage knobs

These four arguments define the explicit mortgage used to convert a home's *value* into the income
needed to **buy in today**. They are inert for `"rent"` and `"own_current"`.

The model: a household spends at most 30% of income on (principal & interest on the financed portion)
plus (taxes & insurance on the full value). Lower the rate or raise the down payment and more homes
become "affordable."

```{r}
# Spell out today's terms instead of the built-in by-year rate
afford_index(
  "06", "075",
  tenure        = "own_buyin",
  interest_rate = 0.069,   # 6.9% annual; NULL uses a built-in per-year approximation
  term_years    = 30,      # loan term in years
  down_pct      = 0.20,    # 20% down
  tax_ins_rate  = 0.0125   # 1.25% of value/yr for taxes + insurance
)
```

| Knob | Default | Effect |
|---|---|---|
| `interest_rate` | `NULL` → built-in by-year (~6.9% for 2024) | higher rate → fewer homes affordable |
| `term_years` | `30` | longer term → slightly more affordable |
| `down_pct` | `0.10` | bigger down payment → smaller loan → more affordable |
| `tax_ins_rate` | `0.0125` | higher carrying cost → fewer homes affordable |

The built-in `interest_rate = NULL` default looks up a rough 30-year average for the requested
`year` (e.g. ~3.1% for 2020, ~6.9% for 2024), so even the bare call is rate-aware — but for a report,
state the rate explicitly so the assumption is auditable.

### `geometry` — attach polygons for mapping

```{r}
# Returns an sf tibble (tract polygons joined to every tenure/tier row)
sf_geo <- afford_index("06", "075", geometry = TRUE)
```

Because the output is long, `geometry = TRUE` **duplicates each tract polygon** across every
tenure × tier combination. Always filter to **one** tenure and **one** tier before mapping (see
*Mapping the index* below).

## The output schema (tidy long)

`afford_index()` returns one row per **tract × tenure × tier**. Verified column order and meaning:

| Column | Meaning |
|---|---|
| `GEOID` | 11-digit tract FIPS |
| `county` | 5-digit county FIPS |
| `year` | ACS endpoint year used |
| `ami_source` | `"acs"` / `"hud"` |
| `tenure` | `"rent"` / `"own_current"` / `"own_buyin"` |
| `ami_tier` | factor: `ELI`/`VLI`/`LI`/`MI` (or your custom names) |
| `ami` | the county's AMI (median family income basis) |
| `income_cutoff` | the tier's income ceiling in dollars (this is `cut_<TIER>` from `ami_cutoffs()`) |
| `accessible` | interpolated count of units affordable to the tier in this tract/tenure |
| `total` | total units of this tenure in the tract |
| `supply` | `accessible / total` — share of this tract's stock affordable to the tier |
| `reg_hh_tier` | regional households in the tier (interpolated demand) |
| `reg_hh_total` | regional households, all tiers |
| `class_prop` | `reg_hh_tier / reg_hh_total` — region's share of households in the tier |
| `ratio` | `supply / class_prop` — **location quotient** (>1 = more than fair share) |
| `rate` | `accessible / reg_hh_tier * 100000` — accessible units per 100k tier households |

A quick read of the analytical columns:

```{r}
sf_idx <- afford_index("06", "075")

# Where is VLI rental supply most over- or under-represented?
vli_rent <- sf_idx[sf_idx$tenure == "rent" & sf_idx$ami_tier == "VLI", ]
head(vli_rent[order(-vli_rent$ratio), c("GEOID", "supply", "class_prop", "ratio", "rate")])
```

- **`supply`** is the within-tract share affordable to the tier.
- **`ratio`** is the analytical heart: `> 1` means the tract carries *more* than its fair share of
  affordable housing for the tier; near `0` means it cannot house the tier at all. It is the same
  idea the EDR reports lead with ("fewer than one affordable unit per N tier households").
- **`rate`** rescales `accessible` to a per-100k-households figure for cross-region comparison.

Reshape to one row per tract when you want side-by-side tenures/tiers:

```{r}
library(tidyr)
wide <- tidyr::pivot_wider(
  sf_idx[, c("GEOID", "tenure", "ami_tier", "ratio")],
  names_from = c(tenure, ami_tier), values_from = ratio
)
```

## Mapping the index

`nt_add_choropleth()` expects **one polygon per tract**, so filter the long `sf` output to a single
tenure and tier first, then pipe through the MapLibre helpers (`vignette("mapping-with-maplibre")`).

```{r}
library(neighborhood)

idx <- afford_index("06", "075", geometry = TRUE)

# ONE tenure + ONE tier => one polygon per tract
vli_rent <- idx[idx$tenure == "rent" & idx$ami_tier == "VLI", ]

nt_map(vli_rent) |>
  nt_add_choropleth(
    data         = vli_rent,
    column       = "ratio",
    type         = "continuous",
    legend_title = "VLI rental supply ratio"
  )
```

```{r afford-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Very-low-income (VLI) rental supply ratio by tract, San Francisco — below 1 means fewer affordable units than VLI renter households."}
# Pre-rendered if available; renders nothing otherwise (build-safe).
fig("map-afford-ratio.png")
```

---

# `ami_cutoffs()` — the income ceilings underneath

`afford_index()` calls `ami_cutoffs()` internally, but it is exported so you can inspect the dollar
thresholds the whole index hinges on. It returns one row **per county**, with each tier's income
ceiling.

```{r}
ami_cutoffs("06", c("075", "081"), 2024)
#> # A tibble: 2 x 7
#>   GEOID NAME                          ami cut_ELI cut_VLI cut_LI cut_MI
#>   <chr> <chr>                       <dbl>   <dbl>   <dbl>  <dbl>  <dbl>
#> ...
```

### Arguments

`ami_cutoffs()` shares its income-side arguments with `afford_index()` (it has no tenure or mortgage
arguments — it is purely the income limits):

| Argument | Default | What it does |
|---|---|---|
| `state` | — (required) | State FIPS or name. |
| `counties` | — (required) | County FIPS or names. |
| `year` | `2024` | ACS 5-year endpoint year. |
| `ami_source` | `"acs"` | `"acs"` / `"hud"` / `"hud_acs"` (see above). |
| `ami_tiers` | `c(ELI=.30, VLI=.50, LI=.80, MI=1.20)` | Named tier fractions of AMI. |
| `hud_hh_size` | `4` | Household size 1–8 (or `NULL` for unadjusted). |

```{r}
# A 2-person household, just the ELI and VLI ceilings, official HUD limits
ami_cutoffs("CA", "San Francisco", 2024,
            ami_source  = "hud",
            ami_tiers   = c(ELI = 0.30, VLI = 0.50),
            hud_hh_size = 2)
```

### Output

| Column | Meaning |
|---|---|
| `GEOID` | 5-digit county FIPS |
| `NAME` | county name (from ACS, or HUD area name under `ami_source = "hud"`) |
| `ami` | the county AMI (median family income basis) |
| `cut_<TIER>` | one income-ceiling column per tier — e.g. `cut_ELI`, `cut_VLI`, … |

Each `cut_<TIER>` is exactly the `income_cutoff` that `afford_index()` uses for that tier in that
county. Inspecting it is the fastest sanity check on a region's results: if the dollar ceilings look
wrong, the supply numbers will too.

---

# From v1 to v2: what `afford_index()` fixes

The table below is the practical crosswalk — each legacy `afford()` limitation and how the new engine
addresses it.

| Concern | Legacy `afford()` (v1) | `afford_index()` (v2) |
|---|---|---|
| **AMI basis** | median of *tract median household* incomes (`B19013`), collapsed to one regional number | county median **family** income (`B19113`), HUD's anchor — one cutoff **per county** |
| **Income brackets** | positional `rep()` of hard-coded bracket vectors | brackets parsed from ACS variable **labels** (robust to table changes) |
| **Within a bracket** | `closest()` snaps the cutoff to the nearest bracket edge | **linear interpolation** within the bracket (no snapping) |
| **Income tiers** | one `ami_limit` fraction per call | standardized **ELI/VLI/LI/MI** (configurable) in one call |
| **Household size** | none | HUD **family-size factors** via `hud_hh_size` |
| **Owner cost** | single fixed `price * 0.188` factor (≈3.8% mortgage baked in) | two explicit bases: `own_current` (actual `B25094` cost) and `own_buyin` (full mortgage model with your rate/term/down/tax) |
| **Owner universe** | mixes owner value `B25075` + price-asked `B25085` | clean single universe per basis (`B25094` or `B25075`) |
| **Official limits** | ACS only | `ami_source = "hud"` for HUD's published limits |
| **Output shape** | wide, one row per tract with `tr_rent_*`/`tr_own_*` | tidy **long**, one row per tract × tenure × tier |
| **Inputs** | FIPS codes only (names silently break joins) | FIPS **or** names, normalized up front |

The headline fix is the owner side: v1's `0.188` factor corresponds to a ~3.8%, 30-year, 0%-down
loan and at 2024 rates (~7%) **overstates** ownership affordability by roughly 30–40%. v2 replaces it
with a transparent mortgage model whose every assumption is an argument you can set.

---

# Appendix: the legacy `afford()` engine

`afford()` is the original v1 implementation, retained for backward compatibility. Prefer
`afford_index()` for new work. Read the limitations before quoting v1 ownership numbers in a report;
the full review lives in `dev/affordability-index-review.md`.

## Quick start

```{r}
library(neighborhood)

# 5-county Bay Area, households at 50% of AMI, 2024 ACS
bay <- afford(
  state     = "06",
  counties  = c("001", "013", "041", "075", "081"),
  ami_limit = 0.5,
  year      = 2024
)

# With tract geometry for mapping (returns an sf object)
wasatch <- afford(
  state     = "49",
  counties  = c("011", "035", "049", "057"),
  ami_limit = 0.5,
  year      = 2024,
  geometry  = TRUE
)
```

`ami_limit` is the income level of interest as a fraction of AMI: `0.3` ≈ extremely low income,
`0.5` ≈ very low income, `0.8` ≈ low income. Units affordable at or below it are counted
cumulatively. **Use FIPS codes** for `state` and `counties` — names silently break the internal
join.

### Arguments

| Argument | Default | What it does |
|---|---|---|
| `state` | — (required) | 2-digit state FIPS (e.g. `"06"`). Names not supported. |
| `counties` | — (required) | 3-digit county FIPS codes. |
| `ami_limit` | — (required) | Income level as a fraction of AMI. |
| `year` | `2024` | ACS 5-year endpoint year. |
| `geometry` | `FALSE` | If `TRUE`, return an `sf` object with tract polygons. |
| `...` | — | Reserved for future use. |

## How it is calculated

1. **Income threshold.** `AMI` = the median of tract median *household* incomes (`B19013_001`);
   `threshold = ami_limit * AMI`.
2. **Cost → required income.** Rentals (`B25063`): `income = (monthly_rent / 0.30) * 12`. Ownership
   (`B25075` value + `B25085` price): `income = price * 0.188`.
3. **Accessible units.** A unit is accessible if its required income is at or below the threshold;
   `supply = accessible / total` per tract and tenure.
4. **Compare to need.** `ratio = supply / class_prop` (location quotient) and
   `rate = accessible / class_pop * 100000`.

## Output columns

`afford()` returns one **wide** row per tract:

| Column | Meaning |
|---|---|
| `tr_rent_total`, `tr_own_total` | rental / owned units in the tract |
| `tr_rent_accessible`, `tr_own_accessible` | units affordable to the income group |
| `tr_rent_supply`, `tr_own_supply` | accessible ÷ total |
| `tr_rent_ratio`, `tr_own_ratio` | supply ÷ regional group share (location quotient) |
| `tr_rent_rate`, `tr_own_rate` | accessible per 100k regional group-households |
| `reg_total_pop`, `reg_class_pop` | region households (all / in the income group) |
| `ami_limit` | the income level used |
| `rent_jenks_cat`, `own_jenks_cat` | 3-class map categories |
| `popup` | preformatted HTML for `leaflet` maps |

## Known limitations

1. **Ownership embeds a fixed interest rate.** The `0.188` factor ≈ a 3.8%, 30-year, 0%-down loan;
   at 2024 rates it overstates ownership affordability ~30–40%. (Fixed in v2 via the explicit
   `own_buyin` model.)
2. **"AMI" is the median of tract medians**, not HUD AMI, and collapses multi-county regions into one
   cutoff. (Fixed in v2: county median family income.)
3. **Occupied stock, not availability.** Scores existing units, not vacant/available ones (Gate 2 in
   the design).
4. **Price/value mixes two census universes** (`B25075` + `B25085`).
5. **No non-price barriers** — discrimination, voucher rejection, eviction-record screening
   ("the scarlet E") are absent (Gate 3 in the design).
6. **Margins of error are dropped.**
7. **Inputs must be FIPS codes** — passing names returns an empty join and distorts regional totals.

---

## Data sources

All inputs are ACS 5-year estimates via `tidycensus::get_acs()` (`CENSUS_API_KEY` required). v2 tables:
`B19113` (median family income), `B19001` (income distribution), `B25063` (gross rent), `B25094`
(owner cost, basis A), `B25075` (home value, basis B). v1 additionally uses `B19013` and `B25085`.
Geometry comes from `tigris::tracts()`. `ami_source = "hud"` uses the `hudr` package + `HUD_API_KEY`.

## See also

- `dev/affordability-index-design.md` — the three-gate "choice" funnel and the full v2 design.
- `dev/affordability-index-review.md` — the v1 methodological review and remediation plan.
- `dev/hud-income-limits-architecture.md` — HUD income-limit reproduction and longevity.
- `vignette("mapping-with-maplibre")` — mapping `afford_index()` output.
- `ntdf()` — neighborhood racial typologies, joinable on `GEOID` for equity analysis.

## Acknowledgments

The affordability index is part of the Eviction Research Network and CiDR Lab housing-precarity
stack. Thanks to the ERN team and collaborators who pressure-tested the methodology.

## Issues and feedback

Hit a bug, or have a feature request or feedback? Please
[open a GitHub issue](https://github.com/evictionresearch/neighborhood/issues) —
problem reports and suggestions are always welcome.
</content>
</invoke>
