---
title: "Color palettes: qualitative, sequential, diverging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Color palettes: qualitative, sequential, diverging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, message = FALSE)
```

```{r setup}
library(neighborhood)
library(ggplot2)
library(dplyr)
```

The package ships one color system, organized by **what the data means** rather
than by what looks nice: qualitative palettes for mutually exclusive categories,
sequential ramps for a single magnitude, diverging ramps for two directions
around a neutral center. All of it leads with the 2022 brand triad — dark navy
`#19222C`, blue navy `#223754`, brand red `#F9322B` — and every palette carries
WCAG 2.1 receipts (the accessibility contrast rules), enforced by the package's
test suite.

## The swatch sheet

Start here. One call opens a titled sheet of every palette, generated from the
live definitions so it always matches your installed version — each card
labeled with the exact string to pass to `ern_palette()` or the scales, a
one-line "use for" rule, and a computed contrast badge per swatch:

```{r, eval = FALSE}
ern_swatch()
```

Pick the card that matches the data's meaning, then use its name below.

## Choosing a palette

| Your data | Palette | ggplot2 scale |
|---|---|---|
| Categories (series, groups) | `"qualitative"` | `scale_color_ern()` / `scale_fill_ern()` |
| Categories where red would read as "bad" | `"neutral"` | `scale_fill_ern("neutral")` |
| The 19 neighborhood typologies | `"typology"` | `scale_fill_ern("typology")` |
| One magnitude, no valence | `"blues"` | `scale_fill_ern_c("blues")` |
| One magnitude, red-themed (filings, displacement) | `"reds"` | `scale_fill_ern_c("reds")` |
| One magnitude, more = good | `"greens"` | `scale_fill_ern_c("greens")` |
| Risk (the HPRM / state-profile convention) | `"ramp"` | `scale_fill_ern_c()` |
| Two directions, one side bad | `"div_brand"` | `scale_fill_ern_div("brand")` |
| Two directions, neither side bad | `"div_gold"` | `scale_fill_ern_div("gold")` |
| Raw hex vectors for anything else | any of the above | `ern_palette("...")` |

## The red rule

Brand red is central — and it is used by role, not banned. Verified against
WCAG 2.1 (ratios on white):

| Red | Ratio | Role |
|---|---|---|
| `#F9322B` brand red | 3.8:1 | **Shapes**: fills, lines, markers, map washes, large text. Passes the 3:1 non-text minimum (WCAG 1.4.11); fails body text. |
| `#CC2118` text red | 5.5:1 | **Words**: direct labels, annotations, links. AA-safe, on white and on the `#e8eef4` tint. |
| `#B01D16` deep red | 6.9:1 | Hover, active, deep emphasis. |

One trap: red text on navy fails (2.9:1) — use white or steel there. All three
live in `ern_palette("brand")` as `accent`, `accent_deep`, `accent_deeper`.

## Qualitative: categories

The default palette opens with the brand triad — blue navy carries the
editorial line, brand red the highlight — and its first four colors are the
most colorblind-separable subset, so the common chart sizes stay legible under
color-vision deficiency. (Gold arrives fifth: it blurs with red for
deuteranopes, so past four series lean on the house style's direct labels.)

```{r qualitative-lines}
big4 <- mn_evictions |>
  filter(county %in% c("Hennepin", "Ramsey", "Dakota", "Anoka")) |>
  summarise(filings = sum(filings), .by = c(county, year))

ggplot(big4, aes(year, filings, color = county)) +
  geom_line(linewidth = 0.8) +
  scale_color_ern() +
  labs(title = "Eviction filings, four largest counties",
       subtitle = "Minnesota, 2017-2024", x = NULL, y = "Filings") +
  theme_ern()
```

When the categories are people or programs, red stops being a highlight and
starts being a verdict — a red band on a race composition reads as "the bad
group." That is what `"neutral"` is for: the same family with no alarm color.

```{r neutral-composition, fig.height = 3.2}
renters <- mn_evictions |>
  filter(year == 2024, county %in% c("Hennepin", "Ramsey", "Olmsted")) |>
  distinct(county, Black = renters_black, White = renters_white,
           Latine = renters_latine, Other = renters_other) |>
  tidyr::pivot_longer(-county, names_to = "race", values_to = "renters")

ggplot(renters, aes(renters, county, fill = race)) +
  geom_col(position = "fill") +
  scale_fill_ern("neutral") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Renter households by race", x = NULL, y = NULL) +
  theme_ern()
```

Two more qualitative options: `"typology"` maps the 19 neighborhood-typology
classes to their canonical key (`vignette("precarity-mapping")` shows it on a
map — `nt_add_choropleth()` applies it automatically), and `"legacy"` is the
pre-1.1.0 palette, kept verbatim so already-published figures can be reproduced
exactly. Don't use `"legacy"` for new work.

## Sequential: one magnitude

Light-to-dark, interpolable to any number of classes with `n`. Filings are a
quantity the network tells in red, so the heatmap below uses `"reds"`; use
`"blues"` when the magnitude has no valence (counts, density, shares) and
`"greens"` when the high end is desirable (affordability, stability). The
4-class `"ramp"` (yellow-orange-red-purple) is the published HPRM / Minnesota /
Washington **risk** convention — reserve it for genuinely bad quantities, and
don't repurpose it for neutral ones.

```{r sequential-heatmap, fig.height = 3.6}
top8 <- mn_evictions |>
  filter(year == 2024) |>
  summarise(f = sum(filings), .by = county) |>
  slice_max(f, n = 8) |>
  pull(county)

monthly <- mn_evictions |>
  filter(year == 2024, county %in% top8) |>
  summarise(filings = sum(filings), .by = c(county, month))

ggplot(monthly, aes(factor(month), reorder(county, filings), fill = filings)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_ern_c("reds") +
  labs(title = "Monthly filings, top-8 counties, 2024",
       x = "Month", y = NULL, fill = "Filings") +
  theme_ern(grid = "none")
```

```{r, eval = FALSE}
# the same scale family, binned for a choropleth legend:
ern_palette("blues", n = 5)   # 5 classes, light -> dark navy
ern_palette("greens", n = 7)
```

## Diverging: two directions

Set `midpoint` to pin the neutral center to a meaningful value (usually `0`),
rather than letting it float at the middle of the data range. When one
direction is genuinely bad — filings rising — use `"brand"`, which puts brand
red on the bad pole: that is exactly what red is for.

```{r diverging-change, fig.height = 4.6}
change <- mn_evictions |>
  filter(year %in% c(2023, 2024)) |>
  summarise(filings = sum(filings), .by = c(county, year)) |>
  tidyr::pivot_wider(names_from = year, values_from = filings,
                     names_prefix = "y") |>
  mutate(chg = y2024 - y2023) |>
  slice_max(abs(chg), n = 12)

ggplot(change, aes(chg, reorder(county, chg), fill = chg)) +
  geom_col() +
  scale_fill_ern_div("brand", midpoint = 0) +
  labs(title = "Change in filings, 2023 to 2024",
       subtitle = "Twelve largest swings; red = filings rose",
       x = "Change in filings", y = NULL) +
  theme_ern(legend = "none")
```

When neither direction is a warning — above vs. below average, owner vs.
renter share — use `"gold"`: blue against gold, the most colorblind-safe
diverging pair, with no alarm on either pole:

```{r, eval = FALSE}
scale_fill_ern_div("gold", midpoint = 0)
```

Every fill scale has a color twin — `scale_color_ern_div()` (and the British
`scale_colour_ern_div()` alias) work identically for point and line aesthetics.

## The interactive charts and maps

Every `nt_*` chart takes a `palette` argument, and `ern_palette()` hands the
same vectors to anything else — so the interactive and static artifacts stay on
one system:

```{r, eval = FALSE}
# interactive bar chart, red-free composition colors
nt_bar(rates, "county", "rate", group = "race",
       palette = ern_palette("neutral"))

# leaflet / MapLibre maps: typology colors are applied automatically by
# nt_pal() and nt_add_choropleth(); numeric layers use the risk ramp
ern_palette("ramp")
```

## What the palettes guarantee

Enforced by `tests/testthat/test-ern-palettes.R`, so they hold in every build:

* every `"qualitative"` and `"neutral"` fill clears **3:1 on white** (WCAG
  1.4.11, the non-text minimum);
* the text reds clear **4.5:1 (AA)** on white and on the light-blue tint;
* sequential ramps darken **monotonically**; diverging ramps peak at the
  center — no false steps in a binned legend;
* brand red sits in slot 2 of the default palette, and `"neutral"` contains no
  red at all;
* `"legacy"` never changes.
