---
title: "Geography utilities: crosswalks and cleaner tract maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geography utilities: crosswalks and cleaner tract maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Census tracts are the package's analysis unit, but the questions rarely stop at
tract lines: cities want *their* numbers, rent indexes arrive by ZIP, PUMS
microdata arrives by PUMA, and coastal tracts drag their maps into the water.
Four small utilities handle the plumbing. All of them exist so that estimates
can move **between geographies without areal hand-waving** — wherever possible
the weights are built on census *blocks*, which nest exactly inside the larger
units.

## `nt_areal_weight()` — tracts ↔ places, block-exact

*"What are Redwood City's numbers?"* is not "the tracts inside Redwood City" —
tracts straddle municipal limits. `nt_areal_weight()` computes, for every
(place, tract) overlap, the share of the tract's population / housing units /
land that falls inside the place, using decennial blocks as the weighting
surface. Because 2020 blocks nest within both 2020 tracts and 2020 places, the
fractions are exact, not approximations.

```{r}
library(neighborhood)

w <- nt_areal_weight("CA", "San Mateo",
                     places = c("Redwood City", "San Mateo"))

# population-weight any tract-level estimate onto the city:
library(dplyr)
w |>
  left_join(tract_estimates, by = c("tract_geoid" = "GEOID")) |>
  group_by(place_name) |>
  summarise(renters_in_city = sum(renters_e * frac_pop, na.rm = TRUE))
```

By default only *incorporated* places (LSAD 25) are kept, which drops
census-designated places for free. Multiply a tract estimate by `frac_pop`
(or `frac_hu`, `frac_area`) and aggregate — that is the whole workflow.

## `nt_zcta_weights()` — ZIPs → tracts, the same way

The same block-exact idea, pointed at ZIP Code Tabulation Areas. This is the
crosswalk under the affordability product's entry-price layer
(`vignette("three-panes")`): Zillow's asking-rent index is ZIP-level, tracts
are the analysis unit, and 2020 blocks nest inside both.

```{r}
w <- nt_zcta_weights("53", "033")     # King County
head(w[, c("zcta", "tract_geoid", "frac_hu")])
```

`frac_hu` sums to ~1 over each tract, so a tract-level version of any ZIP-level
quantity is a housing-unit-weighted average of its ZIPs. The first call
downloads the national ZCTA boundary file through `tigris` (it is not published
by state) — cache it with `options(tigris_use_cache = TRUE)`; every later call
is instant.

## `nt_erase_water()` — tracts that stop at the shoreline

Coastal and lakefront tracts legally extend into open water, which makes
choropleths read as if the bay had a rent burden. `nt_erase_water()` clips the
water out of an `sf` layer for *display* (the data are untouched):

```{r}
balt <- ntdf(state = "MD", county = "Baltimore City", geometry = TRUE) |>
  nt_erase_water()
nt_map(balt)
```

Use it on the mapping copy only — never re-weight estimates by the clipped
areas.

## `get_co_puma()` — counties ↔ PUMAs

PUMS microdata lives on Public Use Microdata Areas. Urban counties contain
several PUMAs; rural PUMAs contain several counties. `get_co_puma()` returns
the county↔PUMA correspondence so microdata summaries can be carried to
counties (and, since tracts nest in PUMAs, aggregated tract work can meet
microdata halfway).

```{r}
cw <- get_co_puma(state = "CA")
```

## See also

- `vignette("three-panes")` — where `nt_zcta_weights()` earns its keep (the
  entry-price layer).
- `vignette("mapping-with-maplibre")` — the mapping toolkit these feed.
