---
title: "Geography utilities: crosswalks and cleaner tract maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geography utilities: crosswalks and cleaner tract maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Census tracts are the package's analysis unit, but the questions rarely stop at
tract lines: cities want *their* numbers, rent indexes arrive by ZIP, PUMS
microdata arrives by PUMA, and coastal tracts drag their maps into the water.
Four small utilities handle the plumbing. All of them exist so that estimates
can move **between geographies without areal hand-waving** — wherever possible
the weights are built on census *blocks*, which nest exactly inside the larger
units.

## `nt_areal_weight()` — tracts ↔ places, block-exact

*"What are Redwood City's numbers?"* is not "the tracts inside Redwood City" —
tracts straddle municipal limits. `nt_areal_weight()` computes, for every
(place, tract) overlap, the share of the tract's population / housing units /
land that falls inside the place, using decennial blocks as the weighting
surface. Because 2020 blocks nest within both 2020 tracts and 2020 places, the
fractions are exact, not approximations.

A worked example — two neighboring cities in San Mateo County (weights computed
2026-07-05 from the 2020 decennial blocks; the chunk is shown, not re-run at
build):

```{r}
library(neighborhood)

w <- nt_areal_weight("CA", "San Mateo",
                     places = c("Redwood City", "San Mateo"))
nrow(w)
#> [1] 52
```

Fifty-two (place, tract) rows cover the two cities. Forty of them are the easy
case — the tract lies fully inside one city (`frac_pop = 1`). The other rows
are why this function exists. Sorted by the population fraction:

```{r}
w[w$frac_pop > 0 & w$frac_pop < 1,
  c("place_name", "tract_geoid", "pop_in", "pop_t", "frac_pop", "frac_hu")]
#>    place_name tract_geoid pop_in pop_t frac_pop frac_hu
#>  Redwood City 06081610700   5108  5200    0.982   0.988
#>  Redwood City 06081611100   5898  6423    0.918   0.924
#>  Redwood City 06081610302   4338  4796    0.905   0.904
#>  Redwood City 06081609800   3763  4733    0.795   0.792
#>     San Mateo 06081606800   1651  3344    0.494   0.516
#>  ...
```

Read the last row: tract `06081606800` straddles the San Mateo city limit
almost exactly in half — 1,651 of its 3,344 residents live inside the city.
Any "the tracts in San Mateo" shortcut either counts all 3,344 or none of
them; the block-built weight counts 49.4% of whatever you are estimating.
(A zero-population water tract in the bay also touches both cities — its
fractions are all 0, so it drops out of any weighted sum on its own.)

Applying the weights is one join and one multiply:

```{r}
# population-weight any tract-level estimate onto the city:
library(dplyr)
w |>
  left_join(tract_estimates, by = c("tract_geoid" = "GEOID")) |>
  group_by(place_name) |>
  summarise(renters_in_city = sum(renters_e * frac_pop, na.rm = TRUE))
```

By default only *incorporated* places (LSAD 25) are kept, which drops
census-designated places for free. Multiply a tract estimate by `frac_pop`
(or `frac_hu`, `frac_area`) and aggregate — that is the whole workflow.

## `nt_zcta_weights()` — ZIPs → tracts, the same way

The same block-exact idea, pointed at ZIP Code Tabulation Areas. This is the
crosswalk under the affordability product's entry-price layer
(`vignette("three-panes")`): Zillow's asking-rent index is ZIP-level, tracts
are the analysis unit, and 2020 blocks nest inside both.

```{r}
w <- nt_zcta_weights("53", "033")     # King County
head(w[, c("zcta", "tract_geoid", "frac_hu")])
```

`frac_hu` sums to ~1 over each tract, so a tract-level version of any ZIP-level
quantity is a housing-unit-weighted average of its ZIPs. The first call
downloads the national ZCTA boundary file through `tigris` (it is not published
by state) — cache it with `options(tigris_use_cache = TRUE)`; every later call
is instant.

## `nt_erase_water()` — tracts that stop at the shoreline

Coastal and lakefront tracts legally extend into open water, which makes
choropleths read as if the bay had a rent burden. `nt_erase_water()` clips the
water out of an `sf` layer for *display* (the data are untouched):

```{r}
balt <- ntdf(state = "MD", county = "Baltimore City", geometry = TRUE) |>
  nt_erase_water()
nt_map(balt)
```

Use it on the mapping copy only — never re-weight estimates by the clipped
areas.

## `get_co_puma()` — counties ↔ PUMAs

PUMS microdata lives on Public Use Microdata Areas. Urban counties contain
several PUMAs; rural PUMAs contain several counties. `get_co_puma()` returns
the county↔PUMA correspondence so microdata summaries can be carried to
counties (and, since tracts nest in PUMAs, aggregated tract work can meet
microdata halfway).

```{r}
cw <- get_co_puma(state = "CA")
```

## See also

- `vignette("three-panes")` — where `nt_zcta_weights()` earns its keep (the
  entry-price layer).
- `vignette("mapping-with-maplibre")` — the mapping toolkit these feed.
