---
title: "Interactive charts: replicating the Minnesota profile"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive charts: replicating the Minnesota profile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
fig <- function(f) {
  p <- file.path("figures", f)
  if (file.exists(p)) knitr::include_graphics(p)
}
```

This tutorial rebuilds the charts from the
[Eviction Research Network Minnesota profile](https://evictionresearch.net/minnesota/)
using `nt_chart()` and the bundled `mn_evictions` data — a monthly eviction trend
with a pre-pandemic baseline and a moratorium band, yearly totals, and rolling
rates by race. Along the way it shows the **"newspaper graphic" hover behavior**
(the y-axis appears only when you hover) and finishes with a real-world example:
pulling Apartment List rent data from a URL to chart regional rents. The second
half is an **"every lever" reference** — one small example per argument — so the
whole surface of `nt_chart()` and its axis-free sibling `nt_spark()` is covered.

The charts are a thin, deterministic wrapper over
[echarts4r](https://echarts4r.john-coene.com/) (John Coene) / Apache ECharts;
`nt_chart()` returns the raw `echarts4r` widget, so you can keep piping `e_*`
functions for anything it doesn't expose.

```{r setup}
library(neighborhood)
library(dplyr)
```

## The two functions at a glance

`nt_chart()` is the full editorial chart; `nt_spark()` is its axis-free sibling
for inline "trend at a glance" graphics. Their complete signatures:

```{r}
nt_chart(
  data, x, y, group = NULL,
  type = c("line", "area", "bar"),
  stack = NULL,
  baseline = NULL, baseline_label = NULL,
  band = NULL, band_label = NULL,
  palette = NULL,
  value_fmt = c("comma", "percent", "currency", "multiple", "none"),
  smooth = NULL,
  crosshair = TRUE,
  yaxis = c("always", "hover", "none"),
  highlight_last = FALSE,
  end_label = FALSE,
  line_styles = NULL,
  source = NULL,
  legend = !is.null(group),
  title = NULL, subtext = NULL,
  x_title = NULL, y_title = NULL,
  height = NULL, ...
)

nt_spark(values, type = c("line", "bar"),
         color = "#19222C", height = "40px", width = "120px")
```

| Argument | What it does |
|---|---|
| `data`, `x`, `y` | data frame and the x / y column **names** (strings) |
| `group` | column name for one series per group (race, county, …) |
| `type` | `"line"`, `"area"`, or `"bar"` |
| `stack` | group key string to stack areas/bars into a composition |
| `baseline`, `baseline_label` | dashed horizontal reference line + its label |
| `band`, `band_label` | shaded vertical region `c(start, end)` + its label |
| `palette` | series colors; **named** vector matches colors to groups |
| `value_fmt` | `"comma"` / `"percent"` / `"currency"` / `"multiple"` / `"none"` |
| `smooth` | curve lines/areas (default `TRUE` for line/area, `FALSE` for bar) |
| `crosshair` | the hover crosshair + on-axis value (default `TRUE`) |
| `yaxis` | `"always"` / `"hover"` / `"none"` — when the y-scale shows |
| `highlight_last` | accent the latest point (leading series, if grouped) |
| `end_label` | direct "name · value" label at each line end (replaces legend) |
| `line_styles` | per-series dash patterns for print/grayscale |
| `source` | small attribution credit, lower-right |
| `legend` | show a legend (default `TRUE` when `group` set, `FALSE` when `end_label`) |
| `title`, `subtext` | chart title and subtitle |
| `x_title`, `y_title` | axis titles |
| `height` | CSS height, e.g. `"380px"` |
| `...` | passed to `echarts4r::e_line()` / `e_area()` / `e_bar()` |

## The data: `mn_evictions`

`mn_evictions` is the public county-month eviction data behind the Minnesota
profile — 87 counties, 2017-01 through 2026-03, with race-imputed counts:

```{r, eval=TRUE}
dplyr::glimpse(neighborhood::mn_evictions)
```

A note on the rate: `rate` is the raw monthly proportion `filings / renters`.
The profiles headline an **annualized rate per 1,000 renter households**, which
you compute over a window as `1000 * sum(filings) * 12 / n_months / renters`.

# Part I — Replicating the profile plates

## 1. The monthly trend (Figure 1)

Roll the counties up to a statewide monthly series, then chart it. The hallmark
ERN touches — a dashed pre-pandemic baseline, a shaded eviction-moratorium band,
and an accented latest point — are each one argument. Set `yaxis = "hover"` so
the y-axis is hidden until the reader hovers, exactly like the live page:

```{r}
statewide <- mn_evictions |>
  group_by(year, month) |>
  summarise(filings = sum(filings), .groups = "drop") |>
  mutate(date = as.Date(sprintf("%d-%02d-01", year, month))) |>
  filter(date >= as.Date("2019-01-01"))

prepan <- mn_evictions |>            # 2017-2019 monthly average = the baseline
  filter(year <= 2019) |>
  group_by(year, month) |>
  summarise(filings = sum(filings), .groups = "drop") |>
  pull(filings) |>
  mean()

nt_chart(
  statewide, "date", "filings", type = "line",
  yaxis          = "hover",                                   # y-axis appears on hover
  baseline       = prepan, baseline_label = "2017-2019 avg",
  band           = c(as.Date("2020-03-01"), as.Date("2021-06-01")),
  band_label     = "Moratorium",
  highlight_last = TRUE
)
```

```{r mn-trend, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Statewide monthly eviction filings (shown mid-hover): the y-axis and gridlines appear only on hover, with the dashed pre-pandemic baseline, the shaded moratorium band, and the accented latest month."}
fig("chart-mn-trend.png")
```

`yaxis = "hover"` is the option to make the y-axis **disappear while you are not
hovering** and reappear (with dashed gridlines) when you are — the minimalist
newspaper look. The default is `"always"` (labels always shown); `"none"` hides
them entirely and relies on the hover crosshair value.

## 2. Yearly totals (Figure 2)

Bars start at zero (the honest baseline for a magnitude), so `type = "bar"`
ignores `smooth` and draws from the axis up:

```{r}
yearly <- mn_evictions |>
  group_by(year) |>
  summarise(filings = sum(filings), .groups = "drop")

nt_chart(yearly, "year", "filings", type = "bar",
         yaxis = "hover", highlight_last = TRUE)   # 2026 is year-to-date
```

```{r mn-bars, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Yearly eviction-filing totals; the latest (partial) year is accented and the y-axis reveals on hover."}
fig("chart-mn-bars.png")
```

A discrete numeric x like `year` is automatically coerced to an ordered category
axis (otherwise ECharts would put it on a 0-based value axis and bunch the bars
up), so the years stay in order and evenly spaced.

## 3. Rates by race — the published plate (Plate V)

The race chart is a rolling **annualized rate per 1,000 renter households** for
each group. Build the 12-month rolling rate from the imputed counts and the
matching ACS denominators, reshape to long, and pass `group`:

```{r}
roll12 <- function(x) as.numeric(stats::filter(x, rep(1, 12), sides = 1))  # trailing 12-mo sum

race <- mn_evictions |>
  arrange(year, month) |>
  group_by(year, month) |>
  summarise(across(c(filings_black, filings_white, filings_latine, filings_other,
                     renters_black, renters_white, renters_latine, renters_other),
                   sum),          # statewide totals (sum across all 87 counties)
            .groups = "drop") |>
  mutate(date = as.Date(sprintf("%d-%02d-01", year, month)),
         Black  = 1000 * roll12(filings_black)  / renters_black,
         White  = 1000 * roll12(filings_white)  / renters_white,
         Latine = 1000 * roll12(filings_latine) / renters_latine,
         Other  = 1000 * roll12(filings_other)  / renters_other) |>
  tidyr::pivot_longer(c(Black, White, Latine, Other),
                      names_to = "race", values_to = "rate") |>
  filter(!is.na(rate))
```

To match the published plate, lean on the editorial arguments: `end_label` writes
each group's name and latest value at the line end (so no legend is needed),
`line_styles` gives each series a print-legible dash pattern, a mostly-ink
`palette` keeps the comparison group (White) light, `highlight_last` accents the
leading line, and `source` prints the data credit. `yaxis = "hover"` keeps the
scale hidden until the reader hovers:

```{r}
nt_chart(
  race, "date", "rate", group = "race", type = "line",
  yaxis          = "hover",
  end_label      = TRUE,                       # "Black · 33.5" at each line end
  highlight_last = TRUE,                       # red dot on the leading series
  line_styles    = c(Black = "solid", Latine = "dotted",
                     Other = "dashed", White = "solid"),
  palette        = c(Black = "#19222C", Latine = "#19222C",
                     Other = "#3f4b57", White = "#9fb6c4"),
  source         = "lsc · acs · wru v2.0 · birdie"
)
```

```{r mn-race, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Filing rate by race, rolling annualized per 1,000 renter households (shown mid-hover). Direct end-labels replace the legend, each series has its own line style, and the y-axis appears only on hover."}
fig("chart-mn-race.png")
```

## 4. The disparity ratio (Plate VI)

The companion plate expresses each group's rate as a **ratio to the white rate** —
above `1.0` means filed against at a higher rate than white renters. Reshape to
wide, divide by `White`, and chart it with a `1.0` parity baseline and
`value_fmt = "multiple"` so values read as `1.52×`:

```{r}
ratio <- race |>
  select(date, race, rate) |>
  tidyr::pivot_wider(names_from = race, values_from = rate) |>
  mutate(Black = Black / White, Latine = Latine / White, Other = Other / White) |>
  tidyr::pivot_longer(c(Black, Latine, Other), names_to = "race", values_to = "ratio") |>
  filter(!is.na(ratio))

nt_chart(
  ratio, "date", "ratio", group = "race", type = "line",
  yaxis          = "hover",
  value_fmt      = "multiple",                 # 1.52x
  end_label      = TRUE, highlight_last = TRUE,
  baseline       = 1, baseline_label = "parity with white renters",
  line_styles    = c(Black = "solid", Latine = "dotted", Other = "dashed"),
  palette        = c(Black = "#19222C", Latine = "#19222C", Other = "#9fb6c4"),
  source         = "lsc · acs · wru v2.0 · birdie"
)
```

```{r mn-ratio, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Black, Latine, and Other filing rates as a ratio of the white filing rate, with a dashed parity line at 1.0 and end-labels in multiples (shown mid-hover)."}
fig("chart-mn-ratio.png")
```

# Part II — Every lever, one at a time

This section walks each argument with a minimal, correct example. Run any of
them to see the live behavior; the prose is the documentation.

## Chart types: `line`, `area`, `bar`

`type` is the basic choice. Lines and areas curve by default; bars start at zero.

```{r}
nt_chart(statewide, "date", "filings", type = "line")   # default
nt_chart(statewide, "date", "filings", type = "area")   # line + soft fill below
nt_chart(yearly,    "year", "filings", type = "bar")    # zero-based magnitude bars
```

## `smooth` — curved vs. straight

`smooth` defaults to `TRUE` for `line`/`area` and `FALSE` for `bar`. Set
`smooth = FALSE` for a literal polyline (each month-to-month step is a straight
segment), which can be the more honest read for a volatile series:

```{r}
nt_chart(statewide, "date", "filings", type = "line", smooth = FALSE)
```

## `stack` — areas and bars as a composition

Pass `stack` (any shared string key) to stack the series so they sum into a total
band — the right read when the parts make up a whole. Here the four race counts
stack into total monthly filings:

```{r}
race_counts <- mn_evictions |>
  group_by(year, month) |>
  summarise(across(c(filings_black, filings_white, filings_latine, filings_other), sum),
            .groups = "drop") |>
  mutate(date = as.Date(sprintf("%d-%02d-01", year, month))) |>
  tidyr::pivot_longer(starts_with("filings_"),
                      names_to = "race", values_to = "filings") |>
  filter(date >= as.Date("2022-01-01"))

# Stacked AREA: composition over time
nt_chart(race_counts, "date", "filings", group = "race", type = "area",
         stack = "total")

# Stacked BAR works the same way
nt_chart(race_counts, "date", "filings", group = "race", type = "bar",
         stack = "total")
```

## `value_fmt` — all five formats

`value_fmt` formats the y-axis ticks, the hover crosshair box, and the tooltip,
keeping them all in agreement. The five styles:

| `value_fmt` | Input → rendered | Note |
|---|---|---|
| `"comma"` (default) | `12345` → `12,345` | thousands separator |
| `"percent"` | `0.25` → `25%` | **expects a proportion in `[0, 1]`** |
| `"currency"` | `1800` → `$1,800`; `-100` → `-$100` | signs negatives as `-$100` |
| `"multiple"` | `1.52` → `1.52×` | ratio / parity charts |
| `"none"` | `1800` → `1800` | raw number, no formatting |

```{r}
# comma (default) — already used above; explicit for completeness
nt_chart(statewide, "date", "filings", value_fmt = "comma")

# percent — note the input must be a proportion in [0, 1]
share <- statewide |>
  mutate(share = filings / max(filings))          # 0..1
nt_chart(share, "date", "share", value_fmt = "percent")

# currency — negatives render as -$100 (e.g. a year-over-year change)
nt_chart(statewide, "date", "filings", value_fmt = "currency")

# multiple — 1.52x (used in the disparity ratio plate above)
nt_chart(ratio, "date", "ratio", group = "race", value_fmt = "multiple")

# none — show the raw number with no separators or symbols
nt_chart(yearly, "year", "filings", type = "bar", value_fmt = "none")
```

## `yaxis` and `crosshair` — the no-axis read

`yaxis` controls when the y-scale appears, and `crosshair` supplies the value on
hover. Together they make the "newspaper graphic" look: with `yaxis = "none"`
there is **no permanent y-axis at all** — the only way to read a value is the
crosshair, which prints the formatted number in a box where the cursor's
horizontal line meets the axis.

```{r}
# always (default): labels always visible
nt_chart(statewide, "date", "filings", yaxis = "always")

# hover: labels + dashed gridlines hidden until you hover (the profile look)
nt_chart(statewide, "date", "filings", yaxis = "hover")

# none: no y-axis ever; the crosshair value is the only readout
nt_chart(statewide, "date", "filings", yaxis = "none")
```

Turning the crosshair off (`crosshair = FALSE`) drops the dashed cross and the
boxed on-axis value, leaving just the tooltip bubble. Pair it with
`yaxis = "none"` only if you don't need readers to read exact values:

```{r}
nt_chart(statewide, "date", "filings", crosshair = FALSE)
```

## A generic multi-series palette

For a non-race comparison (counties, regions, anything), pass `group` and an
unnamed `palette` — colors are applied in series order. The package ships an
8-color default if you pass none:

```{r}
counties <- mn_evictions |>
  filter(county %in% c("Hennepin", "Ramsey", "Dakota", "Anoka")) |>
  group_by(county, year) |>
  summarise(filings = sum(filings), .groups = "drop")

nt_chart(counties, "year", "filings", group = "county", type = "line",
         palette = c("#19222C", "#F9322B", "#223754", "#8BA3BE"))
```

A **named** palette (`c(Hennepin = "#19222C", ...)`) instead pins each color to a
specific group regardless of order — the technique used in the race plates.

## `legend`

A legend appears automatically when `group` is set (and is suppressed when
`end_label = TRUE`, since the direct labels replace it). Override either way:

```{r}
nt_chart(counties, "year", "filings", group = "county", legend = TRUE)   # explicit on
nt_chart(counties, "year", "filings", group = "county", legend = FALSE)  # hide it
```

## Titles: `title`, `subtext`, `x_title`, `y_title`

`title`/`subtext` set the heading and sub-heading; `x_title`/`y_title` label the
axes:

```{r}
nt_chart(statewide, "date", "filings", type = "line",
         title   = "Minnesota eviction filings",
         subtext = "Statewide, monthly",
         x_title = "Month",
         y_title = "Filings")
```

## `height`

`height` takes any CSS length string for the widget's drawn height:

```{r}
nt_chart(statewide, "date", "filings", height = "260px")
```

## Reference marks: `baseline` / `band`

`baseline` draws a dashed horizontal line at a y value (an average, a target, a
`1.0` parity line); `band` shades a vertical region between two x values (a
moratorium window, a recession). Each takes an optional label:

```{r}
nt_chart(
  statewide, "date", "filings", type = "line",
  baseline = prepan, baseline_label = "pre-pandemic avg",
  band = c(as.Date("2020-03-01"), as.Date("2021-06-01")), band_label = "moratorium"
)
```

On a category axis (years, not dates) the `band` edges are matched as category
values, so pass them in the same form as the x values, e.g.
`band = c(2020, 2021)` for a year axis.

## Editorial direct labels: `highlight_last`, `end_label`, `line_styles`

These three replace chart-junk with editorial cues. `highlight_last` puts an
accent dot on the most recent point (the leading series when grouped);
`end_label` writes "name · value" at the end of each line *instead of* a legend;
`line_styles` gives each series a dash pattern so the chart survives grayscale
printing. They are line/area features:

```{r}
nt_chart(
  counties, "year", "filings", group = "county", type = "line",
  highlight_last = TRUE,
  end_label      = TRUE,
  line_styles    = c("solid", "dashed", "dotted", "solid")  # recycled in series order
)
```

`line_styles` can be **named** to pin a pattern to a group
(`c(Hennepin = "solid", Ramsey = "dashed")`); an unnamed vector is recycled in
series order. Valid values are `"solid"`, `"dashed"`, `"dotted"`.

## `source`

`source` prints a small attribution in the lower-right — the credit line on the
profile plates:

```{r}
nt_chart(statewide, "date", "filings", source = "Eviction Research Network")
```

## `...` — drop through to echarts4r, and the return value

Anything in `...` is forwarded to the underlying series function
(`echarts4r::e_line()` / `e_area()` / `e_bar()`). And because `nt_chart()`
returns the **raw echarts4r widget**, you can keep piping any `e_*` function for
features it doesn't expose — a data-zoom slider, a second title, animations:

```{r}
nt_chart(statewide, "date", "filings", type = "line", yaxis = "hover") |>
  echarts4r::e_datazoom(x_index = 0) |>           # draggable zoom slider
  echarts4r::e_title("Minnesota eviction filings")
```

# Part III — Sparklines (`nt_spark`)

`nt_spark()` is the axis-free sibling of `nt_chart()`: a tiny line or bar for a
stat card or a table cell, with no axes, legend, or labels — just the shape. It
takes a bare numeric **vector** (not a data frame) and four styling arguments:

| Argument | Default | What it does |
|---|---|---|
| `values` | — | numeric vector to plot |
| `type` | `"line"` | `"line"` (smoothed, soft fill) or `"bar"` |
| `color` | `"#19222C"` | line/bar color (ERN navy) |
| `height` | `"40px"` | CSS height |
| `width` | `"120px"` | CSS width |

```{r}
# default smoothed line
nt_spark(yearly$filings)

# bar variant
nt_spark(yearly$filings, type = "bar")

# recolor and resize for a wide stat-card slot
nt_spark(yearly$filings, type = "line", color = "#F9322B",
         height = "48px", width = "200px")
```

# Part IV — Pulling data from a URL: Apartment List regional rents

A common workflow is to pull a public dataset straight from the web, reshape it,
and chart it. Here we use [Apartment List's Rent
Estimates](https://www.apartmentlist.com/research/category/data-rent-estimates) —
median new-lease rents by metro/city/county — to chart a metro's rent over time.

The download links are content-hashed and change with each monthly release, so we
scrape the current link off the data page, then read the CSV directly:

```{r}
library(readr); library(tidyr); library(stringr); library(lubridate)

# 1. Resolve the current CSV link off the data page. The hash in the path
#    changes each monthly release, so scrape the link rather than hard-coding it
#    (and keep a pinned fallback in case the page serves bot-filtered content).
page <- "https://www.apartmentlist.com/research/category/data-rent-estimates"
html <- paste(readLines(url(page), warn = FALSE), collapse = "\n")
csv_url <- str_extract(
  html,
  "https://assets\\.ctfassets\\.net/[^\"\\\\]*Apartment_List_Rent_Estimates_2[0-9_]+\\.csv"
)
if (is.na(csv_url)) {
  csv_url <- paste0("https://assets.ctfassets.net/jeox55pd4d8n/2nz8cro1T1bopNPDG6Sbsf/",
                    "54e7e1745db655de15aaa99c5570e6ea/Apartment_List_Rent_Estimates_2026_05.csv")
}

# 2. Read it (keep FIPS as character to preserve leading zeros)
rent <- read_csv(csv_url, col_types = cols(location_fips_code = col_character()))

# 3. Pick a metro (by CBSA code) and bed size, reshape wide -> long
mpls <- rent |>
  filter(location_type == "Metro",
         location_fips_code == "33460",   # Minneapolis-St. Paul-Bloomington, MN-WI
         bed_size == "overall") |>
  pivot_longer(matches("^\\d{4}_\\d{2}$"), names_to = "month", values_to = "rent") |>
  mutate(date = ym(month)) |>
  filter(!is.na(rent))

# 4. Chart it
nt_chart(mpls, "date", "rent", type = "line", value_fmt = "currency",
         yaxis = "hover", title = "Median new-lease rent — Minneapolis metro")
```

```{r apt-rent, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Median new-lease rent for the Minneapolis-St. Paul metro from Apartment List Rent Estimates, pulled live from the URL (shown mid-hover). Data: Apartment List."}
fig("chart-apartmentlist.png")
```

To define your own **region** (a set of cities, or several metros), filter to
those rows and aggregate per month — population-weighting is the defensible
choice since each row is already a median:

```{r}
area <- rent |>
  filter(location_type == "City", bed_size == "overall",
         location_fips_code %in% c("2743000", "2758000")) |>   # Minneapolis, St. Paul
  pivot_longer(matches("^\\d{4}_\\d{2}$"), names_to = "month", values_to = "rent") |>
  mutate(date = ym(month)) |>
  filter(!is.na(rent)) |>
  group_by(date) |>
  summarise(rent = weighted.mean(rent, population), .groups = "drop")

nt_chart(area, "date", "rent", type = "line", value_fmt = "currency", yaxis = "hover")
```

Always **attribute** Apartment List on the chart, read the CSV at runtime rather
than redistributing it, and review their terms of use.

## Saving a chart

A chart is an `htmlwidget`; save a standalone file with
`htmlwidgets::saveWidget()`:

```{r}
chart <- nt_chart(statewide, "date", "filings", type = "line", yaxis = "hover")
htmlwidgets::saveWidget(chart, "mn-trend.html")
```

## Acknowledgments

Charts are built on **echarts4r** (John Coene) and Apache ECharts. Minnesota data
are from the Eviction Research Network; Apartment List rent data are pulled from
[apartmentlist.com](https://www.apartmentlist.com/research/category/data-rent-estimates).
For transparency, the `nt_chart()`/`nt_spark()` functions were developed with the
assistance of **Claude Opus 4.8** (Anthropic); they are deterministic R code and
require no AI to run. See `vignette("mapping-with-maplibre")` for the mapping
toolkit and `vignette("precarity-mapping")` for an advanced interactive
precarity map.

## Issues and feedback

Hit a bug, or have a feature request or feedback? Please
[open a GitHub issue](https://github.com/evictionresearch/neighborhood/issues) —
problem reports and suggestions are always welcome.
