---
title: "Mapping neighborhood data with MapLibre"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping neighborhood data with MapLibre}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
fig <- function(f) {
  p <- file.path("figures", f)
  if (file.exists(p)) knitr::include_graphics(p)
}
```

This tutorial builds an interactive map from nothing to something close to the
[Eviction Research Network](https://evictionresearch.net) state-profile maps
(HPRM, Minnesota, Washington) — and it does so in deterministic R, with no AI in
the loop. We start with a one-line map of neighborhood typologies, then *graft
on* more: a numeric layer, custom breaks and colors, place labels, richer
popups, a layer switcher, and — for the whole United States — vector tiles. The
goal is a complete reference: every argument of every `nt_*` map function, each
with a small, correct example.

The mapping functions are a thin, opinionated layer over
[**mapgl**](https://walker-data.com/mapgl/), Kyle Walker's R package for
MapLibre GL JS. Every `nt_*` function returns an ordinary `mapgl` map, so at any
point you can stop using the `neighborhood` shortcuts and reach for the full
`mapgl` toolbox. Census data throughout comes from Walker's
[**tidycensus**](https://walker-data.com/tidycensus/); his book,
[*Analyzing US Census Data*](https://walker-data.com/census-r/), is the best
companion to this vignette.

A note on what runs here. The map-building chunks need a Census API key and a
web browser, so they are **shown but not executed** when the vignette builds.
The screenshots are pre-rendered. You can copy any chunk into an interactive R
session and it will run.

```{r setup}
library(neighborhood)
library(dplyr)
```

## 1. A map in one line

Map functions need **geometry**. The fastest source is `ntdf()` with
`geometry = TRUE`, which returns an `sf` object of tracts with the
neighborhood-typology columns attached:

```{r}
balt <- ntdf(state = "MD", county = "Baltimore City", geometry = TRUE)

nt_map(balt)
```

```{r typology-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Example output: Baltimore City census tracts colored by neighborhood typology, with the legend, basemap, and address search."}
fig("map-typologies.png")
```

That single call gives you a finished map: an OpenFreeMap *Positron* basemap, the
view framed to Baltimore, every tract colored by its typology (`nt_conc`) using
the canonical palette, a legend, a hover tooltip, a click popup, and the standard
controls (zoom, fullscreen, locate-me, and an address search box).

`nt_map(data, color = NULL, ...)` is the whole signature. When you don't pass
`color`, it looks for an `nt_conc` column and maps it; point it at any other
column to map that instead. Everything in `...` flows straight to
`nt_add_choropleth()`, so every styling argument in Section 4 is available here
too.

```{r}
nt_map(balt, color = "pBlack")                       # share of residents who are Black
nt_map(balt, color = "pBlack", type = "continuous")  # ... as a smooth gradient
```

## 2. The anatomy of a map

`nt_map()` is shorthand. The real workflow is a pipe you compose yourself —
this is what "small or large" means: reach for as many pieces as you need.

```{r}
nt_maplibre(balt) |>                  # 1. basemap + controls + fit to data
  nt_add_choropleth(balt, "nt_conc")  # 2. the colored layer + legend + popup
```

- **`nt_maplibre()`** starts the map: basemap, projection, controls, and (if you
  hand it data) a fit to its bounds (Section 3).
- **`nt_add_choropleth()`** is the workhorse: it colors features by a column and,
  unless told otherwise, adds a matching legend, a hover tooltip, and a click
  popup (Section 4).

Because the return value is a plain `mapgl` map, you can splice in any `mapgl`
call — here, a scale bar and `mapgl`'s own raw layer switcher:

```{r}
nt_maplibre(balt) |>
  nt_add_choropleth(balt, "nt_conc") |>
  mapgl::add_scale_control() |>
  mapgl::add_layers_control()
```

## 3. Starting the map: `nt_maplibre()`

`nt_maplibre()` is the foundation. Called with no arguments it returns an empty
ERN-styled map; hand it spatial `data` and it frames the view to that data's
bounds.

```{r}
nt_maplibre()        # empty map, default view
nt_maplibre(balt)    # framed to Baltimore
```

### Every argument

| Argument | Default | What it does |
|---|---|---|
| `data` | `NULL` | Optional spatial data used only to frame the view (not drawn). `sf`, a shapefile/GeoJSON/GeoPackage path, `sfc`, or `sp`/`terra`; normalized to `sf` and re-projected to WGS84. |
| `style` | `"positron"` | Basemap (see below). |
| `controls` | `TRUE` | Add the standard control set (zoom/compass, fullscreen, geolocate). |
| `search` | `TRUE` | Add an OpenStreetMap/Nominatim address search box (no API key). Ignored when `controls = FALSE`. |
| `fit` | `TRUE` | If `data` is supplied, fit the opening view to its bounds. Ignored when `center` is given. |
| `center` | `NULL` | `c(lng, lat)` to center the map instead of fitting to data. |
| `zoom` | `NULL` | Initial zoom level, paired with `center`. |
| `projection` | `"mercator"` | `"mercator"` (the flat state-profile look) or `"globe"` for the 3-D globe. |
| `...` | — | Passed to `mapgl::maplibre()` (e.g. `height`, `bearing`, `pitch`). |

### Basemaps (`style`)

There are five built-in OpenFreeMap basemaps. Any other value (a style URL, or
the result of `mapgl::carto_style()` / `mapgl::maptiler_style()`) passes straight
through to `mapgl`.

```{r}
nt_maplibre(balt, style = "positron")  # light grey (default)
nt_maplibre(balt, style = "bright")    # full-color OSM
nt_maplibre(balt, style = "liberty")   # OSM "Liberty" theme
nt_maplibre(balt, style = "dark")      # dark mode
nt_maplibre(balt, style = "fiord")     # muted blue-grey

# Anything mapgl accepts also works:
nt_maplibre(balt, style = mapgl::carto_style("positron"))
```

### View: fit, center/zoom, projection

By default a map with `data` fits to that data. To open on a fixed point instead,
give `center` (and usually `zoom`); when `center` is set, `fit` is ignored.

```{r}
# Open centered on downtown Baltimore at zoom 12 instead of fitting to tracts
nt_maplibre(balt, center = c(-76.61, 39.29), zoom = 12)

# Don't auto-fit even though data was supplied
nt_maplibre(balt, fit = FALSE)

# 3-D globe instead of the flat web-mercator map
nt_maplibre(balt, projection = "globe")
```

### Controls and search

The control cluster (zoom/compass, fullscreen, geolocate, address search) is on
by default. Turn the whole cluster off with `controls = FALSE`, or keep the
controls but drop just the search box with `search = FALSE`.

```{r}
nt_maplibre(balt, controls = FALSE)               # bare map, no chrome
nt_maplibre(balt, search = FALSE)                 # controls, but no search box
nt_maplibre(balt, height = "600px", bearing = 20) # `...` -> mapgl::maplibre()
```

## 4. The workhorse: `nt_add_choropleth()`

This is the layer that colors features by a column. It picks sensible defaults
from the data so a one-liner looks right, while leaving every piece overridable.

### Every argument

| Argument | Default | What it does |
|---|---|---|
| `map` | — | A map from `nt_maplibre()` (or `mapgl::maplibre()`). |
| `data` | — | Polygon spatial data (`sf`/path/`sfc`/`sp`/`terra`). |
| `column` | — | Name of the column to color by. |
| `type` | `"auto"` | `"auto"` (categorical for factor/character, binned for numeric), `"categorical"`, `"continuous"` (smooth), or `"step"` (binned). |
| `breaks` | `NULL` | Interior thresholds for `"step"`/`"continuous"`; defaults to data quantiles. |
| `colors` | `NULL` | Fill colors (one per category, or the numeric ramp). |
| `labels` | `NULL` | Legend labels for binned numeric maps. |
| `fill_color` | `NULL` | A raw MapLibre paint expression used verbatim — the full-control escape hatch. |
| `id` | `NULL` | Layer id (defaults to a name derived from `column`). |
| `legend` | `TRUE` | `TRUE`, `FALSE`, or `"interactive"` (clickable legend that filters the map). |
| `legend_title` | `NULL` | Legend title (defaults to `column`). |
| `popup` | `TRUE` | Click popup: `TRUE`, `FALSE`/`NULL`, a column name, or a length-`nrow` vector. |
| `tooltip` | `TRUE` | Hover tooltip: `TRUE` (mapped value), `FALSE`/`NULL`, a column name, or a vector. |
| `opacity` | `0.7` | Fill opacity. |
| `outline_color` | `"#ffffff"` | Polygon outline color. |
| `na_color` | `NULL` | Color for missing values (defaults to ERN grey `#C0C0C0`). |
| `visible` | `TRUE` | Whether the layer starts visible (set `FALSE` for hidden layers in a switcher). |
| `tiles` | `"auto"` | `"auto"`, `"inline"`, or `"pmtiles"` (see Section 8). |
| `max_inline` | `25000` | Feature-count threshold at which `"auto"` switches to PMTiles. |
| `...` | — | Passed to `mapgl::add_fill_layer()`. |

### Categorical maps: typologies

`nt_conc` is a factor with up to 19 levels, and `nt_add_choropleth()` recognizes
it and applies the Eviction Research Network color key automatically (the same
colors as `nt_pal()` for leaflet — see Section 9). You rarely need to touch the
colors, but you can: pass your own `colors`, set `legend = "interactive"` to make
the legend filter the map, change the `legend_title`, or dial down `opacity`.

```{r}
nt_maplibre(balt) |>
  nt_add_choropleth(
    balt, "nt_conc",
    legend       = "interactive",  # click a swatch to isolate that type
    legend_title = "Neighborhood type",
    opacity      = 0.8
  )
```

To force categorical treatment on a column that happens to be numeric (codes, for
example), set `type = "categorical"` and supply your own `colors`:

```{r}
nt_maplibre(balt) |>
  nt_add_choropleth(balt, "some_code",
                    type   = "categorical",
                    colors = c("#1f78b4", "#33a02c", "#e31a1c"))
```

### Numeric maps: binned, continuous, stepped

Any numeric column maps as a **binned choropleth** using the sequential ERN ramp
(yellow → orange → red → purple). `breaks` are the *interior* thresholds; with
the 4-color ramp you supply three of them. Omit `breaks` and the data's quantiles
are used.

```{r}
aff <- afford(state = "24", counties = "510", ami_limit = 0.5, year = 2024,
              geometry = TRUE)

nt_maplibre(aff) |>
  nt_add_choropleth(
    aff, "tr_rent_ratio",
    breaks       = c(0.5, 1, 2),  # 3 interior breaks -> 4 classes
    legend_title = "Affordable-rental ratio"
  )
```

```{r numeric-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Example output: a numeric choropleth binned with the four-class ERN ramp -- Lower, Moderate, High, Extreme."}
fig("map-numeric.png")
```

If you prefer to write *all* bin edges including a floor (the
`c(0, 0.5, 1, 2)` style used in the state configs), that works too — the leading
floor is dropped automatically.

`type` controls the binning style:

```{r}
# Smooth gradient (continuous interpolation) instead of discrete bins
nt_maplibre(aff) |>
  nt_add_choropleth(aff, "tr_rent_ratio", type = "continuous")

# Explicit step (binned) classification, identical to the numeric default
nt_maplibre(aff) |>
  nt_add_choropleth(aff, "tr_rent_ratio", type = "step", breaks = c(0.5, 1, 2))
```

### Custom colors and labels

Override the ramp with `colors`, and rename the bins with `labels`. The default
labels for a 4-class numeric ramp are `c("Lower", "Moderate", "High", "Extreme")`.

```{r}
nt_maplibre(aff) |>
  nt_add_choropleth(
    aff, "tr_rent_ratio",
    breaks = c(0.5, 1, 2),
    colors = c("#fee5d9", "#fcae91", "#fb6a4a", "#cb181d"),
    labels = c("Scarce", "Below fair share", "Fair share", "Abundant"),
    legend_title = "Affordable-rental supply"
  )
```

### Outline, missing values, opacity

```{r}
nt_maplibre(aff) |>
  nt_add_choropleth(
    aff, "tr_rent_ratio",
    opacity       = 0.85,        # more saturated fill
    outline_color = "#444444",   # dark polygon borders instead of white
    na_color      = "#eeeeee"    # how tracts with no data are drawn
  )
```

### Tooltips and popups

The **tooltip** is the hover label; the **popup** is the click panel. Both accept
`TRUE` (auto), `FALSE`/`NULL` (off), a column name, or a length-`nrow` vector.

```{r}
# Default: tooltip shows the mapped value, popup shows id + value
nt_maplibre(balt) |> nt_add_choropleth(balt, "nt_conc")

# Turn the hover tooltip off, keep the click popup
nt_maplibre(balt) |> nt_add_choropleth(balt, "nt_conc", tooltip = FALSE)

# Hover shows a different column than the one being colored
nt_maplibre(balt) |> nt_add_choropleth(balt, "nt_conc", tooltip = "NAME")

# Use a pre-built HTML column for the click popup (see Section 6)
nt_maplibre(balt) |> nt_add_choropleth(balt, "nt_conc", popup = "popup")

# No popup at all
nt_maplibre(balt) |> nt_add_choropleth(balt, "nt_conc", popup = FALSE)
```

`afford()` precomputes a `popup` column for you, so a rich popup is free:

```{r}
nt_map(aff, color = "tr_rent_ratio", popup = "popup",
       legend_title = "Affordable-rental ratio")
```

### The escape hatch: `fill_color`

For bespoke maps you can hand `nt_add_choropleth()` a raw MapLibre paint
expression (built with `mapgl::interpolate()`, `step_expr()`, or `match_expr()`)
and it uses it **verbatim** as the fill, taking full control of coloring. The
column is still used for the popup, tooltip, and `id`. Because a raw expression
carries no legend, supply `colors` *and* `labels` if you want one.

```{r}
expr <- mapgl::interpolate(
  column = "tr_rent_ratio",
  values = c(0, 1, 2),
  stops  = c("#fee5d9", "#fb6a4a", "#cb181d")
)

nt_maplibre(aff) |>
  nt_add_choropleth(
    aff, "tr_rent_ratio",
    fill_color = expr,
    colors     = c("#fee5d9", "#fb6a4a", "#cb181d"),  # only for the legend
    labels     = c("0", "1", "2")
  )
```

## 5. Multiple layers with a switcher: `nt_layers_control()`

A single map can carry several layers and let the reader switch between them.
Give each layer its own `id`, hide all but the first with `visible = FALSE`, then
add `nt_layers_control()` — a compact **radio** control that shows one layer at a
time and reveals only that layer's legend.

```{r}
demos <- c(White = "pWhite", Black = "pBlack", Asian = "pAsian", Latine = "pLatine")

m <- nt_maplibre(balt) |>
  nt_add_choropleth(balt, "nt_conc", id = "Typology",
                    legend_title = "Neighborhood type")

# Add a hidden layer for each demographic share
for (i in seq_along(demos)) {
  m <- nt_add_choropleth(m, balt, demos[[i]], id = names(demos)[i],
                         legend_title = paste("Share", names(demos)[i]),
                         visible = FALSE)
}

nt_layers_control(m, layers = c("Typology", names(demos)), title = "Show layer")
```

```{r layers-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Example output: typology plus a hidden layer for each demographic share, with a switcher (bottom-right) that shows one layer -- and only its legend -- at a time."}
fig("map-layers.png")
```

### Every argument

| Argument | Default | What it does |
|---|---|---|
| `map` | — | A map with two or more layers added via `nt_add_choropleth()`. |
| `layers` | — | Character vector of layer `id`s to switch between, in order. The first shows initially. |
| `labels` | `NULL` | Display labels for the control (defaults to `layers` — use this to show friendly names while switching by id). |
| `title` | `NULL` | Optional title shown above the radio options. |
| `position` | `"bottom-right"` | `"bottom-right"`, `"bottom-left"`, `"top-left"`, or `"top-right"`. The default keeps it clear of the legend (top-left) and the navigation/search controls (top-right). |

```{r}
# Switch by layer id but display readable labels; place the control top-left
nt_layers_control(
  m,
  layers   = c("Typology", "White", "Black", "Asian", "Latine"),
  labels   = c("Neighborhood type", "% White", "% Black", "% Asian", "% Latine"),
  title    = "Measure",
  position = "top-left"
)
```

The same pattern drives the HPRM map's eviction-risk / displacement-risk /
composite toggle. For a raw multi-select (checkbox) toggle instead of a
one-at-a-time switcher, drop to `mapgl::add_layers_control()`.

## 6. Labels: `nt_add_labels()`

Add place or value labels with `nt_add_labels()`. Labels sit at an *interior*
point of each polygon (computed with `sf::st_point_on_surface()`, not a centroid),
so they never drift into a bay or notch, and they carry a halo for legibility.

```{r}
nt_maplibre(balt) |>
  nt_add_choropleth(balt, "nt_conc") |>
  nt_add_labels(balt, "NAME", min_zoom = 11)
```

### Every argument

| Argument | Default | What it does |
|---|---|---|
| `map` | — | A map from `nt_maplibre()`. |
| `data` | — | Spatial data; polygons are reduced to interior label points. |
| `text_column` | — | Column whose values are shown as labels. |
| `id` | `NULL` | Layer id (defaults to a name derived from `text_column`). |
| `size` | `11` | Text size in pixels. |
| `color` | `"#333333"` | Text color (dark grey). |
| `halo_color` | `"#ffffff"` | Halo color (white). |
| `halo_width` | `1.2` | Halo width in pixels. |
| `min_zoom` | `NULL` | Minimum zoom at which labels appear — handy to de-clutter the zoomed-out view. |
| `...` | — | Passed to `mapgl::add_symbol_layer()`. |

```{r}
nt_maplibre(balt) |>
  nt_add_choropleth(balt, "nt_conc") |>
  nt_add_labels(
    balt, "NAME",
    size       = 13,
    color      = "#19222C",   # near-black text
    halo_color = "#ffffff",
    halo_width = 2,           # fatter halo for contrast over busy fills
    min_zoom   = 12           # only show once zoomed in
  )
```

## 7. Popups, from plain to HPRM-rich: `nt_popup()`

By default the choropleth popup lists the tract id and the mapped value. To
control it, build your own HTML with `nt_popup()` and pass the resulting column
to `popup =`. `nt_popup()` returns one HTML string per row and is fully
**deterministic** — same data in, byte-for-byte identical HTML out.

### Every argument

| Argument | Default | What it does |
|---|---|---|
| `data` | — | A data frame or `sf` object (geometry is ignored). |
| `fields` | `NULL` | Named vector/list mapping **display label → column name**, rendered in order. |
| `title` | `NULL` | Bold title at the top. If it names a column, the per-row value is used; otherwise it is a literal string shared by every popup. |
| `html` | `NULL` | Extra HTML (length 1 or `nrow(data)`) appended after the fields — bar charts, SVG, links. |
| `prefix_title` | `TRUE` | When `title` names a column, prefix the value with the column name, e.g. `"GEOID: 24510...";` set `FALSE` for the bare value. |

### Start simple — a title and a few fields

```{r}
balt$popup <- nt_popup(
  balt,
  title  = "GEOID",
  fields = c(
    "Neighborhood type" = "nt_conc",
    "Total population"   = "totraceE",
    "Share Black"        = "pBlack",
    "Share White"        = "pWhite"
  )
)

nt_map(balt, popup = "popup")
```

Numeric fields **format themselves**: values that all fall in `[0, 1]` render as
percentages (e.g. `pBlack` → `"61.4%"`), other numerics get thousands separators
(e.g. `totraceE` → `"4,213"`). You write the columns; the formatting is automatic.

### `prefix_title`: column-name prefix on or off

With `title = "GEOID"`, the default reads `GEOID: 24510010100`. Set
`prefix_title = FALSE` for just the bare value (useful when the title is a place
name rather than an id):

```{r}
# "GEOID: 24510010100"
nt_popup(balt, title = "GEOID", fields = c("Type" = "nt_conc"))

# bare value, no "NAME:" prefix
nt_popup(balt, title = "NAME", fields = c("Type" = "nt_conc"),
         prefix_title = FALSE)

# A literal shared title (not a column) is always shown as-is
nt_popup(balt, title = "Baltimore City", fields = c("Type" = "nt_conc"))
```

### Graft on a chart with `html`

The HPRM popup shows a small racial-composition bar chart. Because a popup is just
HTML, you build it in R and append it through `html`. This is plain string
assembly — no JavaScript, fully reproducible:

```{r}
race_bar <- function(p, color, label) {
  sprintf(
    paste0("<div style='display:flex;align-items:center;gap:4px;margin:1px 0'>",
           "<span style='width:46px;font-size:10px;text-align:right'>%s</span>",
           "<span style='flex:1;background:#eee;height:9px'>",
           "<span style='display:block;height:9px;width:%s%%;background:%s'></span>",
           "</span>",
           "<span style='width:30px;font-size:10px'>%s</span></div>"),
    label, round(p * 100), color, scales::percent(p, accuracy = 1)
  )
}

bars <- with(balt, paste0(
  "<div style='margin-top:6px;font-weight:bold;font-size:11px'>Race &amp; ethnicity</div>",
  race_bar(pBlack,  "#1f78b4", "Black"),
  race_bar(pWhite,  "#C95123", "White"),
  race_bar(pAsian,  "#33a02c", "Asian"),
  race_bar(pLatine, "#e31a1c", "Latine")
))

balt$popup <- nt_popup(
  balt,
  title  = "GEOID",
  fields = c("Neighborhood type" = "nt_conc", "Population" = "totraceE"),
  html   = bars
)

nt_map(balt, popup = "popup")
```

**Bubble tracks** — comparing a tract to its county and the nation — follow the
same idea: join the reference values (a county median and a national median) onto
your data, then build an inline `<svg>` per row positioning a marker on a 0–max
track, and pass it through `html`. Everything stays deterministic because it is
just string math over columns you control. The *precarity-mapping* vignette walks
through that recipe in full.

## 8. Large data: the whole country with PMTiles (`nt_pmtiles()`)

Embedding tens of thousands of polygons directly in a web page is slow. The
state-profile maps solve this with [PMTiles](https://docs.protomaps.com/pmtiles/)
— a single static vector-tile file served over HTTP range requests. The bundled
`us_nt_tracts2024` table covers every U.S. tract.

`nt_add_choropleth()` handles the switch for you. With `tiles = "auto"` (the
default) it draws small layers inline and converts to PMTiles once a layer exceeds
`max_inline` (25,000 features). Force it either way with `tiles = "inline"` or
`tiles = "pmtiles"`:

```{r}
us <- tigris::tracts(cb = TRUE, year = 2024) |>
  left_join(us_nt_tracts2024, by = "GEOID")

nt_maplibre(us) |>
  nt_add_choropleth(us, "nt_conc", tiles = "pmtiles")

# Or raise the inline ceiling so a mid-size layer still draws inline:
nt_maplibre(us) |>
  nt_add_choropleth(us, "nt_conc", tiles = "auto", max_inline = 80000)
```

The PMTiles path calls the
[`tippecanoe`](https://github.com/felt/tippecanoe) command-line tool, an optional
external dependency (install once with e.g. `brew install tippecanoe`). If a layer
exceeds `max_inline` under `tiles = "auto"` but `tippecanoe` is missing, the layer
falls back to drawing inline with a warning; `tiles = "pmtiles"` errors instead.

### Building tiles directly with `nt_pmtiles()`

Need the tiles as a standalone artifact (to host on GitHub Pages, say)? Build them
with `nt_pmtiles()` and wire them up with `mapgl`.

| Argument | Default | What it does |
|---|---|---|
| `data` | — | Spatial data; **all** non-geometry columns are carried into the tiles as feature properties (so popup/tooltip columns survive tiling). |
| `path` | `NULL` | Output `.pmtiles` path (defaults to a temp file). |
| `layer` | `"nt"` | Tile layer name — must equal the `source_layer` used when drawing it. |
| `tippecanoe_args` | `NULL` | Character vector of flags that *replaces* the defaults (`-o`, `-l`, and the input are always set for you). |
| `quiet` | `FALSE` | Suppress tippecanoe's console output and the "wrote ..." message. |

```{r}
tiles <- nt_pmtiles(us, path = "us_typologies.pmtiles", layer = "nt")

nt_maplibre() |>
  mapgl::add_pmtiles_source("us", url = tiles, promote_id = "GEOID") |>
  mapgl::add_fill_layer(id = "nt", source = "us", source_layer = "nt",
                        fill_color = "#1f78b4", fill_opacity = 0.7)
```

The default tippecanoe flags (`-zg --drop-densest-as-needed
--extend-zooms-if-still-dropping --force`) auto-select a zoom range and keep tiles
a reasonable size. Override them wholesale with `tippecanoe_args`, and silence the
console with `quiet = TRUE`:

```{r}
nt_pmtiles(
  us, path = "us_z4_12.pmtiles", layer = "nt",
  tippecanoe_args = c("-Z4", "-z12", "--drop-densest-as-needed", "--force"),
  quiet = TRUE
)
```

Host the resulting `.pmtiles` file alongside your HTML and point the source `url`
at it.

## 9. The leaflet bridge: `nt_pal()`

For **MapLibre** maps you never need `nt_pal()` — `nt_add_choropleth()` applies
the typology color key automatically. But if you are drawing typologies in
**leaflet** instead, `nt_pal(df)` returns a `leaflet::colorFactor()` palette over
the 19 `nt_conc` levels using the same ERN house colors (grey `#C0C0C0` for `NA`).
Both paths read from one internal color vector, so the two engines always agree.

```{r}
pal <- nt_pal(balt)

leaflet::leaflet(balt) |>
  leaflet::addPolygons(fillColor = ~pal(nt_conc), fillOpacity = 0.7,
                       weight = 0.3, color = "#ffffff") |>
  leaflet::addLegend(pal = pal, values = ~nt_conc)
```

## 10. Synchronized maps: `nt_map_sync()`

Comparing layers side by side only works if the reader can look at the *same
place* on every map. `nt_map_sync()` lays two or more maps out in a row and
locks their views together — pan, zoom, or rotate any one and the others follow
instantly. Optional per-panel titles and subtitles ride above each map.

```{r}
md <- ntdf(state = "MD", county = "Baltimore City", geometry = TRUE)

nt_map_sync(
  nt_map(md),                                   # typologies
  nt_map(md, color = "pBlack"),                 # % Black
  nt_map(md, color = "pLatine"),                # % Latine
  titles = c("Neighborhood typologies", "Percent Black", "Percent Latine")
)
```

Maps sync within a *group*: each `nt_map_sync()` call gets its own group by
default, so two rows on one page move independently. To sync maps that are not
side by side (say, one per report section), give both calls the same `group`:

```{r}
nt_map_sync(m1, m2, group = "overview")   # row 1
nt_map_sync(m3, m4, group = "overview")   # elsewhere on the page, same views
```

The synchronization is plain client-side JavaScript (a shared registry and a
re-entrancy guard), so it works in R Markdown, Quarto, the RStudio viewer, and
saved standalone HTML alike. The three-pane affordability comparisons
(`vignette("three-panes")`) are built on it.

## 11. Decluttering the basemap: `nt_declutter_basemap()`

Basemap styles tint parks, landcover, and hillshade. Under a semi-transparent
choropleth those tints read as pale washes over your fill colors — a mottled
"is that a fourth class?" effect. `nt_declutter_basemap()` hides the decorative
land layers once the style loads, leaving roads, water, boundaries, and labels
intact (and it never touches layers this package added):

```{r}
nt_map(md, color = "pBlack") |>
  nt_declutter_basemap()

# a custom pattern, if a style names its layers differently
nt_map(md) |> nt_declutter_basemap(pattern = "park|forest|hillshade")
```

`afford_map()` applies it by default; for any other map it is one pipe step.

## 12. Saving a standalone HTML map

A map is an `htmlwidget`, so a self-contained, shareable HTML file is one call —
the same kind of artifact the ERN state profiles publish:

```{r}
m <- nt_map(balt)
htmlwidgets::saveWidget(m, "baltimore_typologies.html", selfcontained = TRUE)
```

For inline (non-PMTiles) maps the file is fully self-contained. PMTiles maps
additionally need the `.pmtiles` file hosted next to the page.

## 13. When to drop to raw mapgl

The `neighborhood` functions cover the common path. Anything they don't expose,
`mapgl` does — and you never have to leave the pipe to get to it. Reach for
`mapgl` directly for 3-D extrusions (`add_fill_extrusion_layer()`), heatmaps,
custom expressions (`interpolate()`, `step_expr()`, `match_expr()`), draw tools,
bivariate legends, and Shiny integration. See
[the mapgl reference](https://walker-data.com/mapgl/reference/index.html).

## Acknowledgments

This mapping toolkit stands on Kyle Walker's work: **tidycensus** for census data
and **mapgl** for the MapLibre engine. For transparency, the design and
implementation of the `neighborhood` mapping functions were developed with the
assistance of **Claude Opus 4.8** (Anthropic). The functions themselves are
deterministic R code and require no AI to run.

## Issues and feedback

Hit a bug, or have a feature request or feedback? Please
[open a GitHub issue](https://github.com/evictionresearch/neighborhood/issues) —
problem reports and suggestions are always welcome.
</content>
</invoke>
