---
title: "Neighborhood racial typologies: the workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neighborhood racial typologies: the workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
fig <- function(f) {
  p <- file.path("figures", f)
  if (file.exists(p)) knitr::include_graphics(p)
}
```

The neighborhood racial typology is a **descriptive, categorical** measure of
who lives where, at the census-tract scale. Where a single-number segregation
index (such as the [index of
dissimilarity](https://en.wikipedia.org/wiki/Index_of_dissimilarity)) summarizes
a whole region into one figure, the typology labels each tract — "All White",
"Black-White", "Diverse", and so on — so you can map the fine-grained pattern of
who shares (or does not share) a neighborhood.

This vignette walks the end-to-end workflow: build the data with `ntdf()`,
inspect and collapse the types with `ntcheck()`, color a leaflet map with
`nt_pal()` or map it in one line with `nt_map()`, reach for the bundled national
table `us_nt_tracts2024`, pair typologies with `afford_index()`, and — as a
short aside — crosswalk PUMAs to counties with `get_co_puma()` for PUMS
microdata. The classification follows Hall, Crowder, and Spring (2015).[^hall]

[^hall]: Hall, Matthew, Kyle Crowder, and Amy Spring. 2015. "Neighborhood
Foreclosures, Racial/Ethnic Transitions, and Residential Segregation."
*American Sociological Review* 80:526–549.

```{r setup}
library(neighborhood)
library(dplyr)
```

## 1. Build the typologies with `ntdf()`

`ntdf()` pulls the race/ethnicity table (ACS `B03002`) via
[tidycensus](https://walker-data.com/tidycensus/) (Kyle Walker) and classifies
each geography. It computes the population share of five mutually exclusive
groups — White, Black, Asian, Latine, and Other (everyone else, including
multiracial and Native populations) — then assigns a tract to **every group
whose share exceeds 10%**; the combination becomes its type. A group above 90%
makes the tract "All …".

The only required argument is `state`. The simplest call returns a plain
(non-spatial) data frame:

```{r}
balt <- ntdf(state = "MD", county = "Baltimore City")

glimpse(balt)
```

Set `geometry = TRUE` to get back an `sf` object you can map directly:

```{r}
balt_sf <- ntdf(state = "MD", county = "Baltimore City", geometry = TRUE)
```

`ntdf()` needs a Census API key (set once with
`tidycensus::census_api_key("YOUR KEY", install = TRUE)`), so the tidycensus
calls in this vignette are shown but not run.

### Reading the output: `NeighType` vs `nt_conc`

Two classification columns come back:

- **`NeighType`** — the full, fine-grained label. Every group above the 10%
  threshold appears in the name (e.g. `"Black-Latine-White"`), so there are many
  possible combinations.
- **`nt_conc`** — a *concatenated* version that folds the rare multi-group
  combinations into "3 Group Mixed" / "4 Group Mixed" and the single-group
  variants (the "All …" and "…-Shared" labels) into "Mostly …". It is an
  ordered factor with 19 levels, and it is what you usually map.

The **10% concentration threshold** is the heart of the method: a group has to
be at least one-in-ten of a tract's population to "count" toward that tract's
label. That is why a tract that is 95% White and 5% Black reads as "Mostly
White" (via `"White-Shared"`), not "Black-White" — the Black share never clears
the bar. The 19 ordered `nt_conc` levels run from the single-group "Mostly …"
types, through the two-group pairs, up to the mixed and "Diverse" categories and
finally "Unpopulated Tract":

```{r}
levels(balt$nt_conc)
```

### Every `ntdf()` argument

| Argument | Default | What it does |
|---|---|---|
| `state` | *(required)* | One or more states (name, abbreviation, or FIPS), passed straight to tidycensus. |
| `geography` | `"tract"` | Census geography to classify. Any tidycensus level works — `"block group"`, `"county"`, `"county subdivision"`, `"place"`, etc. |
| `county` | `NULL` | Restrict to one or more counties (name or FIPS). `NULL` returns the whole state. |
| `geometry` | `FALSE` | `TRUE` returns an `sf` object with polygons for mapping. |
| `cache_table` | `TRUE` | Passed to tidycensus — cache the variable lookup table locally to speed up repeat calls. |
| `output` | `"wide"` | `"wide"` (one row per geography, the classifier needs this) or `"tidy"`. |
| `year` | `2019` | The ACS 5-year **endpoint** year (so `2019` = the 2015–2019 sample). |
| `...` | — | Any other argument is forwarded to `tidycensus::get_acs()`. |

A different geography — classify block groups instead of tracts:

```{r}
bg <- ntdf(state = "MD", county = "Baltimore City", geography = "block group")
```

A different year — the typology of the same place in an earlier ACS window:

```{r}
balt_2015 <- ntdf(state = "MD", county = "Baltimore City", year = 2015)
```

Because `...` reaches `tidycensus::get_acs()`, you can pass any of its arguments
through. For example, request the 1-year survey instead of the 5-year default:

```{r}
balt_acs1 <- ntdf(state = "MD", county = "Baltimore City",
                  year = 2022, survey = "acs1")
```

The `output = "tidy"` form returns one row per geography × variable (the long
shape) rather than one wide row per geography:

```{r}
balt_tidy <- ntdf(state = "MD", county = "Baltimore City", output = "tidy")
```

Multiple states or counties are vectors:

```{r}
dmv <- ntdf(state = c("MD", "VA", "DC"))                       # whole region
twocty <- ntdf(state = "MD", county = c("Baltimore City", "Howard"))
```

## 2. Inspect with `ntcheck()`

Before mapping, see how the types are distributed. `ntcheck(df)` returns a plain
data frame with one row per `NeighType` and a count `n`, sorted most common
first — useful for deciding whether the default `nt_conc` concatenation is
collapsed enough for your study area, or whether some rare fine-grained types
are worth merging further. Its only argument is the `ntdf()`-created data frame;
the `...` is an unused backward-compatibility slot, so anything you pass there is
silently ignored.

```{r}
ntcheck(balt)
```

It tabulates the detailed `NeighType` column (not `nt_conc`), so it is the right
tool for spotting which long labels are rare. To tabulate the collapsed
`nt_conc` levels instead, a `count()` does the job:

```{r}
balt |> count(nt_conc, sort = TRUE)
```

## 3. The bundled national table: `us_nt_tracts2024`

For a national reference without an API call, the package ships
`us_nt_tracts2024`: a precomputed table of every U.S. tract (plus DC and Puerto
Rico) built with this same `ntdf()` pipeline, using the 5-year ACS centered on
2024 (the 2020–2024 endpoint). It is **non-spatial** (no geometry column — see
§5 for mapping it) and carries two extra columns the live `ntdf()` output does
not: `state` (two-letter postal code) and `year` (2024).

```{r, eval=TRUE}
library(neighborhood)
library(dplyr)
dim(us_nt_tracts2024)
names(us_nt_tracts2024)
```

`ntcheck()` works on it directly — this is the national frequency table:

```{r, eval=TRUE}
head(ntcheck(us_nt_tracts2024), 10)
```

Or slice to one state and tabulate the collapsed levels:

```{r, eval=TRUE}
us_nt_tracts2024 |>
  filter(state == "CA") |>
  count(nt_conc, sort = TRUE)
```

Because the table is national, comparing the racial structure of two states is
one `count()` — and the contrast does real analytical work. California and
Maryland, each state's four most common tract types by share:

```{r, eval=TRUE}
us_nt_tracts2024 |>
  filter(state %in% c("CA", "MD")) |>
  count(state, nt_conc) |>
  group_by(state) |>
  mutate(share = round(n / sum(n), 3)) |>
  slice_max(share, n = 4) |>
  ungroup()
```

Both states lead with three-group mixed tracts, but the mixes are different
mixes: California's integration runs through its Latine population
(Latine–White is its second-largest type, Mostly Latine its third), while
Maryland's runs Black–White (Mostly White, then Black–White, then Mostly
Black). The same typology grammar, applied to two states, surfaces two
different segregation regimes — which is exactly the kind of two-region read
the `afford_index()` pairing in §6 turns into policy analysis.

## 4. Color and map

### Leaflet, via `nt_pal()`

For a **leaflet** map, `nt_pal(df)` returns the canonical color palette — a
`leaflet::colorFactor()` keyed to the 19 `nt_conc` levels, in the Eviction
Research Network house colors, with grey (`#C0C0C0`) for `NA`. Pass it your
`ntdf()` data frame (it reads the `nt_conc` column); like `ntcheck()`, its
trailing `...` is an unused compatibility slot.

```{r}
pal <- nt_pal(balt_sf)

leaflet::leaflet(balt_sf) |>
  leaflet::addProviderTiles("CartoDB.Positron") |>
  leaflet::addPolygons(fillColor = ~pal(nt_conc), fillOpacity = 0.7,
                       weight = 0.3, color = "#ffffff") |>
  leaflet::addLegend(pal = pal, values = ~nt_conc, title = "Neighborhood type")
```

### MapLibre, via `nt_map()`

For an interactive **MapLibre** map you don't even need the palette — `nt_map()`
applies the same color key automatically. When you omit `color` and the data has
an `nt_conc` column, it maps the typologies; otherwise name a column to map.

```{r}
nt_map(balt_sf)                    # typologies (auto-detects nt_conc)
nt_map(balt_sf, color = "pBlack")  # a numeric column instead
```

`nt_map(data, color = NULL, ...)` is shorthand for `nt_maplibre()` piped into
`nt_add_choropleth()`; the `...` forwards to `nt_add_choropleth()` (e.g. `type`,
`breaks`, `legend`, `popup`). See `vignette("mapping-with-maplibre")` for the
full tour.

```{r typology-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Example output: Baltimore City tracts colored by neighborhood typology with the canonical ERN palette."}
fig("map-typologies.png")
```

## 5. Mapping the bundled national table

`us_nt_tracts2024` has no geometry, so join it to tract geometry first. For a
single state this is quick; for the whole country, see the PMTiles section of
`vignette("mapping-with-maplibre")`.

```{r}
md_geo <- tigris::tracts(state = "MD", cb = TRUE, year = 2024) |>
  left_join(filter(us_nt_tracts2024, state == "MD"), by = "GEOID")

nt_map(md_geo)   # nt_conc is present, so it colors typologies automatically
```

## 6. Pairing typologies with affordability

Because both `ntdf()` output and `afford_index()` are keyed on the tract
`GEOID`, you can join them and ask equity questions — for example, *is the
housing affordable to low-income renters concentrated in, or absent from,
neighborhoods of color?*

`afford_index(state, counties, ...)` returns a **long** tibble with one row per
tract × tenure × tier. Its tiers are the standardized ELI/VLI/LI/MI ceilings
(Extremely-/Very-/Low-/Moderate-Income, as cumulative fractions of Area Median
Income), and its tenures are `"rent"`, `"own_current"` (current owner cost), and
`"own_buyin"` (the income needed to buy in today). The key takeaway for mapping:
**filter to a single tenure and tier first**, otherwise each tract appears
several times.

```{r}
idx <- afford_index("24", "510", year = 2024)   # Baltimore City, all tenures/tiers

# Keep one slice: rental supply affordable to Very-Low-Income households
vli_rent <- idx |>
  filter(tenure == "rent", ami_tier == "VLI")

equity <- balt_sf |>
  left_join(vli_rent, by = "GEOID")

# `ratio` is a location quotient: supply vs. the region's share of VLI households
nt_map(equity, color = "ratio")
```

Per row, the most useful columns are `supply` (accessible units ÷ total units in
the tract), `ratio` (that supply divided by the region's share of households in
the tier — a location quotient, where 1 is parity), and `rate` (accessible units
per 100,000 regional tier households). To attach polygons in one step instead of
joining, pass `geometry = TRUE` — but still filter to one tenure/tier before
mapping, since geometry duplicates across slices:

```{r}
idx_sf <- afford_index("24", "510", year = 2024, geometry = TRUE)
idx_sf |>
  filter(tenure == "rent", ami_tier == "VLI") |>
  nt_map(color = "ratio")
```

`afford_index()` exposes the full cost model — the owner-buy-in mortgage
assumptions (`interest_rate`, `term_years`, `down_pct`, `tax_ins_rate`), the AMI
source (`ami_source`), the tier fractions (`ami_tiers`), and the household-size
adjustment (`hud_hh_size`). See `vignette("affordability-index")` for the full
methodology and every argument.

## 7. Aside: PUMA ↔ county crosswalk with `get_co_puma()`

Tract typologies cover the published ACS tables, but individual-level questions
(household structure, age, tenure interactions) need **PUMS microdata**, which is
geocoded to PUMAs (Public Use Microdata Areas), not tracts or counties. PUMAs and
counties do not nest cleanly — in dense cities several PUMAs sit inside one
county; in rural areas one PUMA spans several counties. `get_co_puma(st, yr)`
builds the crosswalk by spatially joining PUMA polygons to county-tagged tract
centroids, so you can attach a county to each PUMA.

```{r}
study_pumas <- get_co_puma("OR", 2021)   # st = state(s), yr = PUMA vintage year
```

It returns a data frame of PUMA and county FIPS, including a `PUMACE…` code
column you can join to `tidycensus::get_pums()` output (Kyle Walker), plus
`state`, `state_name`, and a cleaned `county` name. `st` and `yr` are vectorized
(both are walked in parallel), and `...` forwards to `tigris::tracts()`.

```{r}
pums <- tidycensus::get_pums(
  variables = c("PUMA", "TEN", "AGEP"),
  state = "OR", survey = "acs5", year = 2021)

study_pums <- left_join(study_pumas, pums, by = c("PUMACE10" = "PUMA"))
```

See `?get_co_puma` for a worked renter-by-race-and-sex example end to end.

## Where to go next

- `vignette("affordability-index")` — the `afford_index()` methodology and every
  cost-model argument.
- `vignette("mapping-with-maplibre")` — the full MapLibre toolkit
  (`nt_maplibre()`, `nt_add_choropleth()`, `nt_add_labels()`, PMTiles for
  national-scale maps).

## Acknowledgments

Census data come from Kyle Walker's **tidycensus** and **tigris**. The MapLibre
mapping functions used here were developed with the assistance of **Claude Opus
4.8** (Anthropic) for transparency; they are deterministic and require no AI to
run.

## Issues and feedback

Hit a bug, or have a feature request or feedback? Please
[open a GitHub issue](https://github.com/evictionresearch/neighborhood/issues) —
problem reports and suggestions are always welcome.
