---
title: "Advanced: an interactive precarity map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: an interactive precarity map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
fig <- function(f) {
  p <- file.path("figures", f)
  if (file.exists(p)) knitr::include_graphics(p)
}
```

The Eviction Research Network's flagship maps put a **housing-precarity measure**
on the map and let the reader explore it — switch between related measures, read
a tract's standing in a rich popup, and filter to the places that clear a
threshold. This advanced tutorial rebuilds that whole experience with the
`neighborhood` mapping functions, using **only public data you can pull yourself**
(the ACS via [tidycensus](https://walker-data.com/tidycensus/), plus the bundled
`us_nt_tracts2024` typologies). Nothing here depends on a proprietary index.

The precarity score we build is **illustrative** — a transparent composite of a
few ACS measures, meant to demonstrate the mapping tooling, not to stand in for a
validated model. Swap in your own measure and every step below carries over.

The functions on tour here are `nt_maplibre()` (the base map), `nt_add_choropleth()`
(filled layers), `nt_layers_control()` (the measure switcher), `nt_popup()` (the
rich click panel), and `nt_pmtiles()` (vector tiles for the at-scale version).

```{r setup}
library(neighborhood)
library(dplyr)
library(sf)
library(tigris)
library(tidycensus)
options(tigris_use_cache = TRUE)
```

## 1. Build a precarity measure from the ACS

We work one county at a time so the map stays inline and quick to iterate; step 7
scales the same recipe to the whole country with vector tiles. Here: Hennepin
County, Minnesota (Minneapolis). Pull a handful of renter-precarity ingredients
in one call — renter share, rent burden, median rent and income, and population
(to drop near-empty tracts):

```{r}
acs_vars <- c(
  renters  = "B25003_003", occ_hh   = "B25003_001",   # renter / occupied households
  rb_30_35 = "B25070_007", rb_35_40 = "B25070_008",    # gross rent as a share of income
  rb_40_50 = "B25070_009", rb_50p   = "B25070_010",
  rb_total = "B25070_001", rb_na    = "B25070_011",
  med_rent = "B25064_001", med_inc  = "B19013_001",
  pop      = "B01003_001"
)

tr <- get_acs(geography = "tract", state = "MN", county = "Hennepin",
              variables = acs_vars, year = 2022, output = "wide",
              geometry = TRUE) |>
  st_transform(4326)
```

Derive the component measures, then a composite. Each component is turned into a
**percentile rank** (0–100) within the county and averaged, so the composite
reads as "this tract is more precarious than X% of the county." Lower income
means *higher* precarity, so we rank its negative:

```{r}
pctl <- function(x) dplyr::percent_rank(x) * 100

prec <- tr |>
  transmute(
    GEOID,
    pop          = popE,
    med_rent     = med_rentE,
    med_inc      = med_incE,
    renter_share = rentersE / occ_hhE,
    rent_burden  = (rb_30_35E + rb_35_40E + rb_40_50E + rb_50pE) /
                     pmax(rb_totalE - rb_naE, 1)            # share of renters paying >= 30%
  ) |>
  filter(pop >= 500, !is.na(renter_share), !is.na(rent_burden)) |>
  mutate(
    precarity = rowMeans(cbind(pctl(rent_burden), pctl(renter_share), pctl(-med_inc)))
  )
```

We **filter** out near-empty tracts here so percentiles aren't distorted by a
20-person tract. Section 2.1 shows an alternative: keep them and **grey them out**
on the map instead of dropping them.

Join the bundled racial typologies on `GEOID` — we use the group shares for the
popup bar chart and the typology label in the field table:

```{r}
prec <- prec |>
  left_join(
    us_nt_tracts2024 |>
      select(GEOID, nt_conc, totraceE, pWhite, pBlack, pAsian, pLatine, pOther),
    by = "GEOID"
  )
```

## 2. Three switchable layers (the legend follows the layer)

Add the composite and its two main components as separate layers — each with its
own `id`, all but the first `visible = FALSE` — then wire a switcher with
[nt_layers_control()]. Picking a measure shows that layer **and its legend**, and
hides the others.

A 5-class sequential ramp reads well for a 0–100 percentile; pass it as `colors`
(with matching `labels`) so every layer uses the same legend grammar. For a
`"step"` (binned) layer, each `breaks` vector normally has one fewer entry than
`colors`. The third layer below leaves `colors` unset, so it falls back to the
default 4-class ERN ramp, and shows the convenience case: if you pass *as many*
breaks as colors (a leading floor — here `c(0, 0.25, 0.5, 0.75)` against the
4-class ramp), `nt_add_choropleth()` drops the floor for you, so the same call
works whether or not you include it.

```{r}
ramp5 <- c("#ffffcc", "#fed976", "#fd8d3c", "#f03b20", "#bd0026")

m <- nt_maplibre(prec) |>
  nt_add_choropleth(prec, "precarity", id = "Precarity",
                    breaks = c(20, 40, 60, 80), colors = ramp5,
                    labels = c("0-20", "20-40", "40-60", "60-80", "80-100"),
                    legend_title = "Precarity percentile",
                    popup = "popup", opacity = 0.85) |>
  nt_add_choropleth(prec, "rent_burden", id = "Rent burden",
                    breaks = c(0.35, 0.45, 0.55, 0.65), colors = ramp5,
                    labels = c("<35%", "35-45%", "45-55%", "55-65%", "65%+"),
                    legend_title = "Renters paying 30%+ of income",
                    popup = "popup", opacity = 0.85, visible = FALSE) |>
  nt_add_choropleth(prec, "renter_share", id = "Renter share",
                    breaks = c(0, 0.25, 0.5, 0.75),   # 4 breaks, leading floor dropped
                    labels = c("<25%", "25-50%", "50-75%", "75%+"),
                    legend_title = "Share of households renting",
                    popup = "popup", opacity = 0.85, visible = FALSE)

m <- nt_layers_control(m, layers = c("Precarity", "Rent burden", "Renter share"),
                       title = "Measure")
```

(We pass `popup = "popup"` ahead of building that column — that happens next.)

### Renaming and repositioning the switcher

The `layers` argument names the layer `id`s to toggle, in order; the **first is
shown initially** and the rest must have been added with `visible = FALSE`. By
default the switcher prints those same ids as its option labels. Pass `labels` to
show friendlier text without renaming the layers themselves (it must be the same
length as `layers`), and `position` to move the control. `position` is one of
`"bottom-right"` (the default), `"bottom-left"`, `"top-left"`, or `"top-right"`;
the default keeps it clear of the legend (top-left) and the search/navigation
controls (top-right).

```{r}
m <- nt_layers_control(
  m,
  layers   = c("Precarity", "Rent burden", "Renter share"),
  labels   = c("Composite precarity", "Rent burden (30%+)", "Renters (%)"),
  title    = "Measure",
  position = "bottom-left"
)
```

For a raw multi-select **checkbox** toggle (several layers visible at once)
instead of this one-at-a-time radio, drop to `mapgl::add_layers_control()`.

```{r precarity-map, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="An illustrative precarity map for Hennepin County, MN: the composite percentile layer with the ERN sequential ramp, the measure switcher (Precarity / Rent burden / Renter share), and the legend that follows the chosen layer. Step 7 scales the same recipe to all US tracts via PMTiles."}
fig("map-precarity.png")
```

### 2.1 Grey the filtered tracts instead of dropping them

In step 1 we *removed* low-population tracts. Sometimes you want them on the map
but visibly **not measured** — a neutral grey, so the reader sees coverage gaps
instead of holes. Keep those tracts in the data with a missing (`NA`) measure,
and `nt_add_choropleth()` fills them with `na_color` (default ERN grey
`#C0C0C0`). `outline_color` sets the polygon border (default white); a faint grey
border often reads better over a light basemap.

```{r}
prec_all <- prec |>
  mutate(precarity = ifelse(pop < 500, NA_real_, precarity))   # keep tract, blank the score

nt_maplibre(prec_all) |>
  nt_add_choropleth(prec_all, "precarity", id = "Precarity",
                    breaks = c(20, 40, 60, 80), colors = ramp5,
                    labels = c("0-20", "20-40", "40-60", "60-80", "80-100"),
                    legend_title = "Precarity percentile",
                    na_color = "#d9d9d9",        # grey for the unmeasured tracts
                    outline_color = "#9aa6b2",   # soft border instead of white
                    opacity = 0.85)
```

### `nt_add_choropleth()` argument reference

| Argument | What it does |
|---|---|
| `column` | Column to color by. Numeric → binned ERN ramp; factor/character → categorical (`nt_conc` gets the canonical typology palette). |
| `type` | `"auto"` (default), `"categorical"`, `"continuous"` (smooth), or `"step"` (binned). |
| `breaks` | Interior thresholds for `"step"`/`"continuous"`; one fewer than `colors`, or pass an equal-length leading-floor vector and the floor is dropped. Defaults to data quantiles. |
| `colors` | Fill colors — one per category, or the numeric ramp (defaults to the 4-class ERN ramp). |
| `labels` | Legend labels for binned numeric maps. |
| `fill_color` | A raw MapLibre paint expression used verbatim (escape hatch); pass `colors` + `labels` too if you want a legend. |
| `id` | Layer id (defaults from `column`); the handle the switcher and `setFilter()` use. |
| `legend` | `TRUE` (default), `FALSE`, or `"interactive"` (clickable, filters the map). |
| `legend_title` | Legend title (defaults to `column`). |
| `popup` / `tooltip` | `TRUE` default, `FALSE`/`NULL`, a column name, or a length-`nrow(data)` vector. |
| `opacity` | Fill opacity (default `0.7`). |
| `outline_color` | Polygon border color (default white `#ffffff`). |
| `na_color` | Fill for missing values (default ERN grey `#C0C0C0`). |
| `visible` | Start visible? `FALSE` for all-but-one layer under a switcher. |
| `tiles` / `max_inline` | `"auto"` (PMTiles past `max_inline` = `25000` features), `"inline"`, or `"pmtiles"`. |
| `...` | Passed to `mapgl::add_fill_layer()`. |

## 3. The rich popup: race bars and "bubble tracks"

The signature popup shows two things beyond a field table: a small **bar chart**
of racial composition, and **bubble tracks** that place the tract on a scale
between its county and the nation. Both are just HTML strings you build with
`sprintf()`, so they stay fully deterministic. Pass them to [nt_popup()] via
`html =`.

First the racial-composition bars from the typology shares:

```{r}
race_bars <- function(d) {
  groups <- c(White = "pWhite", Black = "pBlack", Asian = "pAsian",
              Latine = "pLatine", Other = "pOther")
  cols   <- c("#4575b4", "#d73027", "#fc8d59", "#fee090", "#91bfdb")
  rows <- vapply(seq_len(nrow(d)), function(i) {
    bars <- vapply(seq_along(groups), function(g) {
      val <- d[[groups[g]]][i]
      pct <- round(100 * ifelse(is.na(val), 0, val))
      sprintf(paste0("<div style='display:flex;align-items:center;gap:6px;margin:1px 0'>",
                     "<span style='width:46px;font-size:10px;color:#586573'>%s</span>",
                     "<span style='flex:1;background:#eef1f4;border-radius:2px'>",
                     "<span style='display:block;height:7px;width:%d%%;background:%s;border-radius:2px'></span>",
                     "</span><span style='width:30px;font-size:10px;text-align:right'>%d%%</span></div>"),
              names(groups)[g], pct, cols[g], pct)
    }, character(1))
    paste(bars, collapse = "")
  }, character(1))
  rows
}
```

Then a bubble track — a colored rail with a marker for the tract and a tick for
the county reference:

```{r}
bubble <- function(value, county, lo, hi, fmt) {
  pct <- function(x) pmax(0, pmin(100, (x - lo) / (hi - lo) * 100))
  sprintf(
    paste0("<div style='position:relative;height:20px;margin:4px 0'>",
           "<div style='position:absolute;top:8px;left:0;right:0;height:6px;",
           "background:linear-gradient(to right,#d1e5f0,#4393c3,#2166ac);border-radius:3px'></div>",
           "<div style='position:absolute;top:5px;left:%s%%;width:11px;height:11px;",
           "background:#fff;border:2px solid #333;border-radius:50%%;transform:translateX(-50%%)'></div>",
           "<div style='position:absolute;top:4px;left:%s%%;width:2px;height:13px;background:#c0392b'></div>",
           "</div><div style='font-size:10px;color:#586573'>Tract %s &middot; County %s</div>"),
    pct(value), pct(county), fmt(value), fmt(county))
}
```

The bubble tracks need county and national reference values. Pull the county
medians from the ACS and use the published national medians:

```{r}
co <- get_acs(geography = "county", state = "MN", county = "Hennepin",
              variables = c(inc = "B19013_001", rent = "B25064_001"),
              year = 2022, output = "wide")
co_inc <- co$incE; co_rent <- co$rentE
us_inc <- 75149; us_rent <- 1348   # 2022 ACS 1-year national medians

dollars <- function(x) paste0("$", format(round(x), big.mark = ","))

prec$popup <- nt_popup(
  prec,
  title  = "GEOID",
  fields = c("Neighborhood type" = "nt_conc",
             "Precarity percentile" = "precarity",
             "Population" = "pop"),
  html = paste0(
    "<div style='font-size:11px;font-weight:600;margin-top:6px'>Racial composition</div>",
    race_bars(prec),
    "<div style='font-size:11px;font-weight:600;margin-top:6px'>Median gross rent</div>",
    bubble(prec$med_rent, co_rent, 0, 3000, dollars),
    "<div style='font-size:11px;font-weight:600;margin-top:4px'>Median household income</div>",
    bubble(prec$med_inc, co_inc, 0, 200000, dollars)
  )
)
```

With the `popup` column built, re-run the step-2 pipe (it already references
`popup = "popup"`) and every layer gets the rich popup on click.

### Title prefixing

When `title` names a column (as `"GEOID"` does above), `nt_popup()` by default
prefixes the label, so the popup header reads `GEOID: 27053026200`. That is handy
for an ambiguous id, but for a value that already reads as a name you usually want
the **bare value**. Set `prefix_title = FALSE`:

```{r}
# Header reads "27053026200" (bare) rather than "GEOID: 27053026200"
nt_popup(prec, title = "GEOID", prefix_title = FALSE,
         fields = c("Neighborhood type" = "nt_conc"))
```

`prefix_title` only applies when `title` matches a column name. A `title` that is
**not** a column is treated as a literal string shared by every popup, and the
prefix never applies — `title = "Hennepin tract"` is the same banner on every
feature. Numeric `fields` are formatted automatically: values that all sit in
`[0, 1]` render as percentages, other numerics get thousands separators.

## 4. A threshold slider

The other piece of the flagship experience is a slider that filters the map to
the places above a chosen precarity level. mapgl doesn't ship one for a static
widget, but it is a few lines of deterministic JavaScript over the standard
MapLibre API — the same way [nt_layers_control()] is built. It adds a range input
and calls `setFilter()` on the `Precarity` layer as you drag:

```{r}
m <- htmlwidgets::onRender(m, "
function(el, x) {
  var map = el.map || (window.HTMLWidgets && HTMLWidgets.getInstance(el) ?
                       HTMLWidgets.getInstance(el).getMap() : null);
  if (!map) return;
  var box = document.createElement('div');
  box.style.cssText = 'position:absolute;z-index:3;left:10px;bottom:28px;background:#fff;' +
    'padding:8px 11px;border-radius:4px;box-shadow:0 1px 4px rgba(25,34,44,.25);' +
    'font:13px Inter,system-ui,sans-serif;color:#19222C';
  box.innerHTML = '<div style=\"font-weight:600\">Min precarity: <span id=\"pv\">0</span></div>' +
    '<input id=\"ps\" type=\"range\" min=\"0\" max=\"100\" value=\"0\" step=\"5\" style=\"width:170px\">';
  function apply(v) {
    document.getElementById('pv').textContent = v;
    if (map.getLayer('Precarity'))
      map.setFilter('Precarity', ['>=', ['get', 'precarity'], +v]);
  }
  box.querySelector('#ps').addEventListener('input', function(e) { apply(e.target.value); });
  if (map.getLayer('Precarity')) el.appendChild(box); else map.on('load', function(){ el.appendChild(box); });
}
")
```

Because we filtered the data in R rather than on the map, `setFilter()` here is
the only filter on the layer — drag the slider and tracts below the threshold
drop out. (To filter a *categorical* layer instead, swap the expression for
`['in', ['get', 'measure'], ['literal', ['High', 'Extreme']]]`.)

If you took the section-2.1 route and kept low-population tracts as `NA`, note
that `['>=', ['get', 'precarity'], v]` already excludes them (a missing value
fails the comparison), so the grey tracts stay out of the filtered view too.

## 5. Save it

A map is an `htmlwidget`; write a standalone, shareable HTML file with
`htmlwidgets::saveWidget()` — the same delivery format as the live state-profile
maps:

```{r}
htmlwidgets::saveWidget(m, "precarity-hennepin.html", selfcontained = TRUE)
```

## 6. Dropping to mapgl

Every `nt_*` function returns an ordinary `mapgl` map, so anything the wrappers
don't expose is one more pipe to a `mapgl::` function — a dot/hatch
`fill_pattern` overlay for a flagged subset, a second basemap, `add_image()`,
custom controls, and so on:

```{r}
m |>
  mapgl::add_fill_layer(
    id = "high_burden_hatch", source = prec, fill_pattern = "dot-9",
    filter = list(">=", list("get", "rent_burden"), 0.6)
  )
```

## 7. The full map at scale: PMTiles

A statewide or national precarity map is too many tracts to embed inline. Build
the measure for every tract (loop the step-1 recipe over states, or pull national
ACS at once), then tile it once with [nt_pmtiles()] (it shells out to
`tippecanoe`), host the `.pmtiles`, and point a layer at it.

`nt_pmtiles()` carries **all non-geometry columns** into the tiles as feature
properties, so your `popup`/`tooltip` columns survive tiling. The `layer` name
written into the archive (default `"nt"`) **must match** the `source_layer` you
draw with. The default tippecanoe flags auto-select a zoom range and shed the
densest features as needed; pass `tippecanoe_args` to replace them wholesale (the
output `-o`, layer `-l`, and input are always set for you), and `quiet = TRUE` to
silence tippecanoe's console output.

```{r}
tiles <- nt_pmtiles(prec_us, path = "precarity_2022.pmtiles", layer = "precarity")

nt_maplibre() |>
  mapgl::add_pmtiles_source("precarity", url = tiles, promote_id = "GEOID") |>
  mapgl::add_fill_layer(id = "Precarity", source = "precarity",
                        source_layer = "precarity",
                        fill_color = mapgl::step_expr(
                          column = "precarity", base = "#ffffcc",
                          values = c(20, 40, 60, 80),
                          stops  = c("#fed976", "#fd8d3c", "#f03b20", "#bd0026")),
                        fill_opacity = 0.85)
```

`mapgl::step_expr()` is the tile-side equivalent of the `breaks`/`colors` you gave
`nt_add_choropleth()` inline: `base` is the color below the first `values` break,
and each subsequent `stops` color applies above its break. Keeping the same
thresholds and ramp makes the tiled national map read identically to the inline
county one.

`nt_add_choropleth(..., tiles = "pmtiles")` will also do the tiling for you when a
layer exceeds the inline threshold (`max_inline`, default 25,000 features) — it
calls `nt_pmtiles()` under the hood with `layer = "nt"` and wires the source for
you, so you rarely need the explicit three-step form above unless you want to host
the archive separately.

## 8. The model-driver plots: `nt_pd_plot()` and `nt_pd_bart()`

The maps show *where* precarity concentrates; the partial-dependence plots show
*why the model thinks so*. `nt_pd_plot()` draws the house figure used for the
HPRM EDR/EER driver plates — the posterior-mean partial-dependence curve with
its credible ribbon, over a frequency histogram of the predictor — so a reader
sees both the modeled relationship and where the data actually sit.

`nt_pd_plot()` is model-agnostic: it takes a tidy PD table (predictor value,
mean, lower, upper) plus the observed predictor values, and needs no Java.

```{r}
# pd: a tidy table with columns value / mean / lower / upper
# obs: the observed predictor values (for the frequency panel)
nt_pd_plot(pd, obs,
           xlab = "Median rent",
           ylab = "Partial effect on displacement risk")
```

`nt_pd_bart()` is the bridge that *computes* that table from a fitted
`bartMachine` model (server-side only: it needs the `bartMachine` package and a
Java runtime — which is why the plotting is split out):

```{r}
pd <- nt_pd_bart(fit, "med_rent")   # posterior PD table from the fitted model
nt_pd_plot(pd, training_data$med_rent)
```

## Acknowledgments

Mapping is built on Kyle Walker's **mapgl** and **tidycensus**; tiles use
**tippecanoe**. The bundled typologies come from the Eviction Research Network.
The precarity composite here is illustrative — a teaching device, not a validated
model. For transparency, the `neighborhood` mapping functions used here were
developed with the assistance of **Claude Opus 4.8** (Anthropic); they are
deterministic and require no AI to run.

## Issues and feedback

Hit a bug, or have a feature request or feedback? Please
[open a GitHub issue](https://github.com/evictionresearch/neighborhood/issues) —
problem reports and suggestions are always welcome.
</content>
</invoke>
