---
title: "State-profile page builders: the nt_profile_* family"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State-profile page builders: the nt_profile_* family}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(neighborhood)
# a peek helper: show the head of a long markup string
peek <- function(s, n = 400) cat(substr(paste(s, collapse = ""), 1, n), "…\n")
```

The ERN state-profile pages (`evictionresearch.net/<state>/` — Washington is
the reference implementation) are **not** htmlwidget pages. Their figures are
hand-tuned inline SVG, and their interactivity (hover tooltips, the
constant-px chart text, the MapLibre county explorer, the sortable table) is
a vanilla-JS layer that ships with the page template and binds to specific
element ids and classes.

The `nt_profile_*` builders generate those exact fragments as plain character
strings, from the same d3/o3 pipeline outputs the profiles have always used.
A profile's Quarto document calls them from `results: asis` chunks via
`nt_raw_html()`, which wraps the markup in a pandoc `{=html}` block so it
passes through untouched. The complete working page — frontmatter, prose,
chunks, and the format extension carrying the interaction JS — lives in
`evictionresearch/library/templates/state_profile_qmd/`; this vignette
documents each builder on its own.

**The markup is a contract.** The page JS looks up ids like
`chart-monthly-svg` and classes like `.data`, `.focal`, `.moratoria-band`,
`.hover-grid`; the `--sx` scale-watcher sizes in-SVG text via
`font-size: calc(<px> / var(--sx, …))`. Renaming ids/classes in the output
breaks the page's behavior, which is why these functions emit fixed markup
and expose *data* as arguments, not structure.

## Figure 1 — the monthly trend line

`nt_profile_trend_svg()` draws the statewide monthly series: a shaded
moratorium band, gridlines derived from the same value→pixel map as the data
(so labels can never drift from the geometry), the dashed pre-pandemic
baseline, the filings path, and a focal dot + callout on the latest month.

```{r fig1}
monthly <- data.frame(
  date    = seq(as.Date("2019-01-01"), by = "month", length.out = 64),
  filings = round(1500 + 400 * sin(seq_len(64) / 5))
)
svg <- nt_profile_trend_svg(monthly$date, monthly$filings,
                            baseline   = 1378,
                            band_start = as.Date("2020-03-01"),
                            band_end   = as.Date("2021-11-01"))
peek(svg)
```

The hover tooltips need the same series in JS. `nt_profile_monthly_js()`
embeds it as a `window.*` global (the page JS reads
`window.__waMonthly.filings` / `.yearBase` / `.monthBase` / `.yDiv`). The
`y_div` argument **must equal** the y ceiling the SVG derived — the series
max rounded up to `y_step` — or hover positions drift from the drawn line:

```{r fig1-js}
y_div <- max(500, ceiling(max(c(monthly$filings, 1378)) / 500) * 500)
peek(nt_profile_monthly_js(monthly$filings, year_base = 2019, y_div = y_div), 160)
```

## Figure 2 — yearly bars with a projection

`nt_profile_yearly_svg()` draws one labeled bar per year, accents the partial
latest year, and (when given a projection) overlays the 80% prediction-
interval band, the dashed extension to the point estimate, and its tick +
label. Bars carry `data-key` / `data-label` / `data-value` (and
`data-proj-*`) attributes for the tooltip JS.

```{r fig2}
years <- data.frame(year = 2016:2022,
                    filings = c(18157, 17682, 16938, 15190, 4748, 3986, 6200))
svg <- nt_profile_yearly_svg(years$year, years$filings,
                             baseline = 16532, latest_month = 4,
                             projection = list(total = 19500, lo = 17000, hi = 22000))
peek(svg)
```

The projection itself comes from `nt_profile_project_year()`, which fits
`forecast::auto.arima` (seasonal) to the monthly series from `fit_start`
onward, forecasts the rest of the calendar year, and returns
`list(total, lo, hi, ytd, h, order)` — or `NULL` once the year is complete,
so the overlay (and any prose gated on it) disappears on its own:

```{r sarima, eval = FALSE}
proj <- nt_profile_project_year(state_agg$date, state_agg$filings,
                                fit_start = "2023-01-01")
proj$order   # e.g. "SARIMA(3,1,0)(1,0,0)[12]"
```

On the Washington data this refits deterministically — the same series and
forecast version reproduce the shipped 26,314 (80% PI 22,874–29,754).

## Figure 3 — the county-trend chart

This chart is drawn client-side (d3, Catmull-Rom smoothing) from an embedded
payload. Two builders split the work: `nt_profile_county_trend()` emits the
shell (granularity toggle, county-highlight dropdown, the empty `<svg>`, the
caption), and `nt_profile_county_trend_data()` computes the payload from the
county-month aggregate — per-county yearly and monthly vectors plus a
`rec` flag marking counties whose record full year fell in the last
`record_window` years (the chart colors those lines red).

```{r fig3}
county_agg <- expand.grid(month = 1:12, year = 2016:2020,
                          county_code = c("53033", "53053"))
county_agg$county  <- ifelse(county_agg$county_code == "53033", "King", "Pierce")
county_agg$filings <- rpois(nrow(county_agg), 40)

peek(nt_profile_county_trend(
  label = "Figure 3 · Counties on the rise",
  caption_num = "Figure 3",
  caption_desc = "Monthly filings by county.",
  caption_src = "source · demo"), 200)
peek(nt_profile_county_trend_data(county_agg), 220)
```

## Figure 4 — the MapLibre county explorer

`nt_profile_map_explorer()` emits the full explorer section: the map node
with PMTiles + choropleth configuration in `data-*` attributes, the address
search, help/onboarding/loading affordances, the legend node, the time-window
slider stack, and the readout card (stat tiles, sparkline, ACS county
characteristics). Everything live about it — fetching the sidecars, painting
bins with ckmeans, the slider — is the page template's explorer JS.

```{r fig4}
peek(nt_profile_map_explorer(
  county_json = "data/o2b_county_monthly.json?v=1",
  acs_json    = "data/o2c_county_acs.json?v=1",
  pmtiles     = "assets/maps/o2_wa_counties.pmtiles?v=1",
  center = c(-120.74, 47.55), zoom = 6,
  bbox   = c(-124.85, 45.5, -116.92, 49.05)), 260)
```

Every piece of state-specific copy (search placeholder, disclosure text, the
readout's ACS note) is an argument with the Washington wording as default.

## Table 1 — the all-counties table

`nt_profile_county_table()` emits the data payload (a `window.*` global the
page JS renders rows from — and re-sums when the explorer's time window
changes) plus the search/sort shell and caption. The top county by rate gets
`hi: true` (the red highlight).

```{r table1}
co <- data.frame(county  = c("Clark", "Pierce", "King"),
                 renters = c(65952, 123930, 419249),
                 filings = c(2192, 3707, 8850))
co$rate <- round(1000 * co$filings / co$renters, 1)
peek(nt_profile_county_table(co[order(-co$rate), ],
  caption_desc = "Demo counties.", caption_src = "source · demo"), 300)
```

## Figure 5 — the "four P's" diagram

`nt_profile_fourps_svg()` builds the action section's policy-lever graphic:
the eviction-to-displacement pathway, four bands (Prevent / Protect /
Preserve / Produce) of tier-badged lever chips, the "strongest together"
connector, and the footer. All content — bands, levers, tiers, the paired
levers, footer lines — is an argument with the Washington graphic as the
default, so calling it bare reproduces the shipped figure:

```{r fig5}
peek(nt_profile_fourps_svg(), 260)
```

One sizing rule to respect: the in-SVG text uses
`calc(<px> / var(--sx, 8.2))`. The `8.2` fallback is the page CSS's 820px
min-width divided by the 100-unit viewBox — it keeps labels sane before the
page's scale-watcher runs. Don't reset it to 1.

## Emitting from Quarto

Each artifact is one chunk carrying the `results: asis` option:

```{r quarto-pattern, eval = FALSE}
# ```{r} chunk in the profile .qmd, with option `results: asis`
nt_raw_html(
  nt_profile_trend_svg(m$date, m$filings, baseline = pp_monthly,
                       band_start = band0, band_end = band1), "\n",
  nt_profile_monthly_js(m$filings, year_base = 2019, y_div = fig1_ymax)
)
```

`nt_raw_html()` matters: chunk output is re-read as markdown, and indented
HTML would otherwise be parsed into code blocks. The `{=html}` fence it
writes makes pandoc pass the fragment through byte-for-byte.
