---
title: "The three panes: affordability, availability, stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The three panes: affordability, availability, stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
# Chunks that touch the Census/HUD APIs or big local files are shown, not run.
# Chunks marked eval=TRUE run offline (bundled snapshot / hand-built frames).
```

## The simple version

The three-pane product answers one question three times, each time more honestly:

> **Where could a very-low-income family actually live?**

1. **Affordability** — *what would it cost them to live here?* Each tract gets a verdict at the
   rents **current tenants** pay: **affordable** (the typical unit takes ≤ 30% of their income),
   **roughly affordable** (30–50% — attainable only by stretching), or **not affordable** (more
   than half their income).
2. **Availability** — *could they actually get in?* The same verdict, re-priced at **today's
   asking rents** (what the door costs, not what sitting tenants pay), reported next to how many
   units actually **open up** in a year.
3. **Stability** — *could they stay?* Whether the neighborhood is one a low-income household can
   hold onto, from ERN's Housing Precarity Risk Model (displacement + eviction risk): **stable /
   elevated / precarious**.

Each pane is a three-color map — a **verdict, not a gradient** — because policy audiences read
maps by color, and a continuous scale that renders a place "mid" is read as "there isn't a
problem." The three maps multiply into one funnel sentence. King County, WA (VLI renters,
FY2024):

> Of every 100 very-low-income renter households, the standing stock could house **87** at
> ≤ 30% of income — but only **25** of those units open up in a year, **21** open *and* rent at
> ≤ 30% at today's asking prices, and weighting by neighborhood stability leaves about **17**
> openings a year in places a family can also hold onto.

That is the whole product. The rest of this vignette explains each pane in depth — what it
measures, what it deliberately assumes, and the questions readers actually ask.

```{r, eval = FALSE}
library(neighborhood)

# The full three-pane chain, one region:
x <- afford_index("53", "033", 2024, tenure = "rent", geometry = TRUE) |>  # the engine
  afford_verdict() |>                                    # pane 1: the verdict classes
  afford_entry(afford_zori()) |>                         # pane 2: re-priced at entry
  afford_stability("~/data/hprm/hprm_v5_full_2022.parquet")   # pane 3: can they stay?

afford_capacity(x)          # the funnel numbers (per 100 tier households)
afford_map(x)               # pane-1 map   (verdict at incumbent prices)
afford_map(x, verdict_col = "entry_verdict")   # pane-2 map (at asking prices)
afford_map(x, verdict_col = "stable_cat")      # pane-3 map (stability)
```

---

# Pane 1 — Affordability: what would it cost them to live here?

## What the map is saying

Pick the tier first. The pilot group is **VLI** — very low income, at or below 50% of the Area
Median Income, HUD's definition. The pane-1 map then says, tract by tract:

> **This is how much a VLI household will pay if they can move in there:** green — the typical
> unit takes no more than 30% of their income; orange — 30–50% (attainable only by stretching);
> red — more than half their income.

And for the red class, the locked descriptor:

> **Even at the rents current tenants pay, a VLI family would give up more than half their
> income in these neighborhoods — and entry prices run higher still.**

Three "what ifs" must travel with every figure, and `afford_caption()` prints them so they are
never retyped from memory:

```{r, eval = TRUE}
library(neighborhood)
afford_caption("VLI")   # the standard reading + what-ifs block, ready to place under a map
```

## How the verdict is computed

The engine (`afford_index()`, documented in `vignette("affordability-index")`) counts, per
tract, the rental units affordable to the tier at two burden lines:

- **`burden = 0.30`** — the HUD standard: gross rent (utilities included) at or below 30% of the
  tier's income ceiling;
- **`stretch = 0.50`** — HUD's *severe* burden threshold: reachable only by giving up 30–50% of
  income.

`afford_verdict()` then applies a **tract-majority rule**: a tract is *affordable* when at least
half its units clear the 30% line, *roughly affordable* when at least half clear the 50% line,
*not affordable* otherwise. At the default `share_cut = 0.5` each class is literally a
**median-unit sentence** — "the typical rental here would cost a VLI family more than half its
income" — which is why the classes survive a city-council meeting.

A hand-built example, small enough to read (this runs offline):

```{r, eval = TRUE}
toy <- data.frame(
  GEOID   = c("t-green", "t-orange", "t-red"),
  tenure  = "rent", ami_tier = "VLI",
  total   = c(100, 100, 100),
  accessible = c(62, 20, 4),            # units at <= 30% of the VLI ceiling
  supply           = c(0.62, 0.20, 0.04),
  accessible_stretch = c(90, 70, 30),   # units at <= 50% (the stretch line)
  supply_stretch     = c(0.90, 0.70, 0.30),
  reg_hh_tier = 1000, reg_hh_total = 4000,
  available_turnover = c(12, 5, 1), available_vacancy = c(3, 1, 0.2))

v <- afford_verdict(toy)
v[, c("GEOID", "supply", "supply_stretch", "verdict")]
```

The middle tract is the instructive one: only 20% of its units clear the 30% line, but 70%
clear the 50% line — a family *can* get in, by stretching. That "roughly affordable" middle is
usually the **largest class in every market we have run**, and it is the class a continuous
blue gradient hides.

## The income side, and where the dollars come from

The verdict hinges on the tier's income ceiling. Those come from `ami_cutoffs()` (five sources,
most-exact-first) — and the AMI level itself has a dedicated puller. HUD publishes income limits
for *income limit areas* — metro areas (MSAs or HUD's own HMFA subareas) and non-metropolitan
counties — so you query by county and the answer arrives from the county's HUD area. This runs
offline against the bundled FY2024 snapshot:

```{r, eval = TRUE}
hud_ami("53", "033")   # King County -> the Seattle-Bellevue HMFA's AMI, with provenance
```

Note the `NAME`: King County's number *is* the Seattle-Bellevue HUD Metro FMR Area's — there is
no more-local HUD AMI than the area. (The Bay Area splits into three HMFAs, so there is no
single "Bay Area AMI".) `ami_source = "acs"` swaps in the county's own median family income —
Census-only, reproducible, but ~13% low in high-cost coastal areas.

## Questions people ask

- **"Does green mean the neighborhood is cheap?"** No — it means *the typical unit is within
  30% of the VLI income ceiling*. In a high-AMI metro the ceiling is high, so green is relative
  to the region's own income structure, by design (that is what "affordable to VLI" means).
- **"But people already live in those cheap units."** Correct, and pane 1 does not pretend
  otherwise — it prices the *stock*, not the *vacancies*. Whether anything is open is pane 2;
  whether the cheap units are occupied by higher-income households (they often are) is the
  planned competition adjustment (NLIHC-style, HUD CHAS data), and until it lands every count is
  a stated **ceiling**.
- **"Whose income exactly?"** A household at the **top of the tier's range**
  (household-size-adjusted). Families deeper in the band give up a larger share for the same
  unit — so green is "manageable for the best-off VLI household."
- **"Why 30% and 50%?"** They are the federal standards — cost burden and severe cost burden —
  which makes the classes defensible in public: no invented thresholds.
- **"Why not a prettier continuous map?"** Because "0.42 supply share" invites the reader to
  shrug. The three classes are the finding. (Continuous columns — `supply`, `ratio`, `rate` —
  are all still in the data for analysts.)

## The other pane-1 outputs

- `afford_bands()` — turns cumulative ceilings into non-overlapping slices (the 30–50% band).
- `afford_capacity()` — the units-and-people arithmetic (next section).
- `inflow_index()` — **experimental**: where low-income households appear to be *moving*, from
  HPRM's net-migration predictions. Its positive tail failed face validity (exclusionary suburbs
  read as destinations), so use it for the displacement (negative) read only; destinations are
  measured by affordability + stability, not by it.

---

# Pane 2 — Availability: could they actually get in?

## What the map is saying

Pane 1 prices tracts at what **sitting tenants** pay — including the long-tenured cheap leases a
mover will never be offered. Pane 2 re-prices every tract at **entry**:

> **This is how much a VLI household will pay to move into this neighborhood today** — the same
> three classes, priced at today's asking rents instead of what sitting tenants pay.

And its counterpart to the locked descriptor:

> **The door costs more than the rent roll: where the affordability map is green and this one is
> not, the neighborhood is affordable only to those already in it.**

```{r, eval = TRUE}
afford_caption("VLI", pane = "availability")
```

## How the entry re-pricing works

Three steps, three functions:

```{r, eval = FALSE}
zori <- afford_zori()                    # 1. Zillow's ZIP-level asking-rent index (cached)
w    <- nt_zcta_weights("53", "033")     # 2. block-exact ZIP -> tract housing-unit weights
x    <- afford_entry(x, zori, weights = w)   # 3. premium -> entry verdict (weights optional;
                                             #    afford_entry builds them if omitted)
```

1. **`afford_zori()`** downloads (once, then caches) Zillow's Observed Rent Index — the smoothed
   typical **asking rent** of listed units, monthly, ~8,400 ZIPs.
2. **`nt_zcta_weights()`** crosswalks ZIPs to tracts *exactly*: 2020 census blocks nest inside
   both 2020 tracts and 2020 ZCTAs, so each tract's housing-unit split across ZIPs is counted,
   not areally approximated.
3. **`afford_entry()`** computes each ZIP's **entry premium** — asking rent ÷ the same ZIP's
   standing rent — brings it to tracts with the housing-unit weights, **shifts the tract's own
   rent distribution up by it**, and re-applies the same 30/50 verdict grammar
   (`entry_verdict`). Tracts with thin listing coverage fall back to the study-area median
   premium and are flagged (`premium_fallback`).

The ratio-shift matters: Zillow's index is the *typical listing* (mid-market), so using its
level as "the tract's rent" would mis-state the cheap segment. Shifting the tract's own
distribution keeps cheap tracts cheap — just re-priced to the door.

## What the wedge looks like

Median entry premiums and the verdict shift, VLI renters, computed 2026-07-04 (regenerate via
`dev/report_style_maps.Rmd` in the source repo):

| Market | median premium | verdict a / r / n, standing → entry |
|---|---|---|
| Bay Area (9-county) | **1.24×** | 380 / 1026 / 351 → **95 / 953 / 709** |
| Twin Cities (7-county) | 1.14× | 433 / 325 / 16 → 267 / 453 / 54 |
| King County | 1.07× | 139 / 315 / 40 → 87 / 339 / 68 |
| Atlanta (4-county) | 1.04× | 116 / 627 / 182 → 69 / 636 / 220 |

The premium is metro-specific and it is a *diagnosis*: in the Bay, asking rents run a quarter
above standing rents (tenure discounts, rent stabilization), so the green class collapses by
three-quarters at the door — the Bay is affordable mostly to those already housed in it. In
Atlanta asking ≈ standing (fast-turnover market); its constraint is the scarce standing base,
not the door. Same red maps, different diseases, different policy prescriptions.

## Questions people ask

- **"Is this Zillow's rent for my tract?"** No — it is your tract's *own* rent distribution,
  shifted by the ZIP-level asking-vs-standing ratio. Zillow enters only as the shift factor.
- **"Does it include whether a unit is open?"** Yes — availability is *both* the door price and
  the door being open. The turnover/vacancy openings from the engine ride along
  (`available_entry_turnover` = units that open per year **and** are entry-affordable).
- **"What is still missing?"** Competition (a cheap unit occupied by a higher-income household
  still counts — the CHAS adjustment is the next layer) and screening (records, credit,
  vouchers). Both stated on the caption; the counts are ceilings.
- **"Why do standing and entry disagree most in gentrifying places?"** Because that is where the
  gap between what incumbents pay and what the market now asks is widest — the premium *is* a
  gentrification-pressure readout.

---

# Pane 3 — Stability: could they stay?

## What the map is saying

A place you can afford and enter but cannot keep is not a destination. Pane 3 asks whether the
neighborhood is one a low-income household can *hold onto*, using ERN's Housing Precarity Risk
Model:

> **This is whether a VLI household that gets in can stay:** green — stable (no measurable
> displacement and below-average eviction risk, on HPRM's published anchors); orange — elevated
> risk; red — precarious (High/Extreme displacement or eviction pressure).

```{r, eval = TRUE}
afford_caption("VLI", pane = "stability")
```

## Why stability — and not an "opportunity" score

The expected third pane is an opportunity index — Chetty's Opportunity Atlas, the Child
Opportunity Index, the opportunity maps California uses to steer affordable-housing siting.
Pane 3 measures **housing stability instead**, and the choice is a position, not an omission.
Four reasons, each of which shapes how the pane should be read.

**1. Opportunity scores aggregate individual outcomes; the question here is about structural
constraints.** The Opportunity Atlas is the average adult outcome of children raised in a tract
decades earlier, published at tract scale. Its authors are careful about what that can support:
conventional neighborhood characteristics explain only about half of the tract-to-tract
variation in those outcomes, and their own experimental and quasi-experimental analysis
attributes about 60% of the variation to causal neighborhood effects — the remainder is
families sorting themselves (Chetty et al. 2018, rev. 2025). Meanwhile the forces that decide
*who lives where* — widening income inequality, stagnant wages, restructured metropolitan labor
markets — are systemic, not tract amenities (Chapple 2017). Systems create inequality, not
individuals; so the thing this product measures is the **structural constraint on a household's
choice set** — what it can afford, enter, and keep — not an aggregate of individual outcomes
read as a property of land.

**2. "High-opportunity" encodes a contested definition of good.** The tracts those maps score
highest are disproportionately suburban, single-family, and white — a one-way definition of
where low-income families of color should go (Goetz 2018). When Reid (2019) asked the people
the California maps are deployed for — residents of LIHTC housing — the opportunity domains
tracked residents' own assessments of their neighborhoods weakly or backwards and did not
predict their employment outcomes; residents located their barriers in labor- and
housing-market structures, and what they valued was affordability, predictability, and the
ability to stay. Affordable, stable housing is itself the platform for economic mobility — that
finding is the empirical warrant for this pane.

**3. Relocation-to-opportunity is not the lever it promises to be.** The Moving to Opportunity
experiment moved individual families to lower-poverty tracts; the sorted metropolitan structure
they moved within stayed put, and movers largely re-encountered it (Sampson 2008). Routing
households toward designated "good" tracts also carries the mirror-image risk — recreating
segregation under a nicer name. That is why the product's headline is the **funnel per 100
households** (the size of the constrained choice set), never a prescribed destination.

**4. Measurement discipline.** Displacement research is proxy-dependent: tract-composition
("population") measures, individual-mobility measures, and motivational measures rank the same
places differently, and studies using compositional change repeatedly find displacement is *not*
more likely in gentrifying tracts — a proxy artifact, not a reassurance (Carlson 2020). HPRM's
displacement and eviction signals are estimated from **household-level panel data and court
records** — the individual approach in Carlson's typology — rather than inferred from tract
composition change. The residual honesty: the published output is still a tract rate, so pane 3
is a **place-level risk descriptor, never an individual prediction** (the first question in the
Q&A below).

What survives of "opportunity" in this product: nothing is hidden, one signal is defensible.
The share of resident adults who are employed — the one opportunity measure that tracks
residents' own outcomes in Reid's data — can ride along as a clearly labeled side variable, and
an Atlas or COI layer can be added as a **secondary, contested lens** where a client requires
it, with its vintage and sorting caveats stated. But the spine is stability: *a place you can
afford and enter but cannot keep is not a destination.* The full design argument, with the
citation trail, is in `dev/good-place-design.md` (this repo) and the ERN manuscript *A Good
Place* (`hprm` repo).

## How stability is computed

`afford_stability()` joins two HPRM fields onto the result and maps each to a 0–1 score using
**HPRM's own published risk anchors** — no new thresholds are invented:

- **Displacement** (`dis_value`, the EDR net-migration composite): `+50 → 1.0` (no measurable
  outflow), `−300` (Extreme) `→ 0`.
- **Eviction** (`ev_value`, the EER filing-rate ratio): `0.8 → 1.0` (below-average), `2.0`
  (Extreme) `→ 0`.

`stability` is their weighted blend (`stability_source = "both"`, `"edr"`, or `"eer"`), and
`stable_cat` cuts it at 0.8 / 0.5 into **stable / elevated / precarious**. The continuous
headline is `stable_dest = available_turnover × stability` — openings, discounted by how
holdable the neighborhood is. Offline demonstration on the published anchors:

```{r, eval = TRUE}
hp <- data.frame(GEOID = c("t-green", "t-orange", "t-red"),
                 dis_value = c(50, -125, -300),   # +50 -> 1, -125 -> 0.5, -300 -> 0
                 ev_value  = c(0.8, 1.4, 2.0))    # 0.8 -> 1, 1.4 -> 0.5, 2.0 -> 0
s <- afford_stability(toy, hp)
s[, c("GEOID", "stability", "stable_cat", "available_turnover", "stable_dest")]
```

## Questions people ask

- **"Why isn't this pane an opportunity index?"** By design — see *Why stability — and not an
  "opportunity" score* above. Short version: opportunity scores aggregate individual outcomes
  onto tracts and encode a contested definition of "good"; the destination question needs the
  structural constraints on where a household can go and stay.
- **"Is this a prediction about my lease?"** No. HPRM is built from individual-level records but
  published as **tract rates** — it describes the neighborhood's pressure, not any one
  household's outcome.
- **"Why does a cheap tract show red?"** That is the finding, not a bug: affordability and
  instability travel together. Across every market we have measured, the correlation between a
  tract's affordability and its stability is about **−0.5** — the places a VLI family can afford
  are systematically the places under the most displacement pressure. That replicated from the
  published SLC/San Diego reports through the corrected engine.
- **"California looks different — why?"** The eviction model (EER) was trained without
  California, so `ev_value` there is out-of-sample (flagged per tract in `eer_out_of_sample`);
  the displacement side is the robust read. Bay Area maps lean on it.
- **"Vintage?"** HPRM v5 is fitted on 2022 data; the price panes use 2020–24 ACS and current
  asking rents — a stated two-year gap.

---

# The funnel: putting the three panes into one table

`afford_capacity()` turns the columns into units-and-people arithmetic — one row per
tenure × tier, everything per 100 tier households:

```{r, eval = TRUE}
cp <- afford_capacity(s)   # the toy frame, now carrying stability
round(as.data.frame(cp[, c("tier_households", "per100_affordable",
                           "per100_open_year", "per100_stable_dest", "shortfall")]), 1)
```

On real regions (VLI renters, per 100 households; computed 2026-07-04):

| Market | affordable (standing) | …open per year | …open **and** entry-affordable | …**and** stability-weighted |
|---|---|---|---|---|
| King County | 86.8 | 24.8 | 21.0 | **17.4** |
| Bay Area | 78.6 | 14.8 | 8.3 | **6.5** |
| Twin Cities | 103.9 | 26.7 | 21.5 | **14.7** |
| Atlanta | 51.3 | 11.2 | 10.2 | **6.1** |

Read a row left to right and you have the report paragraph: what the stock could hold, what
opens, what opens at a price a mover can pay, and what of *that* sits in neighborhoods they can
keep. Each market loses most of its funnel at a different stage — the Bay at the door, Atlanta
at the base, the Twin Cities at stability — which is exactly the kind of distinction a single
"affordability score" cannot make.

# The maps, together

`afford_map()` renders any pane (`verdict_col = "verdict"` / `"entry_verdict"` /
`"stable_cat"`), with the plain-language legend, an optional HPRM displacement hatch, and a
basemap decluttered of park/landcover tints. `nt_map_sync()` locks two or more maps' views
together, so a reader can pan to one neighborhood and read all three verdicts at once:

```{r, eval = FALSE}
nt_map_sync(
  afford_map(x, "VLI", displacement = hprm),
  afford_map(x, "VLI", verdict_col = "entry_verdict", displacement = hprm),
  afford_map(x, "VLI", verdict_col = "stable_cat"),
  titles = c("1 · affordability", "2 · availability at entry", "3 · stability"))
```

# What the product deliberately does not claim

- **Competition is not yet netted out** — an affordable unit occupied by a higher-income
  household still counts. Nationally, that adjustment roughly halves the affordable stock for
  the lowest incomes (NLIHC's *Gap*: 7.2M ELI-affordable units, 3.8M affordable *and*
  available). The HUD CHAS tract data is the planned source. Until then: ceilings, and the
  captions say so.
- **Screening is outside the measure** — records, credit, vouchers, discrimination ("the scarlet
  E"). It is the theory's third force and a research program of its own (observed-move
  calibration against the Data Axle panel).
- **The verdict is tier-relative** — "affordable to VLI" is a statement about the region's own
  income structure, not about absolute cheapness. And for broad bands (0–80% AMI) a burden
  verdict saturates by construction; use the supply-vs-need `ratio`/`rate` classes there and
  reserve the verdict for the tiers decisions target (VLI/ELI).

## See also

- `vignette("affordability-index")` — the engine underneath: `afford_index()`, `ami_cutoffs()`,
  every argument, the v1→v2 crosswalk.
- `vignette("mapping-with-maplibre")` — the mapping toolkit (`nt_map()`, synced maps,
  decluttering).
- In the source repo, `dev/affordability-deep-dive.md` (the v1 autopsy and replication),
  `dev/availability-design.md` (the entry-price layer design), `dev/good-place-design.md` (the
  full stability-vs-opportunity design argument), and `dev/report_style_maps.Rmd` (the
  four-market evaluation document that produced the numbers above).

## References

- Carlson, H. Jacob. 2020. "Measuring Displacement: Assessing Proxies for Involuntary
  Residential Mobility." *City & Community* 19(3): 573–592.
  [doi:10.1111/cico.12482](https://doi.org/10.1111/cico.12482)
- Chapple, Karen. 2017. "Income Inequality and Urban Displacement: The New Gentrification."
  *New Labor Forum* 26(1): 84–93.
  [doi:10.1177/1095796016682018](https://doi.org/10.1177/1095796016682018)
- Chetty, Raj, John N. Friedman, Nathaniel Hendren, Maggie R. Jones, and Sonya R. Porter. 2018
  (revised 2025). "The Opportunity Atlas: Mapping the Childhood Roots of Social Mobility."
  [NBER Working Paper 25147](https://www.nber.org/papers/w25147).
- Goetz, Edward G. 2018. *The One-Way Street of Integration: Fair Housing and the Pursuit of
  Racial Justice in American Cities.* Ithaca, NY: Cornell University Press.
- Reid, Carolina K. 2019. "Rethinking 'Opportunity' in the Siting of Affordable Housing in
  California: Resident Perspectives on the Low-Income Housing Tax Credit." *Housing Policy
  Debate* 29(4): 645–669.
  [doi:10.1080/10511482.2019.1582549](https://doi.org/10.1080/10511482.2019.1582549)
- Sampson, Robert J. 2008. "Moving to Inequality: Neighborhood Effects and Experiments Meet
  Social Structure." *American Journal of Sociology* 114(1): 189–231.
  [doi:10.1086/589843](https://doi.org/10.1086/589843)
